### 2021-02-02 r script/RegressionWorkup.Rmd
Initial commit

---
title: Preliminary analysis of fire behavior data
date: '`r Sys.Date()`'
author: DAM
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
---

```{r setup, echo=FALSE, warning=FALSE, message = FALSE, results='hide'}
knitr::opts_chunk$set(message = FALSE, warning=FALSE, 
               echo=FALSE, eval=TRUE, fig.path = "./") 
fp = "C:/Users/Devan.McGranahan/GoogleDrive/FireScienceDIY/SpatialFireBehavior"
load(paste0(fp, "/data/DataCombined.Rdata"))
load(paste0(fp, '/r objects/response_CIs.Rdata'))
```

```{r}
# Packages
  pacman::p_load(tidyverse, mice, broom.mixed)
```

# Data

Everything is available in the [SpatialFireBehavior folder](https://github.com/devanmcg/FireScienceDIY/tree/master/SpatialFireBehavior) in the [FireScienceDIY GitHub project](https://github.com/devanmcg/FireScienceDIY).

## Preparations

The raw weather, fuels, and fire behavior data used here are available in an [Excel spreadsheet](https://github.com/devanmcg/FireScienceDIY/blob/master/SpatialFireBehavior/data/AllLocations.xlsx).  
These data were [loaded with this script](https://raw.githubusercontent.com/devanmcg/FireScienceDIY/master/SpatialFireBehavior/r%20script/LoadingData.R) and combined with [this script](https://raw.githubusercontent.com/devanmcg/FireScienceDIY/master/SpatialFireBehavior/r%20script/ConnectingData.R) into [$\texttt{DataCombined.R}$](https://github.com/devanmcg/FireScienceDIY/blob/master/SpatialFireBehavior/data/DataCombined.Rdata).

Data are limited to $\textsf{Rate of spread < 40 m/s}$ and $\textsf{Maximum temperature > 40}^\circ \textsf{C}$.

```{r fig.width = 10, fig.height = 8, fig.cap="A scatterplot matrix of all raw data used in this analysis."}
# Filter data
  AnalysisData <- 
    DataCombined %>%
      filter( ros <= 40, 
              tempC >= 40) %>%
        mutate(FuelMoisture = ifelse(FuelMoisture >= 0, 
                                      FuelMoisture, NA), 
               FuelMoisture = FuelMoisture * 100) %>%
  select(-lys)
# Scatterplot matrix  
GGally::ggpairs(AnalysisData, 
                columns = 8:18, 
                upper = 'blank')
```

## Distributions

```{r fig.width = 8, fig.height =8, fig.cap = "Gamma distributions are fit to all response variables."}
# Distributions of response variables
  ros_dist <-
  AnalysisData %>% 
    ggplot(aes(x=ros)) + theme_bw(14) + 
    geom_histogram(aes(y=..density..),      
                   binwidth=0.5,
                   colour="black", 
                   fill="lightgreen") +
    geom_density(alpha=0.2, 
                 fill="#FF6666") 
  tempC_dist <- 
  AnalysisData %>% 
    ggplot(aes(x=tempC)) + theme_bw(14) + 
    geom_histogram(aes(y=..density..),      
                   binwidth=10,
                   colour="black", 
                   fill="lightgreen") +
    geom_density(alpha=0.2, 
                 fill="#FF6666") 
  soilC_dist <- 
  AnalysisData %>% 
  ggplot(aes(x=SoilMaxC)) + theme_bw(14) + 
  geom_histogram(aes(y=..density..),      
                 binwidth=10,
                 colour="black", 
                 fill="lightgreen") +
  geom_density(alpha=0.2, 
               fill="#FF6666") 
  gridExtra::grid.arrange(tempC_dist, soilC_dist, 
                          ros_dist, ncol=2)
```

## Correlations of interest

```{r fig.width = 5, fig.height =4,fig.cap = "Poor linear relationship between rate of spread and maximum temperature."}
# ROS fit against maximum temperature
  AnalysisData %>% 
    ggplot(aes(x=tempC, y = ros, color=location)) + theme_bw(14) +
      geom_point(alpha=0.5)
```

```{r fig.width = 5, fig.height =4, fig.cap="Unexpected and unexplained relationship between soil moisture and fuel moisture. Dropping soil moisture from predictor variables."}
# Fuel moisture fit against soil moisture
  ggplot(AnalysisData, aes(x=SoilMoisture, y=FuelMoisture, color=location)) + 
    geom_smooth(method = 'lm', se=F) + 
    geom_point(alpha=0.5) +
    coord_cartesian(ylim = c(0,200))
```

# Regression analysis

## Imputation of missing values

There are several missing values in the data:

```{r }
# MICE operations 
#
# Pattern of missing data
AnalysisData %>%
  select(-location, -block, -pasture,-patch,
         -year, -plot, -array, -dbC, -dpF, -SoilMoisture) %>%
    md.pattern(plot = F) %>%
  pander::pander(caption="Table of missing values by variable.") 
```

Therefore I used the multiple imputation method in the **mice** package to fill in missing values. 
The method creates 100 datasets, each with different but reasonable values for the missing data based on patterns in existing data, and performs the regression analysis on each. 
These results are pooled into a single set of results.

## Model specification

Mixed-effect regression models fit with `lme4::glmer` and constructed as follows:

$$y \approx RH + WS + FM + FL + (1|location/block/year/plot),~ 
\text{family=Gamma(link = `log')}$$

where

$RH$ = Relative humidity,

$WS$ = Average wind speed,

$FM$ = Fuel moisture (from clipped samples),

$FL$ = Fuel load (as estimated by LAI)

Weather variables $RH$ and $WS$ were taken from NDAWN stations for the hour in which the fire behavior observation occurred.

All variables were scaled to a common range prior to analysis, and units for each predictor variable are not reported.

```{r eval=FALSE}
# Calculate imputed datasets
  imp <- AnalysisData %>% 
              mutate_at(vars(plot, array), as.character) %>%
              mutate_at(vars(tempC:ros), ~as.numeric(scale(., center=F))) %>%
              select(-SoilMoisture, -dpF) %>%
               mice(m=100, seed = 23109, print=F)
# Rate of spread
  # Fit model 
    ros_all <- with(imp, 
                  suppressMessages(
                    lme4::glmer(ros ~ RH + mean_mph + 
                               FuelMoisture + LAI + 
                            (1|location/block/year/plot), 
                       family=Gamma(link = "log"), 
                       control=lme4::glmerControl(optimizer="bobyqa", 
                                        optCtrl=list(maxfun=100000)) )) )
    # Pool results
    ros_terms <-   
      summary(pool(ros_all)) %>%
        mutate_at(vars(estimate:p.value), ~round(., 2)) %>%
        select(term, estimate) %>%
      bind_cols(confint.mipo(pool(ros_all)) %>%
                  as_tibble() %>%
                  round(2) ) %>%
        arrange(desc(abs(estimate)))
# Maximum temperature  
   # Fit model 
    temp_all <- with(imp, 
                  suppressMessages(
                    lme4::glmer(tempC ~ RH + mean_mph + 
                             FuelMoisture + LAI + 
                             (1|location/block/year/plot), 
                        family=Gamma(link = "log"), 
                        control=lme4::glmerControl(optimizer="bobyqa",
                                      optCtrl=list(maxfun=100000)) )) )

  # Pool results
    temp_terms <-   
     summary(pool(temp_all)) %>%
     mutate_at(vars(estimate:p.value), ~round(., 2)) %>%
     select(term, estimate) %>%
     bind_cols(confint.mipo(pool(temp_all)) %>%
                 as_tibble() %>%
                 round(2) ) %>%
     arrange(desc(abs(estimate)))

#
# Maximum soil surface temperature  
#
  # Fit model 
   soil_all <- with(imp, suppressMessages(
     lme4::glmer(SoilMaxC ~ RH + mean_mph +  
                   FuelMoisture + LAI + 
                   (1|location/block/year/plot), 
                 family=Gamma(link = "log"), 
                 control=lme4::glmerControl(optimizer="bobyqa", 
                                  optCtrl=list(maxfun=100000)) )) )
  # Pool results
   soil_terms <-   
     summary(pool(soil_all)) %>%
     mutate_at(vars(estimate:p.value), ~round(., 2)) %>%
     select(term, estimate) %>%
     bind_cols(confint.mipo(pool(soil_all)) %>%
                 as_tibble() %>%
                 round(2) ) %>%
     arrange(desc(abs(estimate)))

# Combine regression results
  response_CIs <-   
          bind_rows(
            mutate(soil_terms, response = 'Temp. at surface'), 
            mutate(temp_terms, response = 'Temp. above surface'),
            mutate(ros_terms, response = "Rate of spread") 
            )
```

# Results

Fuel moisture reduced maximum temperature ($t_{67} = -2.74, P = 0.01$) whereas weather variables ($\textsf{Wind speed}$, $t_{68} = 2.69, P = 0.01$; $\textsf{RH}$, $t_{70} = -2.07, P = 0.04$ ) influenced rate of spread. 
These data might indicate that fuel variables are more important to first-order effects on basal plant tissue and soil heating.

Interestingly, there was a trend ($t_{66} = -1.94, P = 0.06$) towards higher temperature at the soil surface as RH declined.
This likely indicates higher intensity fires that burn down through litter. 
The negative trend between wind speed and soil surface temperatures was not statistically significant ($t_{61} = -1.62, P = 0.11$).

Further work should include quantifying total heat exposure via infrared sensors.

```{r fig.width=8, fig.height = 3.75}
# Plot confidence intervals
response_CIs %>%
  filter(term != '(Intercept)') %>%
  as_tibble() %>%
  mutate(term = recode(term, LAI = 'Fuel load', 
                             mean_mph = 'Wind speed', 
                             RH = 'Relative humidity', 
                             FuelMoisture = 'Fuel moisture'))  %>%
  ggplot(aes(x = reorder(term, abs(estimate), max))) + theme_bw(16) + 
    geom_hline(yintercept = 0, color="black", linetype = 2) +
    geom_errorbar(aes(ymin = `2.5 %`, ymax = `97.5 %`), 
                  size = 1.1, width = 0.2, 
                  color = "blue") +
    geom_point(aes(y = estimate), pch = 21, 
               stroke = 1.5, size = 4,
               color = "blue", fill="lightblue") +
    labs(x = '', 
         y = 'Regression coefficient with 95% CI') + 
    coord_flip(ylim=c(-1.9,1.9)) + 
    facet_wrap(~ response) + 
    theme(axis.text.y = element_text(color="black", size = 14), 
          strip.text = element_text(size = 14), 
          panel.grid.major.y = element_blank())
```

# Script

```{r script, ref.label=knitr::all_labels(!label %in% c('setup') ), echo=TRUE,eval=FALSE}
```

### 2021-02-02 mapping/script/HREC_fire_towers.R

@@ -1,0 +1,41 @@
+
+ 
+# Importing and converting data
+hrec.FT.shp <- rgdal::readOGR(dsn="C:/Users/devan.mcgranahan/GoogleDrive/QGIS/HREC/sampling/FireBehavior",
+                     layer="FireTowerPlacements")
+ 
+ hrec.FT.shp@data$id <- rownames(hrec.FT.shp@data)
+    hrec.FT.df <- broom::tidy(hrec.FT.shp, region="id")
+    hrec.FT.df <- plyr::rename(hrec.FT.df, c("coords.x1"="lon", 
+                                             "coords.x2"="lat")) 
+    hrec.FT.df$PastureRep <- tolower(hrec.FT.df$PastureRep)
+    hrec.FT.df$Livestock <- tolower(hrec.FT.df$Livestock)
+    # save(hrec.FT.df, file="./r objects/hrec.FT.df.Rdata")
+    
+# Plot tower locations for datasheet
+ pacman::p_load(ggplot2)
+ source("C:/Users/devan.mcgranahan/GoogleDrive/code snippets/ggplot themes/theme_map2.R")
+ load("C:/Users/devan.mcgranahan/GoogleDrive/QGIS/HREC/r objects/hrec.patches.df.Rdata")
+ load("C:/Users/devan.mcgranahan/GoogleDrive/Research/Projects/SpatialFireBehavior/r objects/hrec.FT.df.Rdata")
+ 
+ hett.FT.gg <- ggplot()  + theme_map2() +coord_quickmap() +
+    geom_polygon(data=hrec.patches.df, aes(x=long, y=lat, 
+                                         group=group, fill=status2017), 
+               color="black", size=0.25, 
+               show.legend=FALSE) +
+   geom_point(data=hrec.FT.df, aes(x=lon, y=lat), 
+               show.legend=FALSE) + 
+  scale_fill_manual(values= c("lightgrey", "white"))+ 
+   facet_wrap(~PastureRep, 
+                     scales="free", ncol=1)
+   
+ hett.FT.gg +
+   geom_text(data=hrec.patches.df, aes(x=long, y=lat, label=Patch))
+  geom_point(data=res, aes(x=lon, y=lat))
+  
+  vegan::envfit(X=hrec.patches.df[13:14], P=hrec.patches.df$Patch)
+  
+  patch.cen <- vegan::envfit(c(hrec.patches.df$long, hrec.patches.df$lat) ~ Patch, hrec.patches.df)
+  
+  
+ 

### 2021-02-02 r script/ConnectingData.R

@@ -1,0 +1,45 @@
+pacman::p_load(tidyverse)
+
+load("./r objects/wx_d.Rdata")
+load("./r objects/ROS.Rdata")
+load("./r objects/fuels.Rdata")
+load("./r objects/temps.Rdata")
+load("./r objects/soil.Rdata")
+
+# Summarize to array scale 
+
+  fuels_sum <- 
+    fuels %>%
+      select(-date, -location, -block, -burned) %>%
+      pivot_longer(cols = (LAI:FuelMoisture), 
+                   names_to = "response", 
+                   values_to = "value") %>%
+      group_by(FireCode, plot, array, response) %>%
+        summarize(value = mean(value)) %>%
+      ungroup() %>%
+      pivot_wider(names_from = response, 
+                  values_from = value)
+
+# Summarize & bring in weather data by the hour 
+
+  DataCombined <- 
+    temps %>% 
+    group_by(FireCode,location, plot, array) %>%
+    summarize(timestamp = mean(timestamp), 
+              tempC = mean(tempC)) %>%
+    ungroup() %>%
+    mutate(DayHour = format(timestamp, "%Y-%d-%m %H")) %>% 
+    unite(BurnHourID, c(location, DayHour), sep='_') %>%
+    select(-timestamp) %>%
+    left_join(wx_d, 
+              by = 'BurnHourID')%>%
+    left_join(fuels_sum, 
+              by = c('FireCode', 'plot', 'array')) %>%
+    select(-BurnHourID) %>%
+    left_join(ROS, 
+              by = c('FireCode', 'plot', 'array')) %>%
+    left_join(soil, 
+              by = c('FireCode', 'plot', 'array')) %>%
+    separate(FireCode, c('location','block','pasture','patch','year')) 
+  
+  # save(DataCombined, file="./data/DataCombined.Rdata")

### 2021-02-02 r script/LoadingData.R

@@ -1,0 +1,102 @@
+pacman::p_load(tidyverse, readxl)
+
+wd_fp = "C:/Users/Devan.McGranahan/GoogleDrive/FireScienceDIY/SpatialFireBehavior/data/"
+
+xl_fp = paste0(wd_fp, "/AllLocations.xlsx")
+
+
+# Soil temp data 
+
+soil <-  
+  read_xlsx("./data/fromMZ/Research_Fire_Database.xlsx", "Soil_TC") %>%
+    select(-AirTemp, -RH, -WindSpd, -MaxWindSpd, -WindDir) %>%
+    mutate(date = as.Date(date),
+           L = str_remove(location, "REC"), 
+           B = str_sub(block, 1,3), 
+           Ps = str_replace(pasture, "[.]", ""), 
+           Ps = str_sub(Ps, 1,2), 
+           y = format(date, "%y")) %>%
+  unite("FireCode", c(L,B,Ps,patch,y), sep=".") %>%
+  select(FireCode, plot, array, SoilMaxC) %>%
+    filter(SoilMaxC != is.na(.))
+
+# save(soil, file="./r objects/soil.Rdata")
+
+# Thermocouple data 
+temps <- 
+  read_csv(paste0(wd_fp, "ThermocoupleData.csv")) %>%
+  filter(tempC != is.na(.)) %>%
+    mutate(time = str_remove(time, "[.]+[0-9]"))%>%
+    unite(timestamp, c(date, time), sep = " ") %>%
+  mutate(timestamp = as.POSIXct(timestamp, format = "%m-%d-%Y %H:%M:%S")) %>%
+    mutate(L = str_remove(location, "REC"), 
+           B = str_sub(block, 1,3), 
+           Ps = str_replace(pasture, "[.]", ""), 
+           Ps = str_sub(Ps, 1,2), 
+           y = format(timestamp, "%y"),
+           problematic = as.character(problematic)) %>%
+    unite("FireCode", c(L,B,Ps,patch,y), sep=".") %>%
+    filter(is.na(problematic) ) %>%
+    select(FireCode, timestamp, location, block, plot, array, 
+           TC, tempC) 
+
+# save(temps, file="./r objects/temps.Rdata")
+
+# Fuels data 
+fuels <- 
+  read_xlsx(xl_fp, "fuels") %>%
+    mutate(date = paste(str_sub(date, 1,4), 
+                        str_sub(date, 5,6),
+                        str_sub(date, 7,8),
[…]

### 2021-02-02 r script/RegressionAnalysis.R

@@ -1,0 +1,165 @@
+pacman::p_load(tidyverse)
+
+setwd("C:/Users/Devan.McGranahan/GoogleDrive/Research/Projects/SpatialFireBehavior")
+
+load("./data/DataCombined.Rdata")
+
+AnalysisData <- 
+  DataCombined %>%
+    filter( ros <= 40, 
+            tempC >= 40) %>%
+      mutate(FuelMoisture = ifelse(FuelMoisture >= 0, 
+                                    FuelMoisture, NA), 
+             FuelMoisture = FuelMoisture * 100)
+  
+GGally::ggpairs(AnalysisData, columns = 8:19)
+
+# Distribution
+
+AnalysisData %>% 
+  ggplot(aes(x=ros)) + theme_bw(14) + 
+  geom_histogram(aes(y=..density..),      
+                 binwidth=0.5,
+                 colour="black", 
+                 fill="lightgreen") +
+  geom_density(alpha=0.2, 
+               fill="#FF6666") 
+
+AnalysisData %>% 
+  ggplot(aes(x=tempC)) + theme_bw(14) + 
+  geom_histogram(aes(y=..density..),      
+                 binwidth=10,
+                 colour="black", 
+                 fill="lightgreen") +
+  geom_density(alpha=0.2, 
+               fill="#FF6666") 
+
+AnalysisData %>% 
+  ggplot(aes(x=SoilMaxC)) + theme_bw(14) + 
+  geom_histogram(aes(y=..density..),      
+                 binwidth=10,
+                 colour="black", 
+                 fill="lightgreen") +
+  geom_density(alpha=0.2, 
+               fill="#FF6666") 
+
+AnalysisData %>% 
+  ggplot(aes(x=tempC, y = ros, color=location)) + theme_bw(14) +
+    geom_point(alpha=0.5)
+
+  ggplot(AnalysisData, aes(x=SoilMoisture, y=FuelMoisture, color=location)) + 
+    geom_smooth(method = 'lm') + 
+    geom_point(alpha=0.5) +
+    coord_cartesian(ylim = c(0,200))
+  
+  AnalysisData %>% 
+    ggplot(aes(x=dbC, y = ros, color=location)) + theme_bw(14) +
+    geom_point(alpha=0.5)
+
+# MICE regression 
+
+pacman::p_load(mice, broom.mixed)
+
+  # View pattern of missing data
+    AnalysisData %>%
+        md.pattern() 
+
[…]

### 2021-02-02 r script/SpatializeImputedData.R

@@ -1,0 +1,61 @@
+pacman::p_load(tidyverse, sf)
+
+# 'Flatten" imputed mids object into single tibble
+  imp_d <- 
+    complete(imp, 'long') %>%
+      as_tibble() %>%
+            unite("TreeID", c(location, block, pasture,
+                              year, plot, array), sep = ".") %>%
+            select(-patch, -.id, -.imp) %>%
+            pivot_longer(names_to = "response", 
+                         values_to = "values", 
+                         -TreeID) %>%
+            group_by(TreeID, response) %>%
+              summarize(value = median(values)) %>%
+            ungroup() %>%
+            pivot_wider(names_from = response, 
+                         values_from = value) %>%
+            separate(TreeID, c("location", "block", "pasture",
+                               "year", "plot", "array"))
+
+# Assign triangle coordinates in sf object
+  imp_sf <-        
+    read_sf("mapping/DummyFireTowers/DummyTrianglePoints.shp") %>%
+      select(SubPlot, FireTower) %>%
+        unite("TreeID", c(SubPlot, FireTower), sep=".") %>%
+      full_join(unite(imp_d, "TreeID", c(plot, array), sep=".")) 
+  
+# Convert sf object to tibble with coordinates
+  imp_sf %>% as_tibble() %>%
+    st_write("./data/imp_coords.csv",  layer_options = "GEOMETRY=AS_XY") 
+  imp_coords <- read_csv("./data/imp_coords.csv")
+  
+  ggplot(imp_sf) +
+    geom_sf()
+  
+  mite.xy <- select(imp_coords, X, Y) 
+  mite.dbmem1 <- adespatial::dbmem(mite.xy, thresh=1.012, 
+                                   MEM.autocor = "non-null",
+                                   silent = FALSE)
+  mite.dbmem1  
+# Create sp object 
+  CanTemp_sp <- 
+    imp_coords %>%
+    unite(FireID, c(location, block, pasture, year), sep=".") %>%
+          select(FireID, X, Y, tempC) 
+  
+  
+  CanTemp_sp1 <- filter(CanTemp_sp, FireID == "CG.BAR.NE.18")
+
+
+  sp::coordinates(CanTemp_sp1) = ~X+Y
+  airC_v <- gstat::variogram(tempC~1, CanTemp_sp)
+  
+  v = variogram(log(zinc)~x+y, meuse)
[…]

### 2021-02-02 r script/SummaryGraphing.R

@@ -1,0 +1,23 @@
+
+AnalysisData %>%
+  select(location, tempC, ros, SoilMaxC) %>%
+  pivot_longer(names_to = "response", 
+               values_to = "value", 
+               -location) %>%
+  mutate(location = recode(location, 
+                           CG = "Streeter", 
+                           H = "Hettinger"), 
+         response = recode(response, 
+                           tempC = "Max canopy temp (C)", 
+                           ros = "Rate of spread (m/s)", 
+                           SoilMaxC = "Max soil surface temp (C)")) %>%
+  ggplot() + theme_bw(16) +
+  geom_boxplot(aes(x=location, y = value, fill=location), 
+               size = 1)  + 
+  labs(y = '', 
+       x = 'Research station') + 
+  facet_wrap(~response, scales = "free_y") +
+  scale_fill_viridis_d(option='plasma', guide = F, alpha = 0.5) + 
+  theme(panel.grid.major.x = element_blank(), 
+        axis.text.x = element_text(color='black'), 
+        strip.background = element_rect(fill = 'lightblue')) 

### 2021-02-03 r script/LoadingCompiledData.R
update data and analysis

@@ -1,0 +1,76 @@
+pacman::p_load(tidyverse)
+
+wd_fp = "C:/Users/devan/GitHubProjects/SpatialFireBehavior/" # home PC
+
+# Pull data from .csv created from MZ's compiled Excel file (tab = All)
+
+  AllData <-  
+    read_csv(paste0(wd_fp, "data/fromMZ/CompiledData.csv")) %>%
+    filter(location != "OAK") %>%
+      mutate(date = as.Date(date, format = "%m/%d/%Y"),
+               L = str_remove(location, "REC"), 
+               B = str_sub(block, 1,3), 
+               Ps = str_replace(pasture, "[.]", ""), 
+               Ps = str_sub(Ps, 1,2), 
+               patch = str_replace(patch, "[.]", ""),
+               y = format(date, "%y")) %>%
+        unite("FireCode", c(L,B,Ps,patch,y), sep=".") %>%
+    mutate(time = str_remove(MaxTempTime, "[.]+[0-9]"))%>%
+    unite(timestamp, c(date, time), sep = " ") %>%
+    mutate(timestamp = as.POSIXct(timestamp, format = "%Y-%m-%d %H:%M:%S")) %>%
+        select(FireCode, timestamp, plot, array, TC, MaxC, 
+               AirTemp, RH, MaxWindSpeed, 
+               LAI, FMC, KgHa) 
+
+# Isolate soil surface temperature (TC 4)
+  
+  SoilTemp <-
+    filter(AllData, TC == 4) %>%
+      select(FireCode, plot, array, MaxC) %>%
+        rename(SoilC = MaxC)
+  
+# Summarize array-level data 
+  DataMeans <- 
+    AllData %>%
+      filter(TC %in% c('1', '2', '3')) %>% 
+      select(-timestamp) %>%
+      pivot_longer(cols = c(MaxC:KgHa), 
+                   names_to = "var",
+                   values_to = "value") %>%
+      group_by(FireCode, plot, array, var) %>%
+        summarize(Mean = mean(value) ) %>%
+      ungroup() %>%
+      pivot_wider(names_from = var, 
+                  values_from = Mean) 
+    
+  
+# Calculate ROS 
+
+  D = 1   # Distance between thermocouples (m)
+  
+  ROS <- 
+    AllData %>%
+      filter(TC %in% c('1', '2', '3')) %>%
+      mutate(timestamp = format(timestamp, "%H:%M:%S"), 
[…]

### 2021-02-03 r script/RegressionAnalysis2.R

@@ -1,0 +1,163 @@
+pacman::p_load(tidyverse)
+
+source("./r script/LoadingCompiledData.R")
+
+AnalysisData <- 
+  DataCombined %>%
+    filter( ros <= 40, 
+            tempC >= 40) %>%
+      mutate(FuelMoisture = ifelse(FuelMoisture >= 0, 
+                                    FuelMoisture, NA), 
+             FuelMoisture = FuelMoisture * 100)
+  
+GGally::ggpairs(AnalysisData, columns = 8:19)
+
+# Distribution
+
+AnalysisData %>% 
+  ggplot(aes(x=ros)) + theme_bw(14) + 
+  geom_histogram(aes(y=..density..),      
+                 binwidth=0.5,
+                 colour="black", 
+                 fill="lightgreen") +
+  geom_density(alpha=0.2, 
+               fill="#FF6666") 
+
+AnalysisData %>% 
+  ggplot(aes(x=tempC)) + theme_bw(14) + 
+  geom_histogram(aes(y=..density..),      
+                 binwidth=10,
+                 colour="black", 
+                 fill="lightgreen") +
+  geom_density(alpha=0.2, 
+               fill="#FF6666") 
+
+AnalysisData %>% 
+  ggplot(aes(x=SoilMaxC)) + theme_bw(14) + 
+  geom_histogram(aes(y=..density..),      
+                 binwidth=10,
+                 colour="black", 
+                 fill="lightgreen") +
+  geom_density(alpha=0.2, 
+               fill="#FF6666") 
+
+AnalysisData %>% 
+  ggplot(aes(x=tempC, y = ros, color=location)) + theme_bw(14) +
+    geom_point(alpha=0.5)
+
+  ggplot(AnalysisData, aes(x=SoilMoisture, y=FuelMoisture, color=location)) + 
+    geom_smooth(method = 'lm') + 
+    geom_point(alpha=0.5) +
+    coord_cartesian(ylim = c(0,200))
+  
+  AnalysisData %>% 
+    ggplot(aes(x=dbC, y = ros, color=location)) + theme_bw(14) +
+    geom_point(alpha=0.5)
+
+# MICE regression 
+
+pacman::p_load(mice, broom.mixed)
+
+  # View pattern of missing data
+    AnalysisData %>%
+        md.pattern() 
+
+  # Calculate imputed datasets
+    imp <- AnalysisData %>% 
+                mutate_at(vars(plot, array), as.character) %>%
[…]

### 2021-02-05 r script/LoadingCompiledData.R
update new MZ data

@@ -1,11 +1,13 @@
-pacman::p_load(tidyverse)
+pacman::p_load(tidyverse, lubridate)
 
-wd_fp = "C:/Users/devan/GitHubProjects/SpatialFireBehavior/" # home PC
+# wd_fp = "C:/Users/devan/GitHubProjects/SpatialFireBehavior/" # home PC
+
+wd_fp = getwd() # machine-independent?
 
 # Pull data from .csv created from MZ's compiled Excel file (tab = All)
 
   AllData <-  
-    read_csv(paste0(wd_fp, "data/fromMZ/CompiledData.csv")) %>%
+    read_csv(paste0(wd_fp, "/data/fromMZ/CompiledData2.csv")) %>%
     filter(location != "OAK") %>%
       mutate(date = as.Date(date, format = "%m/%d/%Y"),
                L = str_remove(location, "REC"), 
@@ -19,7 +21,7 @@ wd_fp = "C:/Users/devan/GitHubProjects/SpatialFireBehavior/" # home PC
     unite(timestamp, c(date, time), sep = " ") %>%
     mutate(timestamp = as.POSIXct(timestamp, format = "%Y-%m-%d %H:%M:%S")) %>%
         select(FireCode, timestamp, plot, array, TC, MaxC, 
-               AirTemp, RH, MaxWindSpeed, 
+               AirTemp, RH, dpC, MaxWindSpeed, 
                LAI, FMC, KgHa) 
 
 # Isolate soil surface temperature (TC 4)

### 2021-02-08 writeups/FuelsWeatherFirebehavior.Rmd
update analyses & clean house on old files

---
title: Fuels, fire weather, and Rx fire behavior in North Dakota
subtitle: 
date: '`r Sys.Date()`'
author: DAM
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
---

```{r setup, echo=FALSE, warning=FALSE, message = FALSE, results='hide'}
knitr::opts_chunk$set(message = FALSE, warning=FALSE, 
               echo=FALSE, eval=TRUE, fig.path='figures/', dev=c('png', 'pdf')) 
FilePath = "C:/Users/devan/GitHubProjects/SpatialFireBehavior"
load(paste0(FilePath, '/data/AnalysisData.Rdata'))
load(paste0(FilePath, '/r objects/response_CIs.Rdata'))
load(paste0(FilePath, '/r objects/imp_raw.Rdata'))
source(paste0(FilePath, '/r script/CustomThemes.R'))
source("https://raw.githubusercontent.com/devanmcg/IntroRangeR/master/11_IntroMultivariate/ordinationsGGplot.R")
```

```{r}
# Packages
  pacman::p_load(tidyverse, mice, lubridate, gridExtra, vegan)
```

# Methods

## Study locations

![Hettinger and Central Grasslands (Streeter) Research Extension Centers.](figures/RECmap.png)

## Analysis

### Data & Distributions

The raw weather, fuels, and fire behavior data used here are available in [CSV format](https://github.com/devanmcg/SpatialFireBehavior/raw/main/data/fromMZ/CompiledData2.csv).

```{r data_prep, eval=FALSE}
 AllData <-  
    read_csv(paste0(FilePath, "data/fromMZ/CompiledData2.csv")) %>%
    filter(location != "OAK") %>%
      mutate(date = as.Date(date, format = "%m/%d/%Y"),
               L = str_remove(location, "REC"), 
               B = str_sub(block, 1,3), 
               Ps = str_replace(pasture, "[.]", ""), 
               Ps = str_sub(Ps, 1,2), 
               patch = str_replace(patch, "[.]", ""),
               y = format(date, "%y")) %>%
        unite("FireCode", c(L,B,Ps,patch,y), sep=".") %>%
    mutate(time = str_remove(MaxTempTime, "[.]+[0-9]"))%>%
    unite(timestamp, c(date, time), sep = " ") %>%
    mutate(timestamp = as.POSIXct(timestamp, format = "%Y-%m-%d %H:%M:%S")) %>%
        select(FireCode, timestamp, plot, array, TC, MaxC, 
               AirTemp, RH, dpC, MaxWindSpeed, 
               LAI, FMC, KgHa)

# Isolate soil surface temperature (TC 4)
  SoilTemp <-
    filter(AllData, TC == 4) %>%
      select(FireCode, plot, array, MaxC) %>%
        rename(SoilC = MaxC)

# Summarize array-level data 
  DataMeans <- 
    AllData %>%
      filter(TC %in% c('1', '2', '3')) %>% 
      select(-timestamp) %>%
      pivot_longer(cols = c(MaxC:KgHa), 
                   names_to = "var",
                   values_to = "value") %>%
      group_by(FireCode, plot, array, var) %>%
        summarize(Mean = mean(value) ) %>%
      ungroup() %>%
      pivot_wider(names_from = var, 
                  values_from = Mean)

# Calculate Vapor Pressure Deficit
  DataMeans <- 
    DataMeans %>%
      mutate(e  = 6.11*(10^((7.5*dpC)/(237.3+dpC))), 
             es = 6.11*(10^((7.5*AirTemp)/(237.3+AirTemp))), 
             VPD = es - e) %>%
      select(-e, -es)

# Calculate ROS 
  D = 1   # Distance between thermocouples (m)
  ROS <- 
    AllData %>%
      filter(TC %in% c('1', '2', '3')) %>%
      mutate(timestamp = format(timestamp, "%H:%M:%S"), 
             ArrivalTime = seconds(hms(timestamp)) ) %>%
    select(FireCode, plot, array, ArrivalTime) %>%
    group_by(FireCode, plot, array) %>%
    arrange(ArrivalTime, .by_group = TRUE) %>% 
    mutate(position = order(order(ArrivalTime, decreasing=FALSE)), 
           position = recode(position, "1"="a", "2"="b", "3"="c"), 
           ArrivalTime = as.numeric(ArrivalTime) /60 ) %>%
    spread(position, ArrivalTime)  %>%
    ungroup %>% 
    mutate( theta_rad = atan((2*c - b - a) / (sqrt(3)*(b - a))), 
            ros = case_when(
              a == b ~ (sqrt(3) / 2) / (c - a) , 
              a != b ~  (D*cos(theta_rad) / (b - a) ) 
            )) %>%
    select(-a, -b, -c, -theta_rad)

# Combine, filter into final tibble for analysis 
  AnalysisData  <- 
    full_join(DataMeans, ROS) %>%
              left_join(SoilTemp) %>%
      filter( ros <= 40, 
              MaxC >= 40) %>%
      rename(FuelMoisture = FMC, 
             SoilMaxC = SoilC) %>%
        mutate(FuelMoisture = ifelse(FuelMoisture >= 0, 
                                      FuelMoisture, NA), 
               FuelMoisture = FuelMoisture * 100) %>%
      separate(FireCode, into = c("location", "block", "pasture", 
                           "patch", "year"), 
               remove = F)
```

```{r fig.width = 10, fig.height = 8, fig.cap="A scatterplot matrix of all raw data used in this analysis."}
# Scatterplot matrix  
  GGally::ggpairs(AnalysisData, 
                  columns = 9:18, 
                  upper = 'blank')
```

```{r dist_gg, fig.width = 8, fig.height =8, fig.cap = "Gamma distributions are fit to all response variables. Two outliers above 600 deg C were removed from analysis on soil maximum temperature."}
# Distributions of response variables
  ros_dist <-
  AnalysisData %>% 
    ggplot(aes(x=ros)) + theme_bw(14) + 
    geom_histogram(aes(y=..density..),      
                   binwidth=0.5,
                   colour="black", 
                   fill="lightgreen") +
    geom_density(alpha=0.2, 
                 fill="#FF6666") 
  tempC_dist <- 
  AnalysisData %>% 
    ggplot(aes(x=MaxC)) + theme_bw(14) + 
    geom_histogram(aes(y=..density..),      
                   binwidth=10,
                   colour="black", 
                   fill="lightgreen") +
    geom_density(alpha=0.2, 
                 fill="#FF6666") 
  soilC_dist <- 
  AnalysisData %>% 
  ggplot(aes(x=SoilMaxC)) + theme_bw(14) + 
  geom_histogram(aes(y=..density..),      
                 binwidth=10,
                 colour="black", 
                 fill="lightgreen") +
  geom_density(alpha=0.2, 
               fill="#FF6666") 
  gridExtra::grid.arrange(tempC_dist, soilC_dist, 
                          ros_dist, ncol=2)
```

### Regression analysis

For analysis, are filtered to $\textsf{Rate of spread < 40 m/s}$ and $\textsf{Maximum temperature > 40}^\circ \textsf{C}$.

**Imputation of missing values**

There are several missing values in the data:

```{r missing_table}
# MICE operations 
#
# Pattern of missing data
AnalysisData %>%
  select(-FireCode, -location, -block, -pasture,-patch,
         -year, -plot, -array) %>%
    mice::md.pattern(plot = F) %>%
  pander::pander(caption="Table of missing values by variable.") 
```

Therefore I used the multiple imputation method in the **mice** package to fill in missing values. 
The method creates 100 datasets, each with different but reasonable values for the missing data based on patterns in existing data, and performs the regression analysis on each. 
These results are pooled into a single set of results.

**Model specification**

Mixed-effect regression models fit with `glmmADMB::glmmadmb` and constructed as follows:

$$y \approx RH + WS + FM + FL + (1|location/year/plot),~ 
\text{family = gamma, link = log}$$

where

$RH$ = Relative humidity,

$WS$ = Average wind speed,

$FM$ = Fuel moisture (from clipped samples),

$FL$ = Fuel load (t/ha from clipped samples)

Weather variables $RH$ and $WS$ were taken from NDAWN stations for the hour in which the fire behavior observation occurred.

All variables were scaled to a common range prior to analysis, and units for each predictor variable are not reported.

Response variables included

* Maximum temperature (C) 15 cm above soil surface (mean of three thermocouples)
* Maximum temperature (C) at soil surface (single thermocouple)
* Rate of spread (m/s) through three points of a 1 m equilateral triangle

Thermocouple data recorded with the open-source [FeatherFlame](https://diyfirescience.info/portfolio/work2/) system.

### Multivariate analysis

Principal Components Analysis (PCA) on fire behavior responses fit with `vegan::rda`. 
Post-hoc group (location) and gradient (fire weather) analysis with `vegan::envfit`.

An additional composite variable, vapor pressure deficit (VPD), was calculated as

$\text{VPD} = e - e_s$

where

$e = 6.11 \cdot 10^{\frac{7.5 \cdot \text{Dewpoint}}{237.3 + \text{Dewpoint}}}$

and

$e_s = 6.11 \cdot 10^{\frac{7.5 \cdot \text{Air temperature}}{237.3 + \text{Air temperature}}}$

```{r ADMB_mods, eval=FALSE}
# Max soil surface temperature 
  # Fit model
    soil_ADMB <- glmmADMB::glmmadmb(SoilMaxC ~ RH + MaxWindSpeed +
                            FuelMoisture + tHa,
                          random = ~ 1|location/year/plot ,
                          data=filter(imp_d, SoilMaxC <= 3.0),
                          family="gamma",
                          link="log",  
                          admb.opts=glmmADMB::admbControl(shess = TRUE, 
                                                          noinit = TRUE, 
                                                          impSamp = 10) )
  # Retrieve regression results
    soil_terms <-   
      confint(soil_ADMB) %>%
      as.data.frame %>%
      rownames_to_column("term") %>%
      rename(lwr = `2.5 %`, upr = `97.5 %`) %>%
      full_join(
        coef(soil_ADMB) %>%
          as.data.frame() %>% 
          rownames_to_column("term") %>%
          rename(estimate = '.') )

# Maximum aboveground temp (plant canopy)  
  # Fit model
    canopy_ADMB <- glmmADMB::glmmadmb(MaxC ~ RH + MaxWindSpeed +
                                        FuelMoisture + tHa,
                          random = ~ 1|location/year/plot ,
                          data=imp_d,
                          family="gamma",
                          link="log",  
                          admb.opts=glmmADMB::admbControl(shess = TRUE, 
                                                          noinit = TRUE, 
                                                          impSamp = 10))
  # Retrieve regression results
  canopy_terms <-   
    confint(canopy_ADMB) %>%
    as.data.frame %>%
    rownames_to_column("term") %>%
    rename(lwr = `2.5 %`, upr = `97.5 %`) %>%
    full_join(
      coef(canopy_ADMB) %>%
        as.data.frame() %>% 
        rownames_to_column("term") %>%
        rename(estimate = '.') )

# Rate of spread 
  # Fit model 
    ros_ADMB <- glmmADMB::glmmadmb(ros ~ RH + MaxWindSpeed +
                                    FuelMoisture + tHa,
                          random = ~ 1|location/year/plot ,
                          data=imp_d,
                          family="gamma",
                          link="log",  
                          admb.opts=glmmADMB::admbControl(shess = TRUE, 
                                                          noinit = TRUE, 
                                                          impSamp = 10))
  # Retrieve regression results  
    ros_terms <-   
      confint(ros_ADMB) %>%
      as.data.frame %>%
      rownames_to_column("term") %>%
      rename(lwr = `2.5 %`, upr = `97.5 %`) %>%
      full_join(
        coef(ros_ADMB) %>%
          as.data.frame() %>% 
          rownames_to_column("term") %>%
          rename(estimate = '.') )

# Compile results

    response_CIs <-   
      bind_rows(
        mutate(soil_terms, response = 'Temp. at surface'), 
        mutate(canopy_terms, response = 'Temp. above surface'),
        mutate(ros_terms, response = "Rate of spread") 
      ) %>%
      as_tibble()
```

# Results

## Fuel, weather, and fire behavior across locations

```{r data_summary_gg, fig.height=10, fig.width=6}
LongData <- 
  AnalysisData %>%
    select(-LAI) %>%
    pivot_longer(cols = AirTemp:tHa, 
                 names_to = 'var', 
                 values_to = 'value') %>%
    mutate(type = case_when(
                    var %in% c('MaxC', 
                               'SoilMaxC', 
                               'ros') ~ 'Fire behavior', 
                    var %in% c('tHa',
                               'FuelMoisture') ~ 'Fuel', 
                    TRUE ~ 'Weather' ), 
           var = recode(var, 
                        AirTemp = 'Air temperature (C)', 
                        RH = 'Relative humidity (%)', 
                        dpC = 'Dew point (C)', 
                        FuelMoisture = 'Fuel moisture (%)', 
                        tHa = 'Fuel load (t/ha)',
                        MaxC = 'Flame temp (C)', 
                        MaxWindSpeed = 'Wind speed (m/s)', 
                        ros = 'Rate of spread (m/s)', 
                        SoilMaxC = 'Surface soil temp (C)'), 
           location = recode(location, 
                             H = "Hettinger", 
                             CG = "Central Grasslands"))

wx_d <-
LongData %>%
  filter(type == "Weather") %>%
  ggplot(aes(x = location, y= value,
             fill = location)) + theme_bw(14) +
    geom_violin() +
  labs(y = NULL, x = NULL, 
       title = "Weather") + 
  facet_wrap( ~ var, scales = 'free') +
  scale_fill_manual(values = cbPal5[1:2]) + 
  theme(axis.text.x = element_blank(), 
        panel.grid.major.x = element_blank(), 
        legend.position = "none")

fuel_d <-
LongData %>%
  filter(type == "Fuel") %>%
  ggplot(aes(x = location, y= value,
             fill = location)) + theme_bw(14) +
    geom_violin() +
  labs(y = NULL, x = NULL, 
       title = "Fuel") + 
  facet_wrap( ~ var, scales = 'free') +
  scale_fill_manual(values = cbPal5[1:2]) + 
  theme(axis.text.x = element_blank(), 
        panel.grid.major.x = element_blank(), 
        legend.position = "none")

fb_d <-
  LongData %>%
    filter(type == "Fire behavior", 
           value <= 600) %>%
    ggplot(aes(x = location, y= value,
               fill = location)) + theme_bw(14) +
      geom_violin() +
    labs(y = NULL, x = NULL, 
         title = "Fire behavior") + 
    facet_wrap( ~ var, scales = 'free') +
    scale_fill_manual(values = cbPal5[1:2]) + 
    theme(axis.text.x = element_blank(), 
          panel.grid.major.x = element_blank(), 
          legend.position = "bottom")

# Create list of panels
  gl <- lst(wx_d, fuel_d, fb_d)

  grid.arrange(
    grobs = gl,
    widths = c(1),
    heights = c(0.35, 0.25, 0.35)  )
```

## Variables associated with fire behavior

**Rate of spread** significantly increased with windspeed ($z = 2.61, P = 0.01$) and decreased with greater fuel moisture content ($z = -3.92, P < 0.001$). 
**Aboveground temperatures** in the plant canopy were lower with greater fuel moisture content ($z = -5.48, P < 0.001$) and increased with fuel load ($z = 5.16, P < 0.001$) and wind speed ($z = 2.10, P = 0.04$).
**Temperature at the soil surface** had a statistically-significant inverse relationship with relative humidity ($z = -2.77, P < 0.01$), which likely indicates higher intensity fires that burn down through litter. 
There was a marginally-significant trend for higher fuel moisture to reduce temperature at the soil surface, as well ($z = -1.85, P = 0.06$).

Further work should include quantifying total heat exposure via infrared sensors.

```{r CI_gg, fig.width=8, fig.height = 3.75}
# Plot confidence intervals
response_CIs %>%
  filter(term != '(Intercept)') %>%
  as_tibble() %>%
  mutate(term = recode(term, tHa = 'Fuel load', 
                             MaxWindSpeed = 'Wind speed', 
                             RH = 'Relative humidity', 
                             FuelMoisture = 'Fuel moisture'))  %>%
  ggplot(aes(x = reorder(term, abs(estimate), max))) + theme_bw(16) + 
    geom_hline(yintercept = 0, color="black", linetype = 2) +
    geom_errorbar(aes(ymin = lwr, ymax = upr), 
                  size = 1.1, width = 0.2, 
                  color = "blue") +
    geom_point(aes(y = estimate), pch = 21, 
               stroke = 1.5, size = 4,
               color = "blue", fill="lightblue") +
    labs(x = '', 
         y = 'Regression coefficient with 95% CI') + 
    coord_flip(ylim=c(-1.5,2.4)) + 
    facet_wrap(~ response) + 
    theme(axis.text.y = element_text(color="black", size = 14), 
          strip.text = element_text(size = 14), 
          panel.grid.major.y = element_blank())
```

## Multivariate analysis

```{r fit_pca}
# Fire behavior PCA 
  fb_d <- 
    imp_raw %>%
    select(MaxC, ros, SoilMaxC)

  fb_pca <- rda(fb_d ~ 1, 'euc', scale = T)
```

```{r pca_eig}
 summary(fb_pca)$cont$importance %>% pander::pander("Eigenvalues and proportion explained by composite axes (principal components).")
```

```{r}
# Testing for differences between locations 
  envfit(fb_pca ~ location, imp_raw, 
         choices = c(1:3), 
         strata = imp_raw$year, 
         199)$factors
```

No difference in fire behavior patterns between Hettinger and Central Grasslands.

```{r}
# Testing fire weather against PCA
  envfit(fb_pca ~ MaxWindSpeed + AirTemp + dpC + RH + VPD, 
         data = imp_raw, 
         choices = c(1:3), 
         strata = imp_raw$location)
```

RH and VPD had a statistically-significant associations with variation in fire behavior, but even these relationships were weak.

```{r pca_dat}  
  pca_v <- envfit(fb_pca ~ RH + VPD, imp_raw, 
              choices = c(1:3), 
              strata = imp_raw$location)

  # get ord data for ggplotting
  pca_gg <- gg_ordiplot(fb_pca, groups = imp_raw$location, 
                        plot=FALSE)

  pca_spp <- # species only; sites come later
    scores(fb_pca, display = "species") %>%
    as.data.frame %>%
    as_tibble(rownames="response")

  # Vector 
  pca_vd <- scores(pca_v,  "vectors") %>%
            as.data.frame %>% 
            round(3) %>%
            as_tibble(rownames="gradient")

  pca_scores <- lst(species=pca_spp, 
                    vectors=pca_vd)
  pca_scores$spiders <-
    pca_gg$df_spiders %>% 
    rename(PC1 = x, PC2 = y) %>%
    as_tibble()
```

```{r pca_gg, fig.height=4, fig.width=6, fig.cap='Principal Components Analysis of fire behavior data (response variables in red) for prescribed burns on rangeland at Hettinger (H) and Central Grasslands  (CG) Research Extension Centers. No difference between locations and only the relative humidty (RH) gradient had a significant association with variation in fire behavior.' }
# ggplot PCA
  ggplot() + theme_bw(16) + 
    labs(x="PC 1", 
         y="PC 2") + 
    geom_vline(xintercept = 0, lty=3, color="darkgrey") +
    geom_hline(yintercept = 0, lty=3, color="darkgrey") +
    geom_point(data=pca_scores$spiders, 
                        aes(x=PC1, y=PC2, 
                            shape=Group, 
                            colour=Group), 
                        size=2)  +   
    geom_segment(data=pca_scores$spiders, 
                 aes(x=cntr.x, y=cntr.y,
                     xend=PC1, yend=PC2, 
                     color=Group), 
                 size=1.2, 
                 alpha = 0.75) +
    geom_point(data=pca_scores$spiders, 
               aes(x=PC1, y=PC2, fill=Group), 
               colour="grey40", 
               pch=21, 
               size=3, 
               stroke=2, alpha = 0.75) +
    geom_segment(data=pca_scores$vectors, 
                 aes(x=0, y=0, 
                     xend=PC1*1.6, 
                     yend=PC2*1.6),
                 arrow=arrow(length = unit(0.03, "npc")),
                 lwd=1.5) +
  geom_label(data=pca_scores$spiders,
               aes(x=cntr.x,
                   y=cntr.y,
                   label=Group,
                   color=Group),
               fontface="bold", size=4,
               label.size = 0,
               label.r = unit(0.5, "lines")) +
    geom_text(data=pca_scores$vectors,
              aes(x=PC1*1.75, 
                  y=PC2*1.75, 
                  label=gradient), 
              nudge_x = 0.06, 
              nudge_y =-0.05, 
              size=6, fontface="bold") + 
    geom_label(data=pca_scores$species, 
                aes(x=PC1/3, 
                    y=PC2/3, 
                    label=response), 
                label.padding=unit(0.25,"lines"),
                label.size = 0.5, 
                fontface="bold", 
               color="darkred") +
    scale_color_manual(values = cbPal5[1:2]) + 
    scale_fill_manual(values = cbPal5[1:2])  +
    theme(panel.grid=element_blank(), 
          legend.position="none")

```

# Script

```{r script, ref.label=knitr::all_labels(!label %in% c('setup', 'dist_gg', 'data_summary_gg', 'CI_gg', 'pca_eig', 'pca_dat', 'pca_gg') ), echo=TRUE, eval=FALSE}
```

### 2021-02-08 r script/ADMB_regression.R

@@ -1,0 +1,88 @@
+pacman::p_load(tidyverse)
+
+# Load imputed dataset 
+load(paste0(getwd(), '/r objects/imp_sc.Rdata'))
+
+
+
+# Max soil surface temperature 
+  # Fit model
+    soil_ADMB <- glmmADMB::glmmadmb(SoilMaxC ~ RH + MaxWindSpeed +
+                            FuelMoisture + tHa,
+                          random = ~ 1|location/year/plot ,
+                          data = filter(imp_sc, SoilMaxC <= 3.0),
+                          family="gamma",
+                          link="log",  
+                          admb.opts=glmmADMB::admbControl(shess = TRUE, 
+                                                          noinit = TRUE, 
+                                                          impSamp = 10) )
+  # Retrieve regression results
+    soil_terms <-   
+      confint(soil_ADMB) %>%
+      as.data.frame %>%
+      rownames_to_column("term") %>%
+      rename(lwr = `2.5 %`, upr = `97.5 %`) %>%
+      full_join(
+        coef(soil_ADMB) %>%
+          as.data.frame() %>% 
+          rownames_to_column("term") %>%
+          rename(estimate = '.') )
+
+# Maximum aboveground temp (plant canopy)  
+  # Fit model
+    canopy_ADMB <- glmmADMB::glmmadmb(MaxC ~ RH + MaxWindSpeed +
+                                        FuelMoisture + tHa,
+                          random = ~ 1|location/year/plot ,
+                          data=imp_sc,
+                          family="gamma",
+                          link="log",  
+                          admb.opts=glmmADMB::admbControl(shess = TRUE, 
+                                                          noinit = TRUE, 
+                                                          impSamp = 10))
+  # Retrieve regression results
+  canopy_terms <-   
+    confint(canopy_ADMB) %>%
+    as.data.frame %>%
+    rownames_to_column("term") %>%
+    rename(lwr = `2.5 %`, upr = `97.5 %`) %>%
+    full_join(
+      coef(canopy_ADMB) %>%
+        as.data.frame() %>% 
+        rownames_to_column("term") %>%
+        rename(estimate = '.') )
+  
+# Rate of spread 
+  # Fit model 
+    ros_ADMB <- glmmADMB::glmmadmb(ros ~ RH + MaxWindSpeed +
+                                    FuelMoisture + tHa,
+                          random = ~ 1|location/year/plot ,
+                          data=imp_sc,
+                          family="gamma",
[…]

### 2021-02-08 r script/ConnectingData.R

@@ -1,45 +1,0 @@
-pacman::p_load(tidyverse)
-
-load("./r objects/wx_d.Rdata")
-load("./r objects/ROS.Rdata")
-load("./r objects/fuels.Rdata")
-load("./r objects/temps.Rdata")
-load("./r objects/soil.Rdata")
-
-# Summarize to array scale 
-
-  fuels_sum <- 
-    fuels %>%
-      select(-date, -location, -block, -burned) %>%
-      pivot_longer(cols = (LAI:FuelMoisture), 
-                   names_to = "response", 
-                   values_to = "value") %>%
-      group_by(FireCode, plot, array, response) %>%
-        summarize(value = mean(value)) %>%
-      ungroup() %>%
-      pivot_wider(names_from = response, 
-                  values_from = value)
-
-# Summarize & bring in weather data by the hour 
-
-  DataCombined <- 
-    temps %>% 
-    group_by(FireCode,location, plot, array) %>%
-    summarize(timestamp = mean(timestamp), 
-              tempC = mean(tempC)) %>%
-    ungroup() %>%
-    mutate(DayHour = format(timestamp, "%Y-%d-%m %H")) %>% 
-    unite(BurnHourID, c(location, DayHour), sep='_') %>%
-    select(-timestamp) %>%
-    left_join(wx_d, 
-              by = 'BurnHourID')%>%
-    left_join(fuels_sum, 
-              by = c('FireCode', 'plot', 'array')) %>%
-    select(-BurnHourID) %>%
-    left_join(ROS, 
-              by = c('FireCode', 'plot', 'array')) %>%
-    left_join(soil, 
-              by = c('FireCode', 'plot', 'array')) %>%
-    separate(FireCode, c('location','block','pasture','patch','year')) 
-  
-  # save(DataCombined, file="./data/DataCombined.Rdata")

### 2021-02-08 r script/CustomThemes.R

@@ -1,0 +1,161 @@
+cbPal5 <- c("#009E73","#E69F00",  "#0072B2", "#CC79A7", "#999999")
+cbPal8 <- c("#66c2a5","#fc8d62",  "#8da0cb", "#e78ac3", "#a6d854", "#ffd92f", "#e5c494", "#b3b3b3" )
+ColPts5 <- c(22,24,21,25,23)
+
+# A general theme for book graphics. Size = 18
+
+theme_cb1 <- function (base_size = 18, base_family = "") 
+{
+  theme_grey(base_size = base_size, base_family = base_family) %+replace% 
+    theme(axis.text = element_text(size = rel(0.9)), 
+          axis.title = element_text(face="bold"),
+          axis.ticks = element_line(colour = "black"), 
+          strip.text = element_text(face="bold"),
+          legend.key = element_rect(colour = NA),
+          panel.background = element_rect(fill = "white", colour = NA), 
+          panel.border = element_rect(fill = NA, colour = "grey50"), 
+          panel.grid.major = element_line(colour = "grey90", size = 0.2), 
+          panel.grid.minor = element_line(colour = "grey98", size = 0.5), 
+          strip.background = element_rect(fill = "#A6CEE3", colour = "grey50", size = 0.2))
+}
+
+# A general theme for book graphics; . Size = 18
+
+theme_CF <- function (base_size = 18, base_family = "") 
+{
+  # CF = "clean facet." No shading or lines
+  theme_grey(base_size = base_size, base_family = base_family) %+replace% 
+    theme(axis.text = element_text(size = rel(0.9)), 
+          axis.title = element_text(face="bold"),
+          axis.ticks = element_line(colour = "black"), 
+          strip.text = element_text(face="plain", 
+                                    margin = 
+                                      margin(0,0,1.5,0, "mm")),
+          legend.key = element_rect(colour = NA),
+          panel.background = element_rect(fill = "white", colour = NA), 
+          panel.border = element_rect(fill = NA, colour = "grey50"), 
+          panel.grid.major = element_line(colour = "grey90", size = 0.2), 
+          panel.grid.minor = element_line(colour = "grey98", size = 0.5), 
+          strip.background = element_rect(fill = NA, size = NA))
+}
+
+theme_CFcatX <- function (base_size = 18, base_family = "") 
+{
+    # CF = "clean facet." No shading or lines
+    # catX = no minor X gridlines for categorical variables
[…]

### 2021-02-08 r script/DataAndDistributions.R

@@ -1,0 +1,48 @@
+pacman::p_load(tidyverse)
+
+load(paste0(FilePath, "data/AnalysisData.Rdata"))
+
+GGally::ggpairs(AnalysisData, columns = 9:17)
+
+# Distribution
+
+AnalysisData %>% 
+  ggplot(aes(x=ros)) + theme_bw(14) + 
+  geom_histogram(aes(y=..density..),      
+                 binwidth=0.5,
+                 colour="black", 
+                 fill="lightgreen") +
+  geom_density(alpha=0.2, 
+               fill="#FF6666") 
+
+AnalysisData %>% 
+  ggplot(aes(x=MaxC)) + theme_bw(14) + 
+  geom_histogram(aes(y=..density..),      
+                 binwidth=10,
+                 colour="black", 
+                 fill="lightgreen") +
+  geom_density(alpha=0.2, 
+               fill="#FF6666") 
+
+AnalysisData %>% 
+  ggplot(aes(x=SoilMaxC)) + theme_bw(14) + 
+  geom_histogram(aes(y=..density..),      
+                 binwidth=10,
+                 colour="black", 
+                 fill="lightgreen") +
+  geom_density(alpha=0.2, 
+               fill="#FF6666") 
+
+AnalysisData %>% 
+  ggplot(aes(x=MaxC, y = ros, color=location)) + theme_bw(14) +
+  geom_point(alpha=0.5)
+
+ggplot(AnalysisData, aes(x=SoilMoisture, y=FuelMoisture, color=location)) +
+  theme_bw(14) + 
+  geom_smooth(method = 'lm') + 
+  geom_point(alpha=0.5) +
+  coord_cartesian(ylim = c(0,200))
+
+AnalysisData %>% 
+  ggplot(aes(x=dbC, y = ros, color=location)) + theme_bw(14) +
+  geom_point(alpha=0.5)

### 2021-02-08 r script/DataLoading.R

@@ -1,0 +1,95 @@
+pacman::p_load(tidyverse, lubridate)
+
+FilePath = getwd()
+
+  AllData <-  
+    read_csv(paste0(FilePath, "data/fromMZ/CompiledData2.csv")) %>%
+    filter(location != "OAK") %>%
+    mutate(date = as.Date(date, format = "%m/%d/%Y"),
+           L = str_remove(location, "REC"), 
+           B = str_sub(block, 1,3), 
+           Ps = str_replace(pasture, "[.]", ""), 
+           Ps = str_sub(Ps, 1,2), 
+           patch = str_replace(patch, "[.]", ""),
+           y = format(date, "%y")) %>%
+    unite("FireCode", c(L,B,Ps,patch,y), sep=".") %>%
+    mutate(time = str_remove(MaxTempTime, "[.]+[0-9]"))%>%
+    unite(timestamp, c(date, time), sep = " ") %>%
+    mutate(timestamp = as.POSIXct(timestamp, format = "%Y-%m-%d %H:%M:%S")) %>%
+    select(FireCode, timestamp, plot, array, TC, MaxC, 
+           AirTemp, RH, dpC, MaxWindSpeed, 
+           LAI, FMC, KgHa) 
+
+# Isolate soil surface temperature (TC 4)
+
+  SoilTemp <-
+    filter(AllData, TC == 4) %>%
+    select(FireCode, plot, array, MaxC) %>%
+    rename(SoilC = MaxC)
+
+# Summarize array-level data 
+  DataMeans <- 
+    AllData %>%
+    filter(TC %in% c('1', '2', '3')) %>% 
+    select(-timestamp) %>%
+    pivot_longer(cols = c(MaxC:KgHa), 
+                 names_to = "var",
+                 values_to = "value") %>%
+    group_by(FireCode, plot, array, var) %>%
+    summarize(Mean = mean(value) ) %>%
+    ungroup() %>%
+    pivot_wider(names_from = var, 
+                values_from = Mean)
+  
+# Calculate Hot-Dry-Windy Index
+  DataMeans <- 
+    DataMeans %>%
+      mutate(e  = 6.11 * (10 ^ ( (7.5 * dpC)/ (237.3 + dpC) ) ), 
+             es = 6.11 * (10 ^ ( (7.5 * AirTemp)/ (237.3 + AirTemp) ) ), 
+             VPD = es - e) %>%
+      select(-e, -es)
+
+
+# Calculate ROS 
+
+  D = 1   # Distance between thermocouples (m)
+  
[…]

### 2021-02-08 r script/GLMER_regression.R

@@ -1,0 +1,112 @@
+pacman::p_load(tidyverse)
+
+# Load imputed dataset 
+  load(paste0(getwd(), '/r objects/imp_d.Rdata'))
+
+  
+# Mixed-effect regression models 
+#
+# Rate of spread
+#   
+  # Fit model 
+    ros_all <- with(imp.d, suppressMessages(
+      lme4::glmer(ros ~ RH + MaxWindSpeed + 
+                    FuelMoisture + LAI + 
+                    (1|location/block/year/plot), 
+                  family=Gamma(link = "log"), 
+                  control=lme4::glmerControl(optimizer="bobyqa", 
+                                             optCtrl=list(maxfun=100000)) )) )
+    
+  # Get terms
+    ros_terms <- 
+      suppressWarnings(confint(ros_all,method="boot",nsim=199)) %>%
+      as.data.frame %>%
+      rownames_to_column("term") %>%
+      slice(6:10) %>%
+      rename(lwr = `2.5 %`, upr = `97.5 %`) %>%
+      full_join(
+        lme4::fixef(ros_all) %>%
+          as.data.frame() %>% 
+          rownames_to_column("term") %>%
+          rename(estimate = '.') ) 
+
+
+#
+# Maximum canopy temperature  
+#
+# Fit model 
+temp_all <- with(imp.d, suppressMessages(
+  lme4::glmer(MaxC ~ RH + MaxWindSpeed +  
+                FuelMoisture + LAI + 
+                (1|location/block/year/plot), 
+              family=Gamma(link = "log"), 
+              control=lme4::glmerControl(optimizer="bobyqa", 
+                                         optCtrl=list(maxfun=100000)) )) )
+temp_terms <-   
+  suppressWarnings(confint(temp_all,method="boot",nsim=199)) %>%
+  as.data.frame %>%
+  rownames_to_column("term") %>%
+  slice(6:10) %>%
+  rename(lwr = `2.5 %`, upr = `97.5 %`) %>%
+  full_join(
+    lme4::fixef(temp_all) %>%
+      as.data.frame() %>% 
+      rownames_to_column("term") %>%
+      rename(estimate = '.') ) 
+
+#
+# Maximum soil surface temperature  
+#
+# Fit model 
+soil_glmer <- with(imp_d, suppressMessages(
+                lme4::glmer(SoilMaxC ~ RH + MaxWindSpeed +  
+                    FuelMoisture + LAI + 
[…]

### 2021-02-08 r script/LoadingCompiledData.R

@@ -1,78 +1,0 @@
-pacman::p_load(tidyverse, lubridate)
-
-# wd_fp = "C:/Users/devan/GitHubProjects/SpatialFireBehavior/" # home PC
-
-wd_fp = getwd() # machine-independent?
-
-# Pull data from .csv created from MZ's compiled Excel file (tab = All)
-
-  AllData <-  
-    read_csv(paste0(wd_fp, "/data/fromMZ/CompiledData2.csv")) %>%
-    filter(location != "OAK") %>%
-      mutate(date = as.Date(date, format = "%m/%d/%Y"),
-               L = str_remove(location, "REC"), 
-               B = str_sub(block, 1,3), 
-               Ps = str_replace(pasture, "[.]", ""), 
-               Ps = str_sub(Ps, 1,2), 
-               patch = str_replace(patch, "[.]", ""),
-               y = format(date, "%y")) %>%
-        unite("FireCode", c(L,B,Ps,patch,y), sep=".") %>%
-    mutate(time = str_remove(MaxTempTime, "[.]+[0-9]"))%>%
-    unite(timestamp, c(date, time), sep = " ") %>%
-    mutate(timestamp = as.POSIXct(timestamp, format = "%Y-%m-%d %H:%M:%S")) %>%
-        select(FireCode, timestamp, plot, array, TC, MaxC, 
-               AirTemp, RH, dpC, MaxWindSpeed, 
-               LAI, FMC, KgHa) 
-
-# Isolate soil surface temperature (TC 4)
-  
-  SoilTemp <-
-    filter(AllData, TC == 4) %>%
-      select(FireCode, plot, array, MaxC) %>%
-        rename(SoilC = MaxC)
-  
-# Summarize array-level data 
-  DataMeans <- 
-    AllData %>%
-      filter(TC %in% c('1', '2', '3')) %>% 
-      select(-timestamp) %>%
-      pivot_longer(cols = c(MaxC:KgHa), 
-                   names_to = "var",
-                   values_to = "value") %>%
-      group_by(FireCode, plot, array, var) %>%
-        summarize(Mean = mean(value) ) %>%
-      ungroup() %>%
-      pivot_wider(names_from = var, 
-                  values_from = Mean) 
-    
-  
-# Calculate ROS 
-
-  D = 1   # Distance between thermocouples (m)
-  
-  ROS <- 
-    AllData %>%
-      filter(TC %in% c('1', '2', '3')) %>%
[…]

### 2021-02-08 r script/LoadingData.R

@@ -1,102 +1,0 @@
-pacman::p_load(tidyverse, readxl)
-
-wd_fp = "C:/Users/Devan.McGranahan/GoogleDrive/FireScienceDIY/SpatialFireBehavior/data/"
-
-xl_fp = paste0(wd_fp, "/AllLocations.xlsx")
-
-
-# Soil temp data 
-
-soil <-  
-  read_xlsx("./data/fromMZ/Research_Fire_Database.xlsx", "Soil_TC") %>%
-    select(-AirTemp, -RH, -WindSpd, -MaxWindSpd, -WindDir) %>%
-    mutate(date = as.Date(date),
-           L = str_remove(location, "REC"), 
-           B = str_sub(block, 1,3), 
-           Ps = str_replace(pasture, "[.]", ""), 
-           Ps = str_sub(Ps, 1,2), 
-           y = format(date, "%y")) %>%
-  unite("FireCode", c(L,B,Ps,patch,y), sep=".") %>%
-  select(FireCode, plot, array, SoilMaxC) %>%
-    filter(SoilMaxC != is.na(.))
-
-# save(soil, file="./r objects/soil.Rdata")
-
-# Thermocouple data 
-temps <- 
-  read_csv(paste0(wd_fp, "ThermocoupleData.csv")) %>%
-  filter(tempC != is.na(.)) %>%
-    mutate(time = str_remove(time, "[.]+[0-9]"))%>%
-    unite(timestamp, c(date, time), sep = " ") %>%
-  mutate(timestamp = as.POSIXct(timestamp, format = "%m-%d-%Y %H:%M:%S")) %>%
-    mutate(L = str_remove(location, "REC"), 
-           B = str_sub(block, 1,3), 
-           Ps = str_replace(pasture, "[.]", ""), 
-           Ps = str_sub(Ps, 1,2), 
-           y = format(timestamp, "%y"),
-           problematic = as.character(problematic)) %>%
-    unite("FireCode", c(L,B,Ps,patch,y), sep=".") %>%
-    filter(is.na(problematic) ) %>%
-    select(FireCode, timestamp, location, block, plot, array, 
-           TC, tempC) 
-
-# save(temps, file="./r objects/temps.Rdata")
-
-# Fuels data 
-fuels <- 
-  read_xlsx(xl_fp, "fuels") %>%
-    mutate(date = paste(str_sub(date, 1,4), 
-                        str_sub(date, 5,6),
-                        str_sub(date, 7,8),
[…]

### 2021-02-08 r script/MultivariateAnalysis.R

@@ -1,0 +1,111 @@
+pacman::p_load(tidyverse, vegan) 
+source("https://raw.githubusercontent.com/devanmcg/IntroRangeR/master/11_IntroMultivariate/ordinationsGGplot.R")
+
+# Fire behavior PCA 
+  fb_d <- 
+    imp_raw %>%
+    select(MaxC, ros, SoilMaxC) 
+  
+  fb_pca <- rda(fb_d ~ 1, 'euc', scale = T)
+  
+  summary(fb_pca)$cont$importance 
+  
+  plot(fb_pca)
+  ordispider(fb_pca, groups =imp_raw$location, label = T)
+  text(fb_pca, "species")
+  
+  envfit(fb_pca ~ location, imp_raw, 
+         choices = c(1:3), 
+         strata = imp_raw$year, 
+         199)
+  envfit(fb_pca ~ MaxWindSpeed + AirTemp + dpC + RH + VPD + HDWI, 
+         data = imp_raw, 
+         choices = c(1:3), 
+         strata = imp_raw$location)  
+
+  
+  # get ord data for ggplotting
+  pca_gg <- gg_ordiplot(fb_pca, groups = imp_raw$location, 
+                        plot=FALSE)
+  
+  pca_spp <- # species only; sites come later
+    scores(fb_pca, display = "species") %>%
+    as.data.frame %>%
+    as_tibble(rownames="response")
+  
+  # Vectors 
+  pca_v <- envfit(fb_pca ~ RH + VPD, imp_raw, 
+                  choices = c(1:3), 
+                  strata = imp_raw$location)
+  pca_vd <- scores(pca_v,  "vectors") %>%
+            as.data.frame %>% 
+            round(3) %>%
+            as_tibble(rownames="gradient") 
+  
+  pca_scores <- lst(species=pca_spp, 
+                    vectors=pca_vd)
+  pca_scores$spiders <-
+    pca_gg$df_spiders %>% 
+    rename(PC1 = x, PC2 = y) %>%
+    as_tibble
+
+  # gg plot
+  
+  ggplot() + theme_bw(16) + 
+    labs(x="PC 1", 
+         y="PC 2") + 
+    geom_vline(xintercept = 0, lty=3, color="darkgrey") +
+    geom_hline(yintercept = 0, lty=3, color="darkgrey") +
+    geom_point(data=pca_scores$spiders, 
+                        aes(x=PC1, y=PC2, 
+                            shape=Group, 
+                            colour=Group), 
[…]

### 2021-02-08 r script/RandomEffectsRegression.R

@@ -1,0 +1,56 @@
+load(paste0(FilePath, '/r objects/imp_sc.Rdata'))
+
+imp_sc 
+# RE.gam.var <- function(x) {
+#   require(lme4)
+  var.mod <- lme4::glmer(ros ~ (1|location:year) +
+                         (1|location:pasture:year)
+                         (1|location:pasture:plot:year), 
+                         Gamma(link = "log"), 
+                         data=imp_sc) 
+  transect.var <- (attr(VarCorr(var.mod)$'transect.num:year',"stddev"))^2
+  year.var <- (attr(VarCorr(var.mod)$'year',"stddev"))^2 
+  var.results<-array(NA,c(1,2))
+  colnames(var.results)<-c("transect.var","year.var")
+  var.results[1,] <- round(c(transect.var, year.var), 4) 
+  return(var.results)	
+  # }
+  
+  imp_chr <-  imp_raw  %>% 
+                mutate(across(location:array, as.character)) 
+                
+  var.mod <- lme4::glmer(ros ~ 1 + (1|pasture:plot:array:year) +
+                                (1|pasture:plot:year) +
+                                (1|location) , 
+                         Gamma(link = "log"), 
+                         data=imp_chr, 
+                         control=lme4::glmerControl(optimizer="bobyqa", 
+                                                    optCtrl=list(maxfun=100000)) ) 
+  
+  ros_var <- glmmADMB::glmmadmb(ros ~ 1 ,
+                                 random = ~ (1|pasture:plot:array:year) +
+                                            (1|pasture:plot:year) +
+                                            (1|location),
+                                 data=imp_sc,
+                                 family="gamma",
+                                 link="log") 
+  attr(lme4::VarCorr(ros_var)$'location:year',"StdDev")
+  
+  lme4::VarCorr(ros_var)[3]  
+  
+  lme4::VarCorr(ros_var)  
+  
+  vc <- unlist(lme4::VarCorr(ros_var)) %>%
+    as.data.frame( ) %>%
+    rownames_to_column('scale')
+    colnames(vc) <- c('array', 'plot', 'location')
+  
+  unique(imp_chr$plot)  
+  
+  var_mod <- lme4::glmer(ros ~ 1 + (1|FireCode:plot:array) +
+                           (1|FireCode:plot)  , 
+                         Gamma(link = "log"), 
+                         data=imp_chr, 
+                         control=lme4::glmerControl(optimizer="bobyqa", 
+                                                    optCtrl=list(maxfun=100000)) )
+  

### 2021-02-08 r script/RegressionAnalysis.R

@@ -1,165 +1,0 @@
-pacman::p_load(tidyverse)
-
-setwd("C:/Users/Devan.McGranahan/GoogleDrive/Research/Projects/SpatialFireBehavior")
-
-load("./data/DataCombined.Rdata")
-
-AnalysisData <- 
-  DataCombined %>%
-    filter( ros <= 40, 
-            tempC >= 40) %>%
-      mutate(FuelMoisture = ifelse(FuelMoisture >= 0, 
-                                    FuelMoisture, NA), 
-             FuelMoisture = FuelMoisture * 100)
-  
-GGally::ggpairs(AnalysisData, columns = 8:19)
-
-# Distribution
-
-AnalysisData %>% 
-  ggplot(aes(x=ros)) + theme_bw(14) + 
-  geom_histogram(aes(y=..density..),      
-                 binwidth=0.5,
-                 colour="black", 
-                 fill="lightgreen") +
-  geom_density(alpha=0.2, 
-               fill="#FF6666") 
-
-AnalysisData %>% 
-  ggplot(aes(x=tempC)) + theme_bw(14) + 
-  geom_histogram(aes(y=..density..),      
-                 binwidth=10,
-                 colour="black", 
-                 fill="lightgreen") +
-  geom_density(alpha=0.2, 
-               fill="#FF6666") 
-
-AnalysisData %>% 
-  ggplot(aes(x=SoilMaxC)) + theme_bw(14) + 
-  geom_histogram(aes(y=..density..),      
-                 binwidth=10,
-                 colour="black", 
-                 fill="lightgreen") +
-  geom_density(alpha=0.2, 
-               fill="#FF6666") 
-
-AnalysisData %>% 
-  ggplot(aes(x=tempC, y = ros, color=location)) + theme_bw(14) +
-    geom_point(alpha=0.5)
-
-  ggplot(AnalysisData, aes(x=SoilMoisture, y=FuelMoisture, color=location)) + 
-    geom_smooth(method = 'lm') + 
-    geom_point(alpha=0.5) +
-    coord_cartesian(ylim = c(0,200))
-  
-  AnalysisData %>% 
-    ggplot(aes(x=dbC, y = ros, color=location)) + theme_bw(14) +
-    geom_point(alpha=0.5)
-
-# MICE regression 
-
-pacman::p_load(mice, broom.mixed)
-
-  # View pattern of missing data
-    AnalysisData %>%
-        md.pattern() 
-
[…]

### 2021-02-08 r script/RegressionAnalysis2.R

@@ -1,163 +1,0 @@
-pacman::p_load(tidyverse)
-
-source("./r script/LoadingCompiledData.R")
-
-AnalysisData <- 
-  DataCombined %>%
-    filter( ros <= 40, 
-            tempC >= 40) %>%
-      mutate(FuelMoisture = ifelse(FuelMoisture >= 0, 
-                                    FuelMoisture, NA), 
-             FuelMoisture = FuelMoisture * 100)
-  
-GGally::ggpairs(AnalysisData, columns = 8:19)
-
-# Distribution
-
-AnalysisData %>% 
-  ggplot(aes(x=ros)) + theme_bw(14) + 
-  geom_histogram(aes(y=..density..),      
-                 binwidth=0.5,
-                 colour="black", 
-                 fill="lightgreen") +
-  geom_density(alpha=0.2, 
-               fill="#FF6666") 
-
-AnalysisData %>% 
-  ggplot(aes(x=tempC)) + theme_bw(14) + 
-  geom_histogram(aes(y=..density..),      
-                 binwidth=10,
-                 colour="black", 
-                 fill="lightgreen") +
-  geom_density(alpha=0.2, 
-               fill="#FF6666") 
-
-AnalysisData %>% 
-  ggplot(aes(x=SoilMaxC)) + theme_bw(14) + 
-  geom_histogram(aes(y=..density..),      
-                 binwidth=10,
-                 colour="black", 
-                 fill="lightgreen") +
-  geom_density(alpha=0.2, 
-               fill="#FF6666") 
-
-AnalysisData %>% 
-  ggplot(aes(x=tempC, y = ros, color=location)) + theme_bw(14) +
-    geom_point(alpha=0.5)
-
-  ggplot(AnalysisData, aes(x=SoilMoisture, y=FuelMoisture, color=location)) + 
-    geom_smooth(method = 'lm') + 
-    geom_point(alpha=0.5) +
-    coord_cartesian(ylim = c(0,200))
-  
-  AnalysisData %>% 
-    ggplot(aes(x=dbC, y = ros, color=location)) + theme_bw(14) +
-    geom_point(alpha=0.5)
-
-# MICE regression 
-
-pacman::p_load(mice, broom.mixed)
-
-  # View pattern of missing data
-    AnalysisData %>%
-        md.pattern() 
-
-  # Calculate imputed datasets
-    imp <- AnalysisData %>% 
-                mutate_at(vars(plot, array), as.character) %>%
[…]

### 2021-02-08 r script/imputation.R

@@ -1,0 +1,33 @@
+pacman::p_load(tidyverse)
+
+load(paste0(FilePath, "/data/AnalysisData.Rdata"))
+
+# MICE regression 
+
+pacman::p_load(mice, broom.mixed)
+
+#
+# Imputations
+#
+  # View pattern of missing data
+    AnalysisData %>%
+        md.pattern() 
+
+  # Calculate imputed datasets
+    imp <- AnalysisData %>% 
+                mutate_at(vars(plot, array), as.character) %>%
+                 mice(m=500, seed = 23109, print=F)
+    
+    imp_raw <- 
+      complete(imp) %>%
+              as_tibble() %>% 
+                mutate(across(location:array, as.factor)) 
+    
+    imp_sc <- 
+      imp_raw %>%
+       mutate_at(vars(AirTemp:tHa), ~as.numeric(scale(., center=F)))
+  
+    # save(imp_raw, file = paste0(FilePath, '/r objects/imp_raw.Rdata'))
+    # save(imp_sc, file = paste0(FilePath, '/r objects/imp_sc.Rdata'))
+
+    

### 2021-02-08 writeups/FuelsWeatherFirebehavior.Rmd
add an .md file?

@@
   html_document:
+    keep_md: true
     toc: true

### 2021-02-08 writeups/FuelsWeatherFirebehavior.Rmd
remove md file

--- 2021-02-08 writeups/FuelsWeatherFirebehavior.Rmd
@@
   html_document:
+    keep_md: false
     toc: true

### 2021-02-09 writeups/FuelsWeatherFirebehavior.Rmd
Update FuelsWeatherFirebehavior.Rmd

@@
  AllData <-  
-    read_csv(paste0(FilePath, "data/fromMZ/CompiledData2.csv")) %>%
+    read_csv(paste0(FilePath, "/data/fromMZ/CompiledData2.csv")) %>%
     filter(location != "OAK") %>%

### 2021-02-09 r script/DataLoading.R
Update DataLoading.R

@@ -3,7 +3,7 @@ pacman::p_load(tidyverse, lubridate)
 FilePath = getwd()
 
   AllData <-  
-    read_csv(paste0(FilePath, "data/fromMZ/CompiledData2.csv")) %>%
+    read_csv(paste0(FilePath, "/data/fromMZ/CompiledData2.csv")) %>%
     filter(location != "OAK") %>%
     mutate(date = as.Date(date, format = "%m/%d/%Y"),
            L = str_remove(location, "REC"), 

### 2021-02-12 writeups/FuelsWeatherFirebehavior.Rmd
update with new data from KY

--- 2021-02-08 writeups/FuelsWeatherFirebehavior.Rmd
@@
 FilePath = "C:/Users/devan/GitHubProjects/SpatialFireBehavior"
-load(paste0(FilePath, '/data/AnalysisData.Rdata'))
+AnalysisData <- readr::read_csv(paste0(FilePath, "/data/AnalysisDataKY.csv"))
 load(paste0(FilePath, '/r objects/response_CIs.Rdata'))
 load(paste0(FilePath, '/r objects/imp_raw.Rdata'))
+load(paste0(FilePath, '/r objects/pca_ev.Rdata'))
 source(paste0(FilePath, '/r script/CustomThemes.R'))
@@
 ```{r data_prep, eval=FALSE}
+###
+# Note: Data prep steps included for reference. 
+# Data used in analysis, AnalysisDataKY.csv, has been modified outside of
+# this script to update missing data from another spreadsheet. 
  AllData <-  
-    read_csv(paste0(FilePath, "data/fromMZ/CompiledData2.csv")) %>%
+    read_csv(paste0(FilePath, "/data/fromMZ/CompiledData2.csv")) %>%
     filter(location != "OAK") %>%
@@
                remove = F)
+###
 ```
@@
 # Scatterplot matrix  
-  GGally::ggpairs(AnalysisData, 
-                  columns = 9:18, 
+AnalysisData %>%
+  select(-LAI) %>%
+  GGally::ggpairs(columns = 9:18, 
                   upper = 'blank')
@@
 Therefore I used the multiple imputation method in the **mice** package to fill in missing values. 
-The method creates 100 datasets, each with different but reasonable values for the missing data based on patterns in existing data, and performs the regression analysis on each. 
+The method creates 500 datasets, each with different but reasonable values for the missing data based on patterns in existing data, and performs the regression analysis on each. 
 These results are pooled into a single set of results.
+
+```{r imp, eval=FALSE}
+# Calculate imputed datasets
+  imp <- AnalysisData %>% 
+          select(-LAI) %>%
+              mutate(across(plot:array, ~as.character(.))) %>%
+               mice(m=500, seed = 23109, print=F)
+
+# Return completed datset   
+  imp_raw <- 
+    complete(imp) %>%
+            as_tibble() %>% 
+              mutate(across(location:array, as.factor))
+
+# Scale response variables for regression analysis   
+  imp_sc <- 
+    imp_raw %>%
+     mutate_at(vars(AirTemp:tHa), ~as.numeric(scale(., center=F)))
+```
 
@@
                           random = ~ 1|location/year/plot ,
-                          data=filter(imp_d, SoilMaxC <= 3.0),
+                          data=filter(imp_sc, SoilMaxC <= 3.9),
                           family="gamma",
@@
                           random = ~ 1|location/year/plot ,
-                          data=imp_d,
+                          data=imp_sc,
                           family="gamma",
@@
 
-**Rate of spread** significantly increased with windspeed ($z = 2.61, P = 0.01$) and decreased with greater fuel moisture content ($z = -3.92, P < 0.001$). 
-**Aboveground temperatures** in the plant canopy were lower with greater fuel moisture content ($z = -5.48, P < 0.001$) and increased with fuel load ($z = 5.16, P < 0.001$) and wind speed ($z = 2.10, P = 0.04$).
-**Temperature at the soil surface** had a statistically-significant inverse relationship with relative humidity ($z = -2.77, P < 0.01$), which likely indicates higher intensity fires that burn down through litter.
-There was a marginally-significant trend for higher fuel moisture to reduce temperature at the soil surface, as well ($z = -1.85, P = 0.06$).
-
-Further work should include quantifying total heat exposure via infrared sensors.
+**Rate of spread** significantly increased with windspeed ($z = 3.24, P < 0.01$) and decreased with greater fuel moisture content ($z = -5.84, P < 0.001$). 
+**Aboveground temperatures** in the plant canopy were lower with greater fuel moisture content ($z = -6.45, P < 0.001$) and increased with fuel load ($z = 5.49, P < 0.001$) and wind speed ($z = 2.82, P < 0.01$).
+**Temperature at the soil surface** had a statistically-significant inverse relationship with relative humidity ($z = -2.52, P = 0.01$) and fuel moisture ($z = -3.22, P < 0.01$), which likely indicates higher-intensity fires that burn down through litter.
 
@@
 
-No difference in fire behavior patterns between Hettinger and Central Grasslands.
+No difference in fire behavior patterns between Hettinger and Central Grasslands; this is expected.
 
-```{r}
+```{r envfit, eval=FALSE}
 # Testing fire weather against PCA
@@
 
-RH and VPD had a statistically-significant associations with variation in fire behavior, but even these relationships were weak.
+```{r echo=FALSE}
+pca_ev
+```
+
+RH and VPD had statistically-significant associations with variation in fire behavior, but even these relationships were weak.
 
 ```{r pca_dat}  
-  pca_v <- envfit(fb_pca ~ RH + VPD, imp_raw, 
-              choices = c(1:3), 
-              strata = imp_raw$location)
-
   # get ord data for ggplotting
@@
   # Vector 
-  pca_vd <- scores(pca_v,  "vectors") %>%
+  pca_vd <- scores(pca_ev,  "vectors") %>%
                 as.data.frame %>% 
-            round(3) %>%
-            as_tibble(rownames="gradient")
+                rownames_to_column("gradient") %>%
+              filter(gradient %in% c("RH", "VPD")) %>%
+              select(-PC3) %>%
+            as_tibble( )
 
@@
 ```{r pca_gg, fig.height=4, fig.width=6, fig.cap='Principal Components Analysis of fire behavior data (response variables in red) for prescribed burns on rangeland at Hettinger (H) and Central Grasslands  (CG) Research Extension Centers.
-No difference between locations and only the relative humidty (RH) gradient had a significant association with variation in fire behavior.' }
+No difference between locations and only the vapor pressure deficit (VPD) and relative humidty (RH) gradients had significant associations with variation in fire behavior.' }
 # ggplot PCA
@@
 
+It is interesting that the gradient in temperature at the soil surface (SoilMaxC) is orthogonal to both canopy temperature and rate of spread (MaxC and ros, respectively). 
+This confirms trends observed in the regression analysis that the conditions most conducive to fast, high-intensity fires are not the same that result in surface heating.
+
+Managers might be able to use high windspeeds to mitigate soil heating when fuels are dry, or focus on conditions with low RH to ensure litter burns down to the soil surface.
+
 # Script

### 2021-02-12 r script/ADMB_regression.R

@@ -10,7 +10,7 @@ load(paste0(getwd(), '/r objects/imp_sc.Rdata'))
     soil_ADMB <- glmmADMB::glmmadmb(SoilMaxC ~ RH + MaxWindSpeed +
                             FuelMoisture + tHa,
                           random = ~ 1|location/year/plot ,
-                          data = filter(imp_sc, SoilMaxC <= 3.0),
+                          data = filter(imp_sc, SoilMaxC <= 3.9),
                           family="gamma",
                           link="log",  
                           admb.opts=glmmADMB::admbControl(shess = TRUE, 

### 2021-02-12 r script/imputation.R

@@ -1,21 +1,22 @@
-pacman::p_load(tidyverse)
+pacman::p_load(tidyverse, mice)
 
-load(paste0(FilePath, "/data/AnalysisData.Rdata"))
+
+AnalysisData <- read_csv(paste0(getwd(), "/data/AnalysisDataKY.csv"))
 
 # MICE regression 
-
-pacman::p_load(mice, broom.mixed)
 
 #
 # Imputations
 #
   # View pattern of missing data
     AnalysisData %>%
+      select(AirTemp:tHa, -LAI) %>% 
         md.pattern() 
 
   # Calculate imputed datasets
     imp <- AnalysisData %>% 
-                mutate_at(vars(plot, array), as.character) %>%
+            select(-LAI) %>%
+                mutate(across(plot:array, ~as.character(.))) %>%
                  mice(m=500, seed = 23109, print=F)
     
     imp_raw <- 

### 2021-02-18 writeups/FuelsWeatherFirebehavior2.Rmd
robust imputation update

--- 2021-02-12 writeups/FuelsWeatherFirebehavior.Rmd
@@
 title: Fuels, fire weather, and Rx fire behavior in North Dakota
-subtitle: 
+subtitle: Robust imputation with `mice`
 date: '`r Sys.Date()`'
@@
 source("https://raw.githubusercontent.com/devanmcg/IntroRangeR/master/11_IntroMultivariate/ordinationsGGplot.R")
+# Load for markdown 
+pacman::p_load(tidyverse, mice, lubridate, gridExtra, vegan)
+
+response_CIs <- response_CIs %>%
+  filter(term != '(Intercept)') %>%
+  as_tibble() %>%
+  mutate(term = recode(term, tHa = 'Fuel load', 
+                             MaxWindSpeed = 'Wind speed', 
+                             RH = 'Relative humidity', 
+                             FuelMoisture = 'Fuel moisture')) 
 ```
 
-```{r}
-# Packages
-  pacman::p_load(tidyverse, mice, lubridate, gridExtra, vegan)
+```{r, eval = FALSE}
+# Packages & script for analysis
+  pacman::p_load(tidyverse, mice, broom.mixed, lubridate, gridExtra, vegan)
+  source('C:/Users/devan/GitHubProjects/SpatialFireBehavior/r script/mipo.R')
 ```
@@
 The raw weather, fuels, and fire behavior data used here are available in [CSV format](https://github.com/devanmcg/SpatialFireBehavior/raw/main/data/fromMZ/CompiledData2.csv).
+
+Figures generated by the script below and included in this document are [available on GitHub](https://github.com/devanmcg/SpatialFireBehavior/tree/main/writeups/figures), as well.
 
@@
         select(FireCode, timestamp, plot, array, TC, MaxC, 
-               AirTemp, RH, dpC, MaxWindSpeed, 
+               AirTemp, RH, dpC, WindSpeed, 
                LAI, FMC, KgHa)
@@
 AnalysisData %>%
-  select(-LAI) %>%
+  select(-LAI, -JD) %>%
   GGally::ggpairs(columns = 9:18, 
@@
 
-```{r dist_gg, fig.width = 8, fig.height =8, fig.cap = "Gamma distributions are fit to all response variables. Two outliers above 600 deg C were removed from analysis on soil maximum temperature."}
+```{r dist_gg, fig.width = 8, fig.height =8, fig.cap = "Gamma distributions are fit to all response variables. An outlier well above 600 deg C were removed from analysis on soil maximum temperature."}
 # Distributions of response variables
@@
   select(-FireCode, -location, -block, -pasture,-patch,
-         -year, -plot, -array) %>%
+         -year, -plot, -array, -LAI, -JD) %>%
     mice::md.pattern(plot = F) %>%
@@
 ```{r imp, eval=FALSE}
-# Calculate imputed datasets
-  imp <- AnalysisData %>% 
+# Calculate imputed datasets on scaled data
+    imp_sc <- AnalysisData %>% 
             select(-LAI) %>%
-              mutate(across(plot:array, ~as.character(.))) %>%
-               mice(m=500, seed = 23109, print=F)
-
-# Return completed datset   
+            mutate_at(vars(AirTemp:tHa), ~as.numeric(scale(., center=F))) %>%
+                      mutate(across(location:array, as.factor)) %>%
+                       mice(m=5, seed = 23109, print=F)
+# Reduce mids object to tibble for multivariate analysis    
     imp_raw <- 
-    complete(imp) %>%
+      complete(imp, 'long') %>%
       as_tibble() %>%
+      unite("TreeID", c(location, block, pasture,
+                        year, plot, array), sep = ".") %>%
+      select(-patch, -.id, -.imp, -FireCode) %>%
+      pivot_longer(names_to = "response", 
+                   values_to = "values", 
+                   -TreeID) %>%
+      group_by(TreeID, response) %>%
+      summarize(value = median(values)) %>%
+      ungroup() %>%
+      pivot_wider(names_from = response, 
+                  values_from = value) %>%
+      separate(TreeID, c("location", "block", "pasture",
+                         "year", "plot", "array")) %>% 
       mutate(across(location:array, as.factor))
-
-# Scale response variables for regression analysis   
-  imp_sc <- 
-    imp_raw %>%
-     mutate_at(vars(AirTemp:tHa), ~as.numeric(scale(., center=F)))
 ```
@@
 
-Mixed-effect regression models fit with `glmmADMB::glmmadmb` and constructed as follows:
+Mixed-effect regression models fit with `lme4::glmer` and constructed as follows:
 
@@
 
-```{r ADMB_mods, eval=FALSE}
-# Max soil surface temperature 
+```{r glmer_mods, eval=FALSE}
+# Mixed-effect regression models 
+#
+# Rate of spread
+#   
   # Fit model 
-    soil_ADMB <- glmmADMB::glmmadmb(SoilMaxC ~ RH + MaxWindSpeed +
-                            FuelMoisture + tHa,
-                          random = ~ 1|location/year/plot ,
-                          data=filter(imp_sc, SoilMaxC <= 3.9),
-                          family="gamma",
-                          link="log",  
-                          admb.opts=glmmADMB::admbControl(shess = TRUE, 
-                                                          noinit = TRUE, 
-                                                          impSamp = 10) )
-  # Retrieve regression results
+    ros_imp <- with(imp_sc, suppressMessages(
+                lme4::glmer(ros ~ RH + WindSpeed + 
+                              FuelMoisture + tHa + 
+                              (1|location/block/year/plot), 
+                            family=Gamma(link = "log"), 
+                            control=lme4::glmerControl(optimizer="bobyqa", 
+                                                       optCtrl=list(maxfun=100000)) )) )
+    ros_imp_pooled <- pool(ros_imp)$pooled  %>%
+                          as_tibble()
+    # Get terms
+    ros_terms <- 
+      full_join(
+      summary(pool(ros_imp)) %>% 
+        as_tibble() %>%
+        rownames_to_column("row"),
+
+      confint.mipo(pool(ros_imp)) %>%
+        as_tibble() %>%
+        rownames_to_column("row") )
+
+#
+# Maximum canopy temperature  
+#
+# Fit model 
+canopy_imp <- with(imp_sc, suppressMessages(
+              lme4::glmer(MaxC ~ RH + WindSpeed +  
+                            FuelMoisture + tHa + 
+                            (1|location/block/year/plot), 
+                          family=Gamma(link = "log"), 
+                          control=lme4::glmerControl(optimizer="bobyqa", 
+                                                     optCtrl=list(maxfun=100000)) )) )
+
+  canopy_imp_pooled <- pool(canopy_imp)$pooled  %>%
+                      as_tibble()
+
+  canopy_terms <- 
+    full_join(
+      summary(pool(canopy_imp)) %>% 
+        as_tibble() %>%
+        rownames_to_column("row"),
+
+      confint.mipo(pool(canopy_imp)) %>%
+        as_tibble() %>%
+        rownames_to_column("row") )
+
+#
+# Maximum soil surface temperature  
+#
+# Fit model 
+  soil_imp <- with(imp_sc, suppressMessages(
+                  lme4::glmer(log(SoilMaxC+1) ~ RH + WindSpeed +  
+                      FuelMoisture + tHa + 
+                      (1|location/block/year/plot), 
+                    family=Gamma(link = "log"), 
+                    control=lme4::glmerControl(optimizer="bobyqa", 
+                                               optCtrl=list(maxfun=100000)) )) )
+
+soil_imp_pooled <- pool(soil_imp)$pooled  %>%
+  as_tibble()
+
 soil_terms <- 
-      confint(soil_ADMB) %>%
-      as.data.frame %>%
-      rownames_to_column("term") %>%
-      rename(lwr = `2.5 %`, upr = `97.5 %`) %>%
   full_join(
-        coef(soil_ADMB) %>%
-          as.data.frame() %>% 
-          rownames_to_column("term") %>%
-          rename(estimate = '.') )
+    summary(pool(soil_imp)) %>% 
+      as_tibble() %>%
+      rownames_to_column("row"),
 
-# Maximum aboveground temp (plant canopy)  
-  # Fit model
-    canopy_ADMB <- glmmADMB::glmmadmb(MaxC ~ RH + MaxWindSpeed +
-                                        FuelMoisture + tHa,
-                          random = ~ 1|location/year/plot ,
-                          data=imp_sc,
-                          family="gamma",
-                          link="log",  
-                          admb.opts=glmmADMB::admbControl(shess = TRUE, 
-                                                          noinit = TRUE, 
-                                                          impSamp = 10))
-  # Retrieve regression results
-  canopy_terms <-   
-    confint(canopy_ADMB) %>%
-    as.data.frame %>%
-    rownames_to_column("term") %>%
-    rename(lwr = `2.5 %`, upr = `97.5 %`) %>%
-    full_join(
-      coef(canopy_ADMB) %>%
-        as.data.frame() %>% 
-        rownames_to_column("term") %>%
-        rename(estimate = '.') )
-
-# Rate of spread 
-  # Fit model 
-    ros_ADMB <- glmmADMB::glmmadmb(ros ~ RH + MaxWindSpeed +
-                                    FuelMoisture + tHa,
-                          random = ~ 1|location/year/plot ,
-                          data=imp_sc,
-                          family="gamma",
-                          link="log",  
-                          admb.opts=glmmADMB::admbControl(shess = TRUE, 
-                                                          noinit = TRUE, 
-                                                          impSamp = 10))
-  # Retrieve regression results  
-    ros_terms <-   
-      confint(ros_ADMB) %>%
-      as.data.frame %>%
-      rownames_to_column("term") %>%
-      rename(lwr = `2.5 %`, upr = `97.5 %`) %>%
-      full_join(
-        coef(ros_ADMB) %>%
-          as.data.frame() %>% 
-          rownames_to_column("term") %>%
-          rename(estimate = '.') )
-
-# Compile results
+    confint.mipo(pool(soil_imp)) %>%
+      as_tibble() %>%
+      rownames_to_column("row") )
 
@@
   ) %>%
-      as_tibble()
+  as_tibble() %>%
+  rename(lwr = `2.5 %`, upr = `97.5 %`)
 ```
@@
 
-**Rate of spread** significantly increased with windspeed ($z = 3.24, P < 0.01$) and decreased with greater fuel moisture content ($z = -5.84, P < 0.001$). 
-**Aboveground temperatures** in the plant canopy were lower with greater fuel moisture content ($z = -6.45, P < 0.001$) and increased with fuel load ($z = 5.49, P < 0.001$) and wind speed ($z = 2.82, P < 0.01$).
-**Temperature at the soil surface** had a statistically-significant inverse relationship with relative humidity ($z = -2.52, P = 0.01$) and fuel moisture ($z = -3.22, P < 0.01$), which likely indicates higher-intensity fires that burn down through litter.
+### Rate of spread
+
+```{r }
+response_CIs %>%
+  filter(response == "Rate of spread") %>%
+  rename(t = statistic, 
+         P = p.value) %>%
+    select(term, t, P) %>%
+  mutate(across(t:P, ~round(., 2))) %>%
+  pander::pander("Summary statistics for Rate of spread.") 
+```
+
+### Aboveground temperatures
+
+```{r }
+response_CIs %>%
+  filter(response == "Temp. above surface") %>%
+  rename(t = statistic, 
+         P = p.value) %>%
+    select(term, t, P) %>%
+  mutate(across(t:P, ~round(., 2))) %>%
+  pander::pander("Summary statistics for flame temperature 15 cm above soil surface.") 
+```
+
+### Soil surface temperature
+
+```{r }
+response_CIs %>%
+  filter(response == "Temp. at surface") %>%
+  rename(t = statistic, 
+         P = p.value) %>%
+    select(term, t, P) %>%
+  mutate(across(t:P, ~round(., 2))) %>%
+  pander::pander("Summary statistics for soil surface temperature.") 
+```
 
@@
 response_CIs %>%
-  filter(term != '(Intercept)') %>%
-  as_tibble() %>%
-  mutate(term = recode(term, tHa = 'Fuel load', 
-                             MaxWindSpeed = 'Wind speed', 
-                             RH = 'Relative humidity', 
-                             FuelMoisture = 'Fuel moisture'))  %>%
   ggplot(aes(x = reorder(term, abs(estimate), max))) + theme_bw(16) + 

### 2021-02-18 r script/DataLoading.R

@@ -17,7 +17,7 @@ FilePath = getwd()
     unite(timestamp, c(date, time), sep = " ") %>%
     mutate(timestamp = as.POSIXct(timestamp, format = "%Y-%m-%d %H:%M:%S")) %>%
     select(FireCode, timestamp, plot, array, TC, MaxC, 
-           AirTemp, RH, dpC, MaxWindSpeed, 
+           AirTemp, RH, dpC, WindSpeed, 
            LAI, FMC, KgHa) 
 
 # Isolate soil surface temperature (TC 4)
@@ -93,3 +93,5 @@ FilePath = getwd()
     select(-KgHa)
 
 # save(AnalysisData, file = paste0(FilePath, "/data/AnalysisData.Rdata"))
+  
+ # write_csv(AnalysisData, file = paste0(FilePath, "/data/AnalysisData.csv"))

### 2021-02-18 r script/GLMER_imp_regression.R

@@ -1,0 +1,116 @@
+pacman::p_load(tidyverse, broom.mixed)
+
+# Load imputed dataset 
+  load(paste0(getwd(), '/r objects/imp_sc.Rdata'))
+  source('C:/Users/devan/GitHubProjects/SpatialFireBehavior/r script/mipo.R')
+
+  
+# Mixed-effect regression models 
+#
+# Rate of spread
+#   
+  # Fit model 
+    ros_imp <- with(imp_sc, suppressMessages(
+                lme4::glmer(ros ~ RH + WindSpeed + 
+                              FuelMoisture + tHa + 
+                              (1|location/block/year/plot), 
+                            family=Gamma(link = "log"), 
+                            control=lme4::glmerControl(optimizer="bobyqa", 
+                                                       optCtrl=list(maxfun=100000)) )) )
+    ros_imp_pooled <- pool(ros_imp)$pooled  %>%
+                          as_tibble()
+    # Get terms
+    ros_terms <- 
+      full_join(
+      summary(pool(ros_imp)) %>% 
+        as_tibble() %>%
+        rownames_to_column("row"), 
+  
+      confint.mipo(pool(ros_imp)) %>%
+        as_tibble() %>%
+        rownames_to_column("row") ) 
+    
+
+#
+# Maximum canopy temperature  
+#
+# Fit model 
+  canopy_imp <- with(imp_sc, suppressMessages(
+                lme4::glmer(MaxC ~ RH + WindSpeed +  
+                              FuelMoisture + tHa + 
+                              (1|location/block/year/plot), 
+                            family=Gamma(link = "log"), 
+                            control=lme4::glmerControl(optimizer="bobyqa", 
+                                                       optCtrl=list(maxfun=100000)) )) )
+  
+  canopy_imp_pooled <- pool(canopy_imp)$pooled  %>%
+                            as_tibble()
+  
+  canopy_terms <- 
+    full_join(
+      summary(pool(canopy_imp)) %>% 
+        as_tibble() %>%
+        rownames_to_column("row"), 
+      
+      confint.mipo(pool(canopy_imp)) %>%
+        as_tibble() %>%
+        rownames_to_column("row") ) 
+
+#
+# Maximum soil surface temperature  
+#
+# Fit model 
+  soil_imp <- with(imp_sc, suppressMessages(
[…]

### 2021-02-18 writeups/FuelsWeatherFirebehavior.Rmd
robust imputation update

@@
   select(-FireCode, -location, -block, -pasture,-patch,
-         -year, -plot, -array) %>%
+         -year, -plot, -array, -LAI) %>%
     mice::md.pattern(plot = F) %>%

### 2021-02-18 r script/imputation.R

@@ -14,21 +14,30 @@ AnalysisData <- read_csv(paste0(getwd(), "/data/AnalysisDataKY.csv"))
         md.pattern() 
 
   # Calculate imputed datasets
-    imp <- AnalysisData %>% 
-            select(-LAI) %>%
-                mutate(across(plot:array, ~as.character(.))) %>%
-                 mice(m=500, seed = 23109, print=F)
+    imp_sc <- AnalysisData %>% 
+            select(-LAI, -JD) %>%
+            mutate_at(vars(AirTemp:tHa), ~as.numeric(scale(., center=F))) %>%
+                      mutate(across(location:array, as.factor)) %>%
+                       mice(m=5, seed = 23109, print=F)
     
     imp_raw <- 
-      complete(imp) %>%
-              as_tibble() %>% 
-                mutate(across(location:array, as.factor)) 
+      complete(imp, 'long') %>%
+      as_tibble() %>%
+      unite("TreeID", c(location, block, pasture,
+                        year, plot, array), sep = ".") %>%
+      select(-patch, -.id, -.imp, -FireCode) %>%
+      pivot_longer(names_to = "response", 
+                   values_to = "values", 
+                   -TreeID) %>%
+      group_by(TreeID, response) %>%
+      summarize(value = median(values)) %>%
+      ungroup() %>%
+      pivot_wider(names_from = response, 
+                  values_from = value) %>%
+      separate(TreeID, c("location", "block", "pasture",
+                         "year", "plot", "array")) %>% 
+      mutate(across(location:array, as.factor))  
     
-    imp_sc <- 
-      imp_raw %>%
-       mutate_at(vars(AirTemp:tHa), ~as.numeric(scale(., center=F)))
-  
+
     # save(imp_raw, file = paste0(FilePath, '/r objects/imp_raw.Rdata'))
     # save(imp_sc, file = paste0(FilePath, '/r objects/imp_sc.Rdata'))
-
-    

### 2021-03-01 writeups/FuelsWeatherFirebehaviorFigures.Rmd
begin final draft of figures

---
title: Fuels, fire weather, and Rx fire behavior in North Dakota
subtitle: Finalizing figures and results
date: '`r Sys.Date()`'
author: DAM
output:
  html_document:
    keep_md: false
    toc: true
    toc_float: true
    theme: united
    highlight: tango
---

```{r setup, echo=FALSE, warning=FALSE, message = FALSE, results='hide'}
knitr::opts_chunk$set(message = FALSE, warning=FALSE, 
               echo=FALSE, eval=TRUE, fig.path='figures/', dev=c('png', 'pdf')) 
```

```{r data_loading}
# Packages 
  pacman::p_load(tidyverse, readr, mice, broom.mixed, vegan,
                 lubridate, gridExtra, wesanderson)
set.seed(42)
# Data
  fp = url('https://raw.githubusercontent.com/devanmcg/SpatialFireBehavior/main/data/AnalysisDataKY.csv') 
  AnalysisData <- read_csv(fp) 
# Extra scripts
  source("https://raw.githubusercontent.com/devanmcg/IntroRangeR/master/11_IntroMultivariate/ordinationsGGplot.R")
  source('https://raw.githubusercontent.com/cran/mice/master/R/mipo.R')
```

# Methods

## Regression analysis

Figures generated by the script below and included in this document are [available on GitHub](https://github.com/devanmcg/SpatialFireBehavior/tree/main/writeups/figures), as well.

```{r data_prep, eval=FALSE}
###
### N O T   R U N
###
 AllData <-  
    read_csv(paste0(FilePath, "/data/fromMZ/CompiledData2.csv")) %>%
    filter(location != "OAK") %>%
      mutate(date = as.Date(date, format = "%m/%d/%Y"),
               L = str_remove(location, "REC"), 
               B = str_sub(block, 1,3), 
               Ps = str_replace(pasture, "[.]", ""), 
               Ps = str_sub(Ps, 1,2), 
               patch = str_replace(patch, "[.]", ""),
               y = format(date, "%y")) %>%
        unite("FireCode", c(L,B,Ps,patch,y), sep=".") %>%
    mutate(time = str_remove(MaxTempTime, "[.]+[0-9]"))%>%
    unite(timestamp, c(date, time), sep = " ") %>%
    mutate(timestamp = as.POSIXct(timestamp, format = "%Y-%m-%d %H:%M:%S")) %>%
        select(FireCode, timestamp, plot, array, TC, MaxC, 
               AirTemp, RH, dpC, WindSpeed, 
               LAI, FMC, KgHa) 
# Isolate soil surface temperature (TC 4)
  SoilTemp <-
    filter(AllData, TC == 4) %>%
      select(FireCode, plot, array, MaxC) %>%
        rename(SoilC = MaxC)
# Summarize array-level data 
  DataMeans <- 
    AllData %>%
      filter(TC %in% c('1', '2', '3')) %>% 
      select(-timestamp) %>%
      pivot_longer(cols = c(MaxC:KgHa), 
                   names_to = "var",
                   values_to = "value") %>%
      group_by(FireCode, plot, array, var) %>%
        summarize(Mean = mean(value) ) %>%
      ungroup() %>%
      pivot_wider(names_from = var, 
                  values_from = Mean) 
# Calculate Vapor Pressure Deficit
  DataMeans <- 
    DataMeans %>%
      mutate(e  = 6.11*(10^((7.5*dpC)/(237.3+dpC))), 
             es = 6.11*(10^((7.5*AirTemp)/(237.3+AirTemp))), 
             VPD = es - e) %>%
      select(-e, -es)
# Calculate ROS 
  D = 1   # Distance between thermocouples (m)
  ROS <- 
    AllData %>%
      filter(TC %in% c('1', '2', '3')) %>%
      mutate(timestamp = format(timestamp, "%H:%M:%S"), 
             ArrivalTime = seconds(hms(timestamp)) ) %>%
    select(FireCode, plot, array, ArrivalTime) %>%
    group_by(FireCode, plot, array) %>%
    arrange(ArrivalTime, .by_group = TRUE) %>% 
    mutate(position = order(order(ArrivalTime, decreasing=FALSE)), 
           position = recode(position, "1"="a", "2"="b", "3"="c"), 
           ArrivalTime = as.numeric(ArrivalTime) /60 ) %>%
    spread(position, ArrivalTime)  %>%
    ungroup %>% 
    mutate( theta_rad = atan((2*c - b - a) / (sqrt(3)*(b - a))), 
            ros = case_when(
              a == b ~ (sqrt(3) / 2) / (c - a) , 
              a != b ~  (D*cos(theta_rad) / (b - a) ) 
            )) %>%
    select(-a, -b, -c, -theta_rad)
# Combine, filter into final tibble for analysis 
  AnalysisData  <- 
    full_join(DataMeans, ROS) %>%
              left_join(SoilTemp) %>%
      filter( ros <= 40, 
              MaxC >= 40) %>%
      rename(FuelMoisture = FMC, 
             SoilMaxC = SoilC) %>%
        mutate(FuelMoisture = ifelse(FuelMoisture >= 0, 
                                      FuelMoisture, NA), 
               FuelMoisture = FuelMoisture * 100) %>%
      separate(FireCode, into = c("location", "block", "pasture", 
                           "patch", "year"), 
               remove = F)
###
###
###
```

For analysis, data are filtered to $\textsf{Rate of spread < 40 m/s}$ and $\textsf{Maximum temperature > 40}^\circ \textsf{C}$.

### Imputation of missing values

We used the multiple imputation method in the **mice** package to fill in missing values. 
The method creates 5 datasets, each with different but reasonable values for the missing data based on patterns in existing data, and performs the regression analysis on each. 
These results are pooled into a single set of results.

```{r imp, eval=TRUE}
# Calculate imputed datasets on scaled data
  imp_sc <- AnalysisData %>% 
          select(-LAI, -JD) %>%
          mutate_at(vars(AirTemp:tHa), ~as.numeric(scale(., center=F))) %>%
                    mutate(across(location:array, as.factor)) %>%
                     mice(m=5, seed = 23109, print=F)
# Reduce mids object to tibble for multivariate analysis    
    imp_raw <- 
      complete(imp_sc, 'long') %>%
      as_tibble() %>%
      unite("TreeID", c(location, block, pasture,
                        year, plot, array), sep = ".") %>%
      select(-patch, -.id, -.imp, -FireCode) %>%
      pivot_longer(names_to = "response", 
                   values_to = "values", 
                   -TreeID) %>%
      group_by(TreeID, response) %>%
      summarize(value = median(values)) %>%
      ungroup() %>%
      pivot_wider(names_from = response, 
                  values_from = value) %>%
      separate(TreeID, c("location", "block", "pasture",
                         "year", "plot", "array")) %>% 
      mutate(across(location:array, as.factor))
```

### Model specification

Mixed-effect regression models fit with `lme4::glmer` and constructed as follows:

$$y \approx RH + WS + FM + FL + (1|location/year/plot),~ 
\text{family = gamma, link = log}$$

where

$RH$ = Relative humidity,

$WS$ = Average wind speed,

$FM$ = Fuel moisture (from clipped samples),

$FL$ = Fuel load (t/ha from clipped samples)

Weather variables $RH$ and $WS$ were taken from NDAWN stations for the hour in which the fire behavior observation occurred.

All variables were scaled to a common range prior to analysis, and units for each predictor variable are not reported.

Response variables included

* Maximum temperature (C) 15 cm above soil surface (mean of three thermocouples)
* Maximum temperature (C) at soil surface (single thermocouple)
* Rate of spread (m/s) through three points of a 1 m equilateral triangle

Thermocouple data recorded with the open-source [FeatherFlame](https://diyfirescience.info/portfolio/work2/) system.

### Multivariate analysis

Principal Components Analysis (PCA) on fire behavior responses fit with `vegan::rda`. 
Post-hoc group (location) and gradient (fire weather) analysis with `vegan::envfit`.

An additional composite variable, vapor pressure deficit (VPD), was calculated as

$\text{VPD} = e - e_s$

where

$e = 6.11 \cdot 10^{\frac{7.5 \cdot \text{Dewpoint}}{237.3 + \text{Dewpoint}}}$

and

$e_s = 6.11 \cdot 10^{\frac{7.5 \cdot \text{Air temperature}}{237.3 + \text{Air temperature}}}$

# Results

## Fuel, weather, and fire behavior across locations

```{r data_summary_gg, fig.height=10, fig.width=6}
LongData <- 
  AnalysisData %>%
    select(-LAI, -JD, -MaxWindSpeed) %>%
    pivot_longer(cols = AirTemp:tHa, 
                 names_to = 'var', 
                 values_to = 'value') %>%
    mutate(type = case_when(
                    var %in% c('MaxC', 
                               'SoilMaxC', 
                               'ros') ~ 'Fire behavior', 
                    var %in% c('tHa',
                               'FuelMoisture') ~ 'Fuel', 
                    TRUE ~ 'Weather' ), 
           var = recode(var, 
                        AirTemp = 'Air temperature (°C)', 
                        RH = 'Relative humidity (%)', 
                        dpC = 'Dew point (°C)', 
                        FuelMoisture = 'Fuel moisture (%)', 
                        tHa = 'Fuel load (t/ha)',
                        MaxC = 'Flame temp (°C)', 
                        WindSpeed = 'Wind speed (m/s)', 
                        ros = 'Rate of spread (m/s)', 
                        SoilMaxC = 'Soil surface temp (°C)'), 
           location = recode(location, 
                             H = "Hettinger", 
                             CG = "Central Grasslands"))

wx_d <-
LongData %>%
  filter(type == "Weather") %>%
  ggplot(aes(x = location, y= value,
             fill = location)) + theme_bw(14) +
    geom_violin() +
  labs(y = NULL, x = NULL, 
       title = "Weather") + 
  facet_wrap( ~ var, scales = 'free') +
  scale_fill_manual(name = "Location", 
                    values = wes_palette("Cavalcanti1")[4:5]) + 
  theme(axis.text.x = element_blank(), 
        panel.grid.major.x = element_blank(), 
        legend.position = "none")

fuel_d <-
LongData %>%
  filter(type == "Fuel") %>%
  ggplot(aes(x = location, y= value,
             fill = location)) + theme_bw(14) +
    geom_violin() +
  labs(y = NULL, x = NULL, 
       title = "Fuel") + 
  facet_wrap( ~ var, scales = 'free') +
  scale_fill_manual(name = "Location", 
                    values = wes_palette("Cavalcanti1")[4:5]) + 
  theme(axis.text.x = element_blank(), 
        panel.grid.major.x = element_blank(), 
        legend.position = "none")

fb_d <-
  LongData %>%
    filter(type == "Fire behavior", 
           value <= 600) %>%
    ggplot(aes(x = location, y= value,
               fill = location)) + theme_bw(14) +
      geom_violin() +
    labs(y = NULL, x = NULL, 
         title = "Fire behavior") + 
    facet_wrap( ~ var, scales = 'free') +
    scale_fill_manual(name = "Location", 
                      values = wes_palette("Cavalcanti1")[4:5]) + 
    theme(axis.text.x = element_blank(), 
          panel.grid.major.x = element_blank(), 
          legend.position = "bottom")

# Create list of panels
  gl <- lst(wx_d, fuel_d, fb_d)

  grid.arrange(
    grobs = gl,
    widths = c(1),
    heights = c(0.35, 0.25, 0.35)  )
```

## Variables associated with fire behavior

```{r glmer_mods, eval=TRUE}
# Mixed-effect regression models 
#
# Rate of spread
#   
  # Fit model 
    ros_imp <- with(imp_sc, suppressMessages(
                lme4::glmer(ros ~ RH + WindSpeed + 
                              FuelMoisture + tHa + 
                              (1|location/block/year/plot), 
                            family=Gamma(link = "log"), 
                            control=lme4::glmerControl(optimizer="bobyqa", 
                                                       optCtrl=list(maxfun=100000)) )) )
    ros_imp_pooled <- pool(ros_imp)$pooled  %>%
                          as_tibble()
    # Get terms
    ros_terms <- 
      full_join(
      summary(pool(ros_imp)) %>% 
        as_tibble() %>%
        rownames_to_column("row"),

      confint.mipo(pool(ros_imp)) %>%
        as_tibble() %>%
        rownames_to_column("row") )

#
# Maximum canopy temperature  
#
# Fit model 
canopy_imp <- with(imp_sc, suppressMessages(
              lme4::glmer(MaxC ~ RH + WindSpeed +  
                            FuelMoisture + tHa + 
                            (1|location/block/year/plot), 
                          family=Gamma(link = "log"), 
                          control=lme4::glmerControl(optimizer="bobyqa", 
                                                     optCtrl=list(maxfun=100000)) )) )

  canopy_imp_pooled <- pool(canopy_imp)$pooled  %>%
                      as_tibble()

  canopy_terms <- 
    full_join(
      summary(pool(canopy_imp)) %>% 
        as_tibble() %>%
        rownames_to_column("row"),

      confint.mipo(pool(canopy_imp)) %>%
        as_tibble() %>%
        rownames_to_column("row") )

#
# Maximum soil surface temperature  
#
# Fit model 
  soil_imp <- with(imp_sc, suppressMessages(
                  lme4::glmer(log(SoilMaxC+1) ~ RH + WindSpeed +  
                      FuelMoisture + tHa + 
                      (1|location/block/year/plot), 
                    family=Gamma(link = "log"), 
                    control=lme4::glmerControl(optimizer="bobyqa", 
                                               optCtrl=list(maxfun=100000)) )) )

soil_imp_pooled <- pool(soil_imp)$pooled  %>%
  as_tibble()

soil_terms <- 
  full_join(
    summary(pool(soil_imp)) %>% 
      as_tibble() %>%
      rownames_to_column("row"),

    confint.mipo(pool(soil_imp)) %>%
      as_tibble() %>%
      rownames_to_column("row") )

response_CIs <-   
  bind_rows(
    mutate(soil_terms, response = 'Temp. at surface'), 
    mutate(canopy_terms, response = 'Temp. above surface'),
    mutate(ros_terms, response = "Rate of spread") 
  ) %>%
  as_tibble() %>%
  rename(lwr = `2.5 %`, upr = `97.5 %`) %>%
  filter(term != '(Intercept)') %>%
  mutate(term = recode(term, tHa = 'Fuel load', 
                             WindSpeed = 'Wind speed', 
                             RH = 'Relative humidity', 
                             FuelMoisture = 'Fuel moisture')) 
```

### Rate of spread

```{r }
response_CIs %>%
  filter(response == "Rate of spread") %>%
  rename(t = statistic, 
         P = p.value) %>%
    select(term, t, P) %>%
  mutate(across(t:P, ~round(., 2))) %>%
  pander::pander("Summary statistics for Rate of spread.") 
```

### Aboveground temperatures

```{r }
response_CIs %>%
  filter(response == "Temp. above surface") %>%
  rename(t = statistic, 
         P = p.value) %>%
    select(term, t, P) %>%
  mutate(across(t:P, ~round(., 2))) %>%
  pander::pander("Summary statistics for flame temperature 15 cm above soil surface.") 
```

### Soil surface temperature

```{r }
response_CIs %>%
  filter(response == "Temp. at surface") %>%
  rename(t = statistic, 
         P = p.value) %>%
    select(term, t, P) %>%
  mutate(across(t:P, ~round(., 2))) %>%
  pander::pander("Summary statistics for soil surface temperature.") 
```

```{r CI_gg, fig.width=8, fig.height = 3.75}
# Plot confidence intervals
response_CIs %>%
  mutate(response =case_when(
                            response == "Temp. above surface" ~ "°C, canopy", 
                            response == "Temp. at surface" ~ "°C, soil", 
                            TRUE ~ response              )
                          ) %>%
  ggplot(aes(x = reorder(term, abs(estimate), max))) + theme_bw(16) + 
    geom_hline(yintercept = 0, color="black", linetype = 2) +
    geom_errorbar(aes(ymin = lwr, ymax = upr), 
                  size = 1.1, width = 0.2, 
                  color = "blue") +
    geom_point(aes(y = estimate), pch = 21, 
               stroke = 1.5, size = 4,
               color = "blue", 
               fill="lightblue") +
    labs(x = '', 
         y = 'Regression coefficient with 95% CI') + 
    coord_flip(ylim=c(-2,2.4)) + 
    facet_wrap(~ response) + 
    theme(axis.text.y = element_text(color="black", size = 14), 
          strip.text = element_text(size = 14), 
          panel.grid.major.y = element_blank(), 
          panel.grid.minor.x = element_blank())
```

## Multivariate analysis

```{r fit_pca}
# Fire behavior PCA 
  fb_d <- 
    imp_raw %>%
    select(MaxC, ros, SoilMaxC)

  fb_pca <- rda(fb_d ~ 1, 'euc', scale = T)
```

```{r pca_eig}
 summary(fb_pca)$cont$importance %>% pander::pander("Eigenvalues and proportion explained by composite axes (principal components).")
```

```{r}
# Testing for differences between locations 
pca_fc <-  envfit( fb_pca ~ location, imp_raw, 
                   choices = c(1:3), 
                   strata = imp_raw$year, 
                   199)$factors
pca_fc
```

No difference in fire behavior patterns between Hettinger and Central Grasslands; this is expected.

```{r envfit, eval=TRUE}
# Testing fire weather against PCA
pca_ev <- envfit(fb_pca ~ MaxWindSpeed+AirTemp+dpC+RH+VPD, 
                 data = imp_raw, 
                 choices = c(1:3), 
                 strata = imp_raw$location)
pca_ev
```

dpC, RH, and VPD had statistically-significant associations with variation in fire behavior, but even these relationships were weak.

```{r pca_dat}  
  # get ord data for ggplotting
  pca_gg <- gg_ordiplot(fb_pca, groups = imp_raw$location, 
                        plot=FALSE)

  pca_spp <- 
    scores(fb_pca, display = "species") %>%
    as.data.frame %>%
    as_tibble(rownames="response") %>%
    mutate(response = recode(response, 
                             "MaxC" = "canopy °C", 
                             "SoilMaxC" = 'soil °C', 
                             "ros" = "ROS"))

  # Vector 
  pca_vd <- scores(pca_ev,  "vectors") %>%
                as.data.frame %>% 
                rownames_to_column("gradient") %>%
              filter(gradient %in% c("dpC", "RH", "VPD")) %>%
              select(-PC3) %>%
            as_tibble( ) %>%
              mutate(gradient = ifelse(gradient == "dpC", 
                                       "DP", gradient))

  pca_scores <- lst(species=pca_spp, 
                    vectors=pca_vd)
  pca_scores$spiders <-
    pca_gg$df_spiders %>% 
    rename(PC1 = x, PC2 = y) %>%
    as_tibble()
```

```{r pca_gg, fig.height=4, fig.width=6, fig.cap='Principal Components Analysis of fire behavior data (response variables in red) for prescribed burns on rangeland at Hettinger (H) and Central Grasslands  (CG) Research Extension Centers. No difference between locations and only the vapor pressure deficit (VPD) and relative humidty (RH) gradients had significant associations with variation in fire behavior.' }
# ggplot PCA
  ggplot() + theme_bw(16) + 
    labs(x="PC 1", 
         y="PC 2") + 
    geom_vline(xintercept = 0, lty=3, color="darkgrey") +
    geom_hline(yintercept = 0, lty=3, color="darkgrey") +
    geom_point(data=pca_scores$spiders, 
                        aes(x=PC1, y=PC2, 
                            shape=Group, 
                            colour=Group), 
                        size=2)  +   
    geom_segment(data=pca_scores$spiders, 
                 aes(x=cntr.x, y=cntr.y,
                     xend=PC1, yend=PC2, 
                     color=Group), 
                 size=1.2, 
                 alpha = 0.75) +
    geom_point(data=pca_scores$spiders, 
               aes(x=PC1, y=PC2, fill=Group), 
               colour="grey40", 
               pch=21, 
               size=3, 
               stroke=2, alpha = 0.75) +
    geom_segment(data=pca_scores$vectors, 
                 aes(x=0, y=0, 
                     xend=PC1*1.6, 
                     yend=PC2*1.6),
                 arrow=arrow(length = unit(0.03, "npc")),
                 lwd=1.5) +
  geom_label(data=pca_scores$spiders,
               aes(x=cntr.x,
                   y=cntr.y,
                   label=Group,
                   color=Group),
               fontface="bold", size=4,
               label.size = 0,
               label.r = unit(0.5, "lines")) +
    geom_text(data=pca_scores$vectors,
              aes(x=PC1*1.75, 
                  y=PC2*1.75, 
                  label=gradient), 
              nudge_x = 0.06, 
              nudge_y =-0.05, 
              size=6, fontface="bold") + 
    geom_label(data=pca_scores$species, 
                aes(x=PC1/3, 
                    y=PC2/3, 
                    label=response), 
                label.padding=unit(0.25,"lines"),
                label.size = 0.5, 
                fontface="bold", 
               color="darkblue") +
    scale_color_manual(values = wes_palette("Cavalcanti1")[4:5]) + 
    scale_fill_manual(values = wes_palette("Cavalcanti1")[4:5])  +
    theme(panel.grid=element_blank(), 
          legend.position="none")

```

It is interesting that the gradient in temperature at the soil surface (soil °C) is orthogonal to both canopy temperature and rate of spread (canopy °C and ROS, respectively). 
This confirms trends observed in the regression analysis that the conditions most conducive to fast, high-intensity fires are not the same that result in surface heating.

Managers might be able to use high windspeeds to mitigate soil heating when fuels are dry, or focus on conditions with low RH to ensure litter burns down to the soil surface.

# Script

```{r script, ref.label=knitr::all_labels(!label %in% c('setup', 'dist_gg', 'data_summary_gg', 'CI_gg', 'pca_eig', 'pca_dat', 'pca_gg') ), echo=TRUE, eval=FALSE}
```

### 2021-08-06 paper/FigureGenerator.Rmd
finish first draft

---
output:
  word_document:
    reference_docx: word_template.docx
    keep_md: yes
---

```{r setup, echo=FALSE, warning=FALSE, message = FALSE, results='hide'}
knitr::opts_chunk$set(message = FALSE, warning=FALSE, 
                      echo=FALSE, eval=TRUE)

# Packages 
pacman::p_load(tidyverse, gridExtra, wesanderson, captioner)

wes_cb1 = c('#68d2ea', '#800000' )

  t_nums <- captioner::captioner(prefix = "Table")
  f_nums <- captioner::captioner(prefix = "Figure")
```

```{r data_loading}
# Data
load("C:/Users/devan.mcgranahan/GithubProjects/SpatialFireBehavior/r objects/response_CIs.Rdata")
load("C:/Users/devan.mcgranahan/GithubProjects/SpatialFireBehavior/r objects/pca_scores.Rdata")

# Extra scripts
  source("https://raw.githubusercontent.com/devanmcg/IntroRangeR/master/11_IntroMultivariate/ordinationsGGplot.R")
```

<!-- Table 1: Results of generalized linear mixed effect regression models testing three measure of fire behavior against four potential predictor variables. Statistics reflect pooled results of 50 imputed datasets using the *mice* package in R; see Methods. Vapor pressure deficit included for Rate of spread only due to statistically-significant difference between GLMM regression results that included VPD compared to RH alone (Wald = 5.32, P = 0.02), while temperature models had no such difference. -->

<!-- |Response           |Model term             |       t df  |      P| -->
<!-- |:------------------|:----------------------|------------:|------:| -->
<!-- |Rate of spread     |                       |             |       | -->
<!-- |                   |Wind speed             |  2.92 108.0 | < 0.01| -->
<!-- |                   |Vapor pressure deficit | -2.31 108.1 |   0.02| -->
<!-- |                   |Relative humidity      | -1.66 80.8  |   0.10| -->
<!-- |                   |Fuel load              |  1.16 49.9  |   0.25| -->
<!-- |                   |Fuel moisture          | -0.68 62.9  |   0.50| -->
<!-- |Canopy temperature |                       |             |       | -->
<!-- |                   |Fuel load              |  2.82 54.2  |   0.01| -->
<!-- |                   |Fuel moisture          | -2.16 40.9  |   0.04| -->
<!-- |                   |Relative humidity      | -1.19 120.4 |   0.24| -->
<!-- |                   |Wind speed             |  0.02 132.4 |   0.99| -->
<!-- |Surface temperature|                       |             |       | -->
<!-- |                   |Relative humidity      | -1.19 74.8  |   0.24| -->
<!-- |                   |Fuel load              | -0.48 20.1  |   0.64| -->
<!-- |                   |Fuel moisture          | -0.47 40.1  |   0.64| -->
<!-- |                   |Wind speed             |  0.06 49.5  |   0.95| -->

```{r }
violin_cap <- f_nums(name = "violin_plots", 
                   caption = "Distribution of weather, fuel, and fire behavior data for fires in southwestern North Dakota (Hettinger, dark maroon) and central North Dakota (Central Grasslands, light blue) sampled from 2017 to 2019. Summary statistics include median (horizontal gray lines) and means (triangles).  ")
```

`r f_nums('violin_plots')`
```{r data_summary_gg, fig.cap = violin_cap, fig.height=7.5, fig.width=6}
LongData <- 
    read_csv(url('https://raw.githubusercontent.com/devanmcg/SpatialFireBehavior/main/data/AnalysisDataKY.csv')) %>% 
    select(-LAI, -JD, -MaxWindSpeed) %>%
    mutate(ros = ifelse(ros >= 10, NA, ros), 
           tHa = ifelse(tHa >=4, NA, tHa)) %>%
    pivot_longer(cols = AirTemp:tHa, 
                 names_to = 'var', 
                 values_to = 'value') %>%
    mutate(type = case_when(
                    var %in% c('MaxC', 
                               'SoilMaxC', 
                               'ros') ~ 'Fire behavior', 
                    var %in% c('tHa',
                               'FuelMoisture') ~ 'Fuel', 
                    TRUE ~ 'Weather' ), 
           var = recode(var, 
                        AirTemp = 'Air temperature (°C)', 
                        RH = 'Relative humidity (%)', 
                        dpC = 'Dew point (°C)', 
                        FuelMoisture = 'Fuel moisture (%)', 
                        tHa = 'Fuel load (t/ha)',
                        MaxC = 'Canopy temp (°C)', 
                        WindSpeed = 'Wind speed (m/s)', 
                        ros = 'Rate of spread (m/s)', 
                        SoilMaxC = 'Soil surface temp (°C)'), 
           location = recode(location, 
                             H = "Hettinger", 
                             CG = "Central Grasslands"))

ld_sums <-
  LongData %>%
    group_by(location, type, var) %>%
    summarize(mean = mean(value, na.rm = T), 
              .groups = 'drop')

wx_d <-
LongData %>%
  filter(type == "Weather") %>%
  ggplot(aes(x = location, y= value,
             fill = location)) + theme_bw(14) +
    geom_violin(draw_quantiles = 0.5, 
                color = "darkgrey", 
                size = 1) +
  geom_point(data = filter(ld_sums, type == "Weather"), 
             aes(x = location, 
                 y = mean), 
                 fill = 'black', 
             color = 'white', 
             pch = 24, 
             size = 3, 
             stroke = 1.1) + 
  labs(y = NULL, x = NULL, 
       title = "Weather") + 
  facet_wrap( ~ var, scales = 'free') +
  scale_fill_manual(name = "Location", 
                    values = wes_cb1) + 
  theme(axis.text.x = element_blank(), 
        panel.grid.major.x = element_blank(), 
        legend.position = "none")

fuel_d <-
LongData %>%
  filter(type == "Fuel") %>%
  ggplot(aes(x = location, y= value,
             fill = location)) + theme_bw(14) +
        geom_violin(draw_quantiles = 0.5, 
                color = "darkgrey", 
                size = 1) +
  geom_point(data = filter(ld_sums, type == "Fuel"), 
             aes(x = location, 
                 y = mean), 
                 fill = 'black', 
             color = 'white', 
             pch = 24, 
             size = 3, 
             stroke = 1.1) + 
  labs(y = NULL, x = NULL, 
       title = "Fuel") + 
  facet_wrap( ~ var, scales = 'free') +
  scale_fill_manual(name = "Location", 
                    values = wes_cb1) + 
  theme(axis.text.x = element_blank(), 
        panel.grid.major.x = element_blank(), 
        legend.position = "none")

fb_d <-
  LongData %>%
    filter(type == "Fire behavior", 
           value <= 600) %>%
    ggplot(aes(x = location, y= value,
               fill = location)) + theme_bw(14) +
          geom_violin(draw_quantiles = 0.5, 
                color = "darkgrey", 
                size = 1) +
  geom_point(data = filter(ld_sums, type == "Fire behavior"), 
             aes(x = location, 
                 y = mean), 
                 fill = 'black', 
             color = 'white', 
             pch = 24, 
             size = 3, 
             stroke = 1.1) + 
    labs(y = NULL, x = NULL, 
         title = "Fire behavior") + 
    facet_wrap( ~ var, scales = 'free') +
    scale_fill_manual(name = "Location", 
                      values = wes_cb1) + 
    theme(axis.text.x = element_blank(), 
          panel.grid.major.x = element_blank(), 
          legend.position = "bottom")

# Create list of panels
  gl <- lst(wx_d, fuel_d, fb_d)

  grid.arrange(
    grobs = gl,
    widths = c(1),
    heights = c(0.35, 0.25, 0.35)  )
```

```{r }
pca_cap <- f_nums(name = "pca", 
                   caption = "Principal Components Analysis of fire behavior data (response variables in blue; rate of spread (ROS), temperature above surface (canopy ºC), and temperature at soil surface (soil ºC) for prescribed burns on rangeland at Hettinger (H), in southwestern North Dakota, and Central Grasslands (CG), in central North Dakota. No difference between locations (P = 0.11). Total variance explained in these two axes = 86%. ")
```

`r f_nums('pca')`
```{r pca_gg, fig.height=4, fig.width=6.5, fig.cap=pca_cap }
# ggplot PCA
  ggplot() + theme_bw(16) + 
    labs(x="PC 1", 
         y="PC 2") + 
    geom_vline(xintercept = 0, lty=3, color="darkgrey") +
    geom_hline(yintercept = 0, lty=3, color="darkgrey") +
    geom_point(data=pca_scores$spiders, 
                        aes(x=PC1, y=PC2, 
                            shape=Group, 
                            colour=Group), 
                        size=2)  +   
    geom_segment(data=pca_scores$spiders, 
                 aes(x=cntr.x, y=cntr.y,
                     xend=PC1, yend=PC2, 
                     color=Group), 
                 size=1.2, 
                 alpha = 0.75) +
    geom_point(data=pca_scores$spiders, 
               aes(x=PC1, y=PC2, fill=Group), 
               colour="grey40", 
               pch=21, 
               size=3, 
               stroke=2, alpha = 0.75) +
    geom_segment(data=pca_scores$vectors, 
                 aes(x=0, y=0, 
                     xend=PC1*1.6, 
                     yend=PC2*1.6),
                 arrow=arrow(length = unit(0.03, "npc")),
                 lwd=1.5) +
  geom_label(data=pca_scores$spiders,
               aes(x=cntr.x,
                   y=cntr.y,
                   label=Group,
                   color=Group),
               fontface="bold", size=4,
               label.size = 0,
               label.r = unit(0.5, "lines")) +
    geom_text(data=pca_scores$vectors,
              aes(x=PC1*1.75, 
                  y=PC2*1.75, 
                  label=gradient), 
              nudge_x = 0.06, 
              nudge_y =-0.05, 
              size=6, fontface="bold") + 
    geom_label(data=pca_scores$species, 
                aes(x=PC1*c(0.28, 0.3, 0.3), 
                    y=PC2*c(0.001, 0.3, 0.3), 
                    label=response), 
                label.padding=unit(0.25,"lines"),
               nudge_y =c(-0.05, 0, 0.05),
                label.size = 0.5, 
                fontface="bold", 
               color="darkblue") +
    scale_color_manual(values = wes_cb1) + 
    scale_fill_manual(values = wes_cb1)  +
    theme(panel.grid=element_blank(), 
          legend.position="none")

```

```{r }
CIs_cap <- f_nums(name = "CI_plots", 
                   caption = "Regression coefficients and 95% confidence intervals for fuel and weather terms from models for maximum temperature at 15 cm above the soil surface (canopy), maximum temperature at the soil surface (soil) and rate of spread. ")
```

`r f_nums('CI_plots')`
```{r CI_gg, fig.cap = CIs_cap, fig.width=8, fig.height = 5}
# Plot confidence intervals
response_CIs %>%
  mutate(response =case_when(
                            response == "Temp. above surface" ~ "°C, canopy", 
                            response == "Temp. at surface" ~ "°C, soil", 
                            TRUE ~ response ), 
         term = str_replace_all(term, " ", "\n")  ) %>%
  ggplot(aes(x = reorder(term, abs(estimate), max))) + theme_bw(16) + 
    geom_hline(yintercept = 0, color="black", linetype = 2) +
    geom_errorbar(aes(ymin = lwr, ymax = upr), 
                  size = 1.3, width = 0.2, 
                  color = wes_palette("GrandBudapest1")[4]) +
    geom_point(aes(y = estimate), pch = 21, 
               stroke = 1.5, size = 4,
               color = wes_palette("GrandBudapest1")[4], 
               fill=wes_palette("GrandBudapest1")[1]) +
    labs(x = '', 
         y = 'Regression coefficient with 95% CI') + 
    coord_flip(ylim=c(-2,2)) + 
    facet_wrap(~ response) + 
    theme(axis.text.y = element_text(color="black", size = 14, hjust = 0.5), 
          strip.text = element_text(size = 14), 
          panel.grid.major.y = element_blank(), 
          panel.grid.minor.x = element_blank())
```

### 2021-08-06 paper/GrasslandFireBehaviorNGP.Rmd
finish first draft

---
title: "Weather and fuel as modulators of grassland fire behavior in the northern Great Plains"
author: "Devan Allen McGranahan^A,C^, Megan E. Zopfi^B^, Kathryn A. Yurkonis^B^ ^A^ USDA Agricultural Research Service, Livestock and Range Research Laboratory, Miles City, Montana, USA ^B^ Department of Biology, University of North Dakota, Grand Forks, North Dakota, USA ^C^ Corresponding author. Email: Devan.McGranahan@usda.gov"
output:
  word_document:
    reference_docx: word_template.docx
    keep_md: yes
csl: ijwf.csl
bibliography: FireBehaviorBib.bib
---

```{r setup, include=FALSE, eval = TRUE}
knitr::opts_chunk$set(echo = FALSE, eval = FALSE)
```

# Abstract

Fuel and fire weather conditions interact to affect wildland fire behavior, but little is known about how these variables affect fire behavior in the northern Great Plains of North America.
Data that are available consist mostly of reports based on temperature-time data from thermocouples, and neither consider rate of spread or statistically test the influence of fire environmental variables. 
We measured fuel load and fuel moisture ahead of prescribed fires in southwestern and central North Dakota, USA, and used a unique multi-channel thermocouple array to measure rate of spread, soil surface temperature, and flame temperature in the plant canopy, which we compared with fire weather data taken from nearby weather stations. 
Canopy temperatures averaged 225^o^C during spring burns in central North Dakota and 250^o^C during fall burns in southwestern North Dakota. 
Surface temperatures averaged just above 100^o^C, although 50% of observations were 60^o^C or less. 
Regression analysis indicated that wind speed drove faster rates of spread while higher fuel loads and lower fuel moisture produced higher canopy temperatures. 
None of our measurements explained variability in soil surface temperature, likely because ground-level heating remained low. 
We highlight the differential responses among fire behavior metrics to different components of the wildland fire environment. 
These results can help wildland fire managers better match burn conditions to desired outcomes and fire ecologists match measurements to specific ecological responses. 
We also discuss knowledge gaps that remain in understanding how land management and ignition patterns interact to effect variability in fire behavior, and what fire scientists need to learn about the relationships between atmospheric moisture and fuel curing in grassland systems.

**Key words:** Grassland fire ecology and management; Prescribed fire; Rangeland fire management; Robust wildland fire science; Wildland fire science in working landscapes

# Introduction

More than simply the result of combustion of vegetation, fire behavior in wildland environments is multi-faceted, with different components producing different effects on the surrounding environment and organisms within. 
Most wildland fire scientists describe fire behavior in terms of *rate of spread*—how quickly a flame front moves through a fuelbed—and *intensity*—a suite of measurements of how much energy is released by combustion, often expressed as a rate of energy release over time [@mcgranahan2021a].

Wildland fire behavior is controlled by interactions between several abiotic and biotic factors, and understanding them is critical to safe and effective wildland fire management [@benson2009]. 
*Abiotic factors* include those determined by the physical environment, such as wind speed and atmospheric moisture content. 
Wind speed has long been recognized as a primary driver of fire behavior, especially in well-cured grassland fuels [@whittaker1961; @cheney1995; @kidnie2015].
Two measures of atmospheric moisture content—relative humidity and vapor pressure deficit—are associated with fire growth [@evett2008; @reid2010; @sedano2014].

*Biotic factors* relate principally to the vegetation biomass available for combustion. 
Overall, fire spread and energy release rates increase as more fuel is available to burn; the structure and arrangement of plant biomass is also important. 
Greater fuel load attributable to longer time-since-fire increased fire temperature; spatial variability in fuel load and patchy distribution of fine fuels contributed to spatial variability in fire behavior [@patten1984; @gibson1990; @gomes2020].
Furthermore, fine-leaved grasses burn more completely and hotter than an equal mass of forbs [@wragg2018].
Finally, fuel moisture content is an especially important driver outside of the highly-cured context of wildfire seasons [@sparling1966; @kidnie2015]. 
Together, variability in flammability traits and curing rates among species that comprise grassland fuelbeds contributes to variability in fire behavior [@mcgranahan2016; @cardoso2018].

How and where within the wildland fire environment fire behavior measurements are made matters a great deal to assessing fire effects. 
For decades, fire ecologists have measured fire behavior as *flame temperature* via various methods, including arrays of temperature-sensitive paints [e.g., @whittaker1961; @smith1966; @bailey1980] or by recording air temperature as a flame front passes over a thermocouple connected to a datalogger [e.g., @strong2013; @russell2015]. 
Despite its popularity among fire ecologists, temperature alone is a poor response variable fraught by several issues in collecting and interpreting thermocouple data [see review by @mcgranahan2020]. 
Firstly, a considerable amount of variability in temperature attributable to sensor placement relative to both the ground and the fire. 
There is neither a standard for placing thermocouple probes in the wildland fire environment nor consistency in vertical temperature profiles. 
Most observations of surface fire temperature profiles describe an inverse, linear relationship between height and temperature [@smith1966; @patten1984; @archibold2003], although @ramsay1996 found the highest temperatures at the top of a 1 m profile and the lowest temperatures at 30 cm, while @frost1987 and @bailey1980 present evidence that the highest temperatures occur midway up the profile.
At least some of this variability might be due to differences in surface vs. canopy temperature among head and back fires [@trollope1978]. 
Secondly, many factors that contribute to variability in temperatures recorded by thermocouples are attributable to the nature of the sensor rather than the nature of the fire. 
Thus, reporting 'device temperatures' alone impedes comparisons between studies; @bova2008 present a standard calibration of thermocouple probes, while @mcgranahan2021b simply uses the timestamps of peak heating across an array of several thermocouples to calculate rate of spread.
Using rate of spread as the response variable makes moot the third issue with temperature as a fire behavior metric: temperature of the media around a probe is a poor proxy for the thermal experience of an organism.
Measures of intensity or energy flux are more biologically relevant [@kremens2012; @smith2016].

In the North American Great Plains, most reports of grassland fire behavior consist of temperatures derived from thermocouples, and there are few data on rate of spread. 
Soil surface temperatures in South Dakota tallgrass prairie ranged from 200-500^o^C during spring burns, and were greatest under lower fuel loads [@ohrtman2015]; fires in Saskatchewan mixed grass prairie exceeded 300^o^C 5-10cm above the soil surface in spring, summer, and fall [@archibold2003]. 
Mean temperatures in experimental burns in eastern Montana ranged from 172-222^o^C in the summer to 253^o^C in the spring [@strong2013; @russell2015]. 
To our knowledge, no studies on grassland fire in the northern Great Plains region has explicitly tested the effect of fire weather on fire behavior.

Our objectives were to (1) describe the range of variability in three measures of fire behavior—rate of spread, canopy temperature, and soil surface temperature—during prescribed burns in typical fuelbeds of the northern US Great Plains, and (2) explain variability in fire behavior in terms of abiotic and biotic conditions. 
Our analysis emphasizes the differential effects of environmental variables among the three responses, and the multidimensional relationship between these different measures of grassland fire behavior.

# Methods

## Study locations

We sampled 25 prescribed fires at two locations in central and southwestern North Dakota, USA. 
At both locations, sampled grasslands are included in a patch-burn grazing study that requires a portion of each experimental unit to be burned each year [@spiess2020]. 
Fire sizes ranged from 8-16 ha. 
Typical ignition patterns consisted of downwind backing fires followed by either ring ignition and primarily head fire spread, when fuels were conducive; when fuels were sparse or higher-moisture, flanking fires and strip ignitions were employed as necessary to ensure fire spread through the entire burn unit.

In central North Dakota, we sampled 15 spring (May) fires at the North Dakota State University Central Grassland Research Extension Center near Streeter, ND (46.718686 N, 99.448521 W). 
Burned grasslands at this location are divided into two 260 ha blocks with four, 65 ha pastures each in which either an 8- or 16-ha patch is burned each spring.
Located in a mixed-grass prairie ecoregion, this location has a rolling topography and receives an average of 468 mm annual precipitation. 
Vegetation is mixed-grass prairie invaded by exotic, C3 grasses; stands are dominated by *Pascopyrum smithii*, *Nassella viridula*, *Poa pratensis*, *Bromus inermis*, *Koeleria macrantha*, *Artemisia* spp., *Solidago* spp., and clumps of *Glycyrrhiza lepidota* and *Symphoricarpos occidentalis*.

In southwestern North Dakota, we sampled ten, 16-ha fall (October) fires in two blocks at the North Dakota State University Hettinger Research Extension Center, Hettinger, ND (46.004443 N, 100.646477 W) with mean annual precipitation of 380 mm. 
Topography is consistently flat.
Located in a shortgrass prairie ecoregion, these pastures are dominated by introduced C3 grasses *Thinopyrum intermedium*, *Bromus inermis*, *Agropyron cristatum*, and *Poa pratensis*, along with the non-native legume *Medicago sativa*.

## Data collection

We measured each fire with 27 a set of 9, 1 m equilateral triangle plots arranged in a nested fashion such that three 10 m triangles, each containing 3, 1m plots, were placed 100 m apart to form a total plot area of 0.433 ha with 27 sample points positioned at the centroid of each burn unit. 
This fractal design is modified from the Sierpinksi triangle described by @dorrough2007 and applied to measuring wildland fire spread by @mcgranahan2021b.
Although the logarithmically-scaled nested design was intended for geospatial analysis of point-level data, for our analyses here we calculate averages from the finest (1 m) scale.

**Fuel data** were collected no more than three hours prior to fire ignition. 
We clipped and collected all fuels in a 25 × 25 cm quadrat positioned 0.5 m away from each 1 m triangle vertex; the three measurements per plot were averaged prior to analysis. 
Fuel samples were initially placed in airtight plastic bags to retain moisture, and then weighed, dried to constant mass at 60^o^ for 48 hours, and reweighed. 
These data were used to calculate percent fuel moisture content (expressed on a dry-weight basis) and fuel load (kg/ha) for each plot (n = 9 subsamples for each fire).

**Fire behavior data** were recorded as temperature (^o^) associated with the advancing flame front at each of the 27 points arranged in 9, 1 m triangles at the center of each burn unit.
Data were recorded with the open-source FeatherFlame thermocouple datalogger system [@mcgranahan2021b].
Briefly, the FeatherFlame system reads overbraided, ceramic fiber-insulated K-type thermocouples (Omega, Norwalk, CT) connected to an Arduino-based datalogger assembled from Adafruit Feather breakout boards (M0 Adalogger, datalogging shield, and OLED display; Adafruit Industries, LLC, New York City, NY) and housed inside water-resistant Pelican (Pelican Products, Inc, Torrance, California) cases. 
The low cost of open-source systems make multiple units more affordable than proprietary data loggers with no sacrifice in data quality [@mcgranahan2021].

At each 1-m triangle—the individual observational unit in the nested plot design—we used four thermocouples to a single FeatherFlame datalogger.
Three thermocouples measured flame temperature 15 cm above the soil surface at each 1 m vertex while a fourth thermocouple recorded soil surface temperature at a representative point within the 1 m array.  
The soil thermocouple was placed on mineral ground, perpendicular to the soil surface, below plant litter. 
Dataloggers recorded thermocouple temperatures at 1.5 Hz.

For each fire event, we determined the time and value of the maximum temperature (^o^) as the flame front encountered each thermocouple. 
We calculated the Maximum flame temperature (^o^) 15 cm above soil surface for each plot as the mean of the thermocouple readings from its vertices. 
We calculated the rate of spread (m/s) of the flame front as it passed through each plot using the maximum temperature timestamps following equations 1 and 2 from @simard1984.

**Fire weather data** were obtained after each fire from records made available by the North Dakota Agriculture Weather Network, the statewide mesonet system with sensor arrays at both experimental stations. 
We downloaded hourly relative humidity (%), dew point (^o^), air temperature (^o^), and average wind speed (km/h).
From these data we calculated atmospheric vapor pressure (e) and saturation vapor pressure (es) and used these quantities to determine the vapor pressure deficit (VPD = e - es) for the hour in which each fire behavior observation occurred. 
These data capture hourly trends in weather at the meso-gamma scale [2-20 km; @orlanski1975], and are reliably connected to our fire behavior measurements via time stamps provided by the dataloggers. 
The Hettinger mesonet array is 3–8 km from burned pastures, while the Central Grasslands mesonet array is 1–7.5 km  from burned pastures. 
We found a high degree of consistency between these meso-scale data and fire weather records made on the fireline during operational periods, and the open rangeland physiognomy with flat to rolling terrain precludes substantial microsite differences in weather between these records and the conditions at each fire behavior sample point.

## Data Analysis

Prior to analysis, we ensured statistical power across 167 observational units by using multiple imputation to interpolate missing datapoints, as missing field data occurred for three rate of spread samples (2% of total), 29 fuel load values (17%), and 46 soil surface temperature values (27%). 
We used the multiple imputation method in the *mice* package 
[@vanbuuren2011] in the R statistical environment [@rct2020] to fill in these missing values. 
The procedure simulated 100 datasets with different, but reasonable, values for the missing data based on patterns in the existing data. 
We then scaled all variables to a common range within each imputed dataset, performed regression analysis on each *mice*-generated dataset, and report composite statistical results pooled from the results of the 100 individual regression models.

**Multivariate analysis**
We first conducted a multivariate analysis to assess composite relationships among the fuel, weather, and fire behavior responses. 
We used Principal Components Analysis (PCA) fit with the 'rda' function in the *vegan* package for R [@oksanen2017].
We performed post-hoc group (location) and gradient (fire weather) analysis with the vegan 'envfit' function, stratified by year.

**Regression analysis** 
We assessed weather and fuel effects on three fire behavior response variables: Maximum flame temperature (^o^) 15 cm above the soil surface (mean of three thermocouples), Maximum soil surface temperature (^o^; single thermocouple), and Rate of spread (m/s) through each 1 m equilateral triangle plot. 
Because all three response variables were best modeled with a gamma distribution, we fit generalized linear mixed-effect regression models for each response with the 'glmer' function from the *lme4* package in R [@bates2015]. 
Fixed effects consisted of weather and fuel variables, as described above. 
The random-effect term was constructed to account for spatial non-independence within locations and nested variance within sample plots, and the effect of repeated measurements within each location. 
Due to concerns about colinearity between relative humidity and vapor pressure deficit because they are derived from the same variables, vapor pressure deficit was excluded from regression analysis for all three response variables.

# Results

Most measures of fuel, fire weather, and fire behavior showed considerable variability within each location, although rates of spread were generally low (Fig. 1). 
Principal Components Analysis indicated fire behavior patterns were consistent across locations (P = 0.11), although the fall season in Hettinger—our semi-arid location in southwestern North Dakota—tended to have drier air and hotter fires (Fig. 2).
Spring fires in the Central Grasslands were conducted under warmer and more evaporative (VPD) conditions than fall fires at Hettinger. 
The first two axes of the Principal Components Analysis (Fig. 2) explained 86% of overall variance in the fire behavior dataset. 
The first axis (52% variance explained) was most strongly associated with canopy temperature and rate of spread, while the second axis was more strongly associated with soil surface temperature. 
Dew point was marginally related (p < 0.05) and inversely associated with canopy temperatures and rate of spread dew point had marginally significant association with variation in fire behavior.

Fires at both locations were characterized by considerable variability among sub-plots. 
Within the plant canopy, half of the fires (13) exceeded 325^o^C, but only four of those fires had >50% of individual sample plots within the burns reach an average of 325^o^C. 
Sparse fuels meant that fire did not spread to some individual plots in some burns despite strip ignitions.
Among plots that burned, less than half exceeded 100^o^C at the soil surface (49 of 121 plots).
A majority of the 21 fires had at least one plot exceed 100ºC at the soil surface (18 of 21 fires), seven had over half of the plots exceed 100^o^C at the soil surface and only one fire had all plots reach this (a spring Central Grasslands fire). 
Only 10 fires reached 325^o^C on the soil surface, with most of these fires reaching this point in only one plot and never in more than half of the plots.

The effects of fuel and fire weather predictor variables varied across the three response variables (Fig. 3). 
Fires spread faster with higher wind speeds (t = 2.92, P < 0.01), but no other variable had a statistically-significant association with rate of spread (Table 1). 
Temperatures in the plant canopy (15 cm above the soil surface) increased as fuel load increased (t = 2.82, P = 0.01) and decreased as fuel moisture increased (t = -2.16, P = 0.04). 
No fuel or weather variable included here had statistically-significant associations with soil surface temperature (Table 1).

# Discussion

In our comparison of three measurements of fire behavior—rate of spread and maximum temperature recorded on the soil surface and 15 cm above the soil surface—against fuel and fire weather variables, we found considerable variability in which predictor variables were associated with different measures of fire behavior. 
These data directly support the safe and effective application of prescribed fire in the region. 
Some results are straightforward and consistent with decades of fire safety science; e.g., faster rates of fire spread associated with higher wind speed and lower relative humidity. 
Other results add nuance to an ecological understanding of how fire behavior relates to fire effects—e.g., factors like fuel load and fuel moisture were related with canopy temperature but not soil surface temperature, which suggests that direct effects on belowground plant tissue and soil biota are not correlated with aboveground heating and fire spread.

To our knowledge, this is the first study from the northern Great Plains to scrutinize the factors that influence fire behavior, and the first to combine reports of fire spread and temperature data from thermocouples. 
Most published research on fire spread in the Great Plains is derived from computer simulations [@mcgranahan2013; @overholt2014; @yurkonis2019]. 
The few field studies from the region mostly report temperature data from thermocouples and rarely incorporate fuel and fire weather data into the analysis; when such information is provided, it is typically included in the study description, not as data. 
Given the high degree of variability in the wildland fire environment, a mechanistic understanding of grassland fire dynamics will require collect fuel, fire weather, and fire behavior data in a spatially and temporally consistent manner to facilitate statistical analyses of their relationships [@hiers2020; @mcgranahan2018].

Mean temperatures recorded in this study are consistent with other reports from northern rangelands. 
In central Alberta grassland, @bailey1980 observed that surface temperatures varied between 110^o^C and 165^o^C for backfires and headfires, respectively, and headfires averaged 200^o^C 15 cm above the ground; temperatures generally tracked with fuel load. 
Surface fires through jack pine barrens in Ontario had a similar range as ours: 140-545^o^C [@smith1966]. 
In our study, average 15-cm temperatures were 225^o^C during spring burns in central North Dakota and 250^o^C during fall burns in southwestern North Dakota; surface temperatures at both locations generally averaged just above 100^o^C (Fig. 1).

Discrepencies between our data and others reported in the region are consistent with what would be expected when differences in the fire environment are considered.
For example, @ohrtman2015 reported a wide range of maximum temperatures at the soil surface---150-500^o^C---that was generally explained by variability in annual productivity and clipping frequency. 
Our fires were also cooler than those reported by @archibold2003 in Saskatchewan: using the mid-point of observations made at 10 cm and 20 cm as a comparison, spring fires reached 314^o^C and fall fires reached 298^o^C. 
But @archibold2003 also reported substantially lower fuel moisture in each season and they had approximately three times the fuel load, likely due to an absence of grazing management on the remnant prairie. 
With greater variability in fuel load and fuel moisture, we might also expect to see these factors have greater influence on aboveground flame temperatures. 
For example, in Colorado, @augustine2014 observed a strong linear relationship between fuel load and  temperatures ranging from 60-200^o^C, but their fuel load also ranged from 0.2 to 1.2 t/ha.

Although one must look beyond the northern Great Plains for comparable data on fire spread, our results are consistent with international reports. 
For example, our data match the pattern from Australian grassland, where @cheney1993 found a statistical relationship between rate of spread and both wind speed and fuel moisture, but not total fuel load. 
Their fires traveled at rates similar to the vast majority of ours (1-2 m/s), with slightly higher  fuel loads and slightly lower relative humidity. 
Wind was by far their most important variable, followed by dead fuel moisture, consistent with patterns in our data. 
Likewise, @cruz2015 observed fire spread rates up to 2 m/s in fully-cured grassland fuels with fuel loads at the upper end of our samples.  
In comparing our results to these studies, it appears the trigonometric method of calculating rate of spread through the thermocouple array [@simard1984] overestimates rate of spread, but there is no indication that using this measure of spread as a scaled response variable in regression models presents error in interpreting the relative effect of fuel and weather predictors.

Interestingly, @cheney1993 found that fires spread faster in undisturbed pastures compared to those that had been cut, which they attribute to differences in fuel structure (height, bulk density) rather than fuel load. 
This might have implications for fire behavior in our region, where invasive species like *Poa pratensis* generally increase aboveground plant biomass but do so by adding thick dense litter at the soil surface, rather than standing dead fuel in the plant canopy [@gasch2020]. 
While difficult to tease apart statistically in the present data, many burn units in our mesic location in central North Dakota were dominated by *P. pratensis* and indeed, that location tended to have higher fuel loads and lower rates of spread (Fig. 1), consistent with simulations of fire spread through those *P. pratensis*-dominated prairies [@yurkonis2019].

Much is made of the difference in fire behavior between head and back fires in the fire ecology literature, and while we expect these differences translate to different fire effects in our system, making distinctions between fire types is difficult in both our data and our management. 
@trollope1978 emphasized that while head fires move faster and generally release more energy, back fires effect greater heating at the ground level. 
One would expect, then, that back fires would have more opportunity to burn down through even thick *P. pratensis* litter to mineral soil. 
Unfortunately, our results offer little insight into what fuel or weather variables enhance litter combustion, likely because most of our fires never got very hot at the soil surface---50% of our observations were less than 60^o^C, and 60% less than 100^0^C (Fig. 1). 
Nor can we differentiate the direction of fire spread with the current trigonometry applied to the triangular thermocouple arrays [@simard1984], although it would theoretically be possible to compare spread direction to wind direction if the latter data were available at a fine enough scale. 
But in these pastures, the functional difference between head and back fires might be moot. Because our fuels were often sparse, sometimes only marginally cured, and prescriptions precluded taking advantage of higher wind or lower relative humidity to mitigate fuel limitations, we often employed substantial interior ignitions using strip, point, flanking, and spiral patterns that sent flame fronts towards our sensors in all possible directions at different times. 
Differences in ignition patterns create additional spatial variability in fire behavior and burn severity [@williams2015]. 
Our data are certainly useful in describing the variability in fire behavior across these burns, but do not inform the relationship between ignition pattern and fire behavior. 
Thus, future research on fire behavior in the northern Great Plains should use experimental plots with consistent fuelbeds to explicitly compare head and back fires set via line ignitions, akin to the experimental burning program described by @cruz2015 in Australia.

A more detailed analysis of fuelbed effects on fire behavior also ought to separate fuels into live and dead components, and consider the moisture content of each along with fuel load ratios. 
We report here the fine fuel moisture content of the entire fuelbed, consistent with descriptive, post hoc statistical approaches to describing fire behavior [@trollope1985; @mcgranahan2016; @cruz2015]. 
But predictive fire behavior models accommodate inputs for live and dead fuel categories [@scott2005], and @kidnie2015a found that four categories of live, dead, and senescent fuels best represented differences in grassland fuel moisture scenarios. 
Although parsing live and dead fuel moisture in the present analysis would probably not better explain variability in our dataset, it would likely contribute to better predictions of fire behavior relative to management objectives if information on fuel moisture content were available prior to ignition. 
Unfortunately, the standard clipping and drying method is not compatible with providing day-of fuel moisture data, and visual assessments based on color tend to over-predict curedness [@kidnie2015a]. 
However, electronic devices can provide accurate and instantaneous measurements of grassland fuel moisture [@mcgranahan2019].

Research must also address the influence of atmospheric moisture conditions on prescribed fire behavior. 
Several broad-scale, post-hoc analyses of wildfire conditions conclude that atmospheric moisture  is an important driver of burned area [@evett2008; @reid2010; @sedano2014]. 
But experiments that explicitly test the immediate effect of relative humidity on fire behavior report no appreciable effect on surface fire temperatures or rate of spread [@sparling1966; @trollope1985]. 
It is likely that atmospheric moisture plays a larger role in modulating fuel moisture content prior to combustion than affecting instantaneous fire behavior itself---consider how fire behavior models take fuel moisture as a parameter and not relative humidity, but include relative humidity as an input to determine fuel moisture content [@rothermel1983; @cruz2016].

The most appropriate measure of atmospheric moisture content might also be unresolved. 
We focused our analysis here on relative humidity because it is so common in fire behavior models and fire weather forecasts. 
But vapor pressure deficit has also been identified as an important driver of fire spread and intensity [@gomes2020]. 
In fact, @srock2018 suggest vapor pressure deficit might be a better measure of atmospheric moisture content for fire predictions, but the Hot-Dry-Windy index they developed to incorporate vapor pressure deficit operates at synoptic scales beyond the spatial extent and operational periods of prescribed burns. 
Given that substantial changes in atmospheric moisture changes in recent decades are expected to strengthen over the 21^st^ century [@seager2015; @ficklin2017], understanding how these dynamics affect fire behavior will be an essential component of managing resilent fire regimes.

# Acknowledgements

We appreciate support from the North Dakota State Agricultural Experiment Station, including K. Sedivec at Central Grasslands REC and C. Schauer and B. Geaumont at Hettinger REC. 
We recognize funding from the University of North Dakota Department of Biology, USDA-NIFA Hatch project number ND02393, and USDA-NIFA AFRI award number 2018-67020-27856. 
We appreciate the technical assistance of several North Dakota State University graduate students and faculty for assistance with prescribed burning. 
L. LaFond, C. LaFond, and E. Wahl assisted with datalogger assembly. 
J. Spiess, J. Cutter, M. Lakey, B. Poling, and A. Steele assisted with sensor deployment and data collection.

# References

### 2021-08-06 paper/GrasslandFireBehaviorNGP.tex
finish first draft

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  11pt,
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
  \setmainfont[]{cochineal}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}
    }
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering

\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{Test}
\author{true \and true \and true}
\date{September 11, 2020}

% Jesus, okay, everything above this comment is default Pandoc LaTeX template. -----
% ----------------------------------------------------------------------------------
% I think I had assumed beamer and LaTex were somehow different templates.

\usepackage{kantlipsum}

\usepackage{abstract}
\renewcommand{\abstractname}{}    % clear the title
\renewcommand{\absnamepos}{empty} % originally center

\renewenvironment{abstract}
 {{%
    \setlength{\leftmargin}{0mm}
    \setlength{\rightmargin}{\leftmargin}%
  }%
  \relax}
 {\endlist}

\makeatletter
\def\@maketitle{%
  \newpage
%  \null
%  \vskip 2em%
%  \begin{center}%
  \let \footnote \thanks
      {\fontsize{18}{20}\selectfont\raggedright  \setlength{\parindent}{0pt} \@title \par}
    }
%\fi
\makeatother

\title{Test }

%\author{\Large Devan Allen McGranahan\vspace{0.05in} \newline\normalsize\emph{USDA-Agricultural Research Service, Livestock and Range Research
Station, Miles City, Montana, USA}   \and \Large Megan Zopfi\vspace{0.05in} \newline\normalsize\emph{Department of Biology, University of North Dakota, Grand Forks, North
Dakota, USA}   \and \Large Kathryn Yurkonis\vspace{0.05in} \newline\normalsize\emph{Department of Biology, University of North Dakota, Grand Forks, North
Dakota, USA}  }

\date{}

\usepackage{titlesec}

% 
\titleformat*{\section}{\large\bfseries}
\titleformat*{\subsection}{\normalsize\itshape} % \small\uppercase
\titleformat*{\subsubsection}{\normalsize\itshape}
\titleformat*{\paragraph}{\normalsize\itshape}
\titleformat*{\subparagraph}{\normalsize\itshape}

% add some other packages ----------

% \usepackage{multicol}
% This should regulate where figures float
% See: https://tex.stackexchange.com/questions/2275/keeping-tables-figures-close-to-where-they-are-mentioned
\usepackage[section]{placeins}

\makeatletter
\@ifpackageloaded{hyperref}{}{%
\ifxetex
  \PassOptionsToPackage{hyphens}{url}\usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \PassOptionsToPackage{hyphens}{url}\usepackage[draft,unicode=true]{hyperref}
\fi
}

\@ifpackageloaded{color}{
    \PassOptionsToPackage{usenames,dvipsnames}{color}
}{%
    \usepackage[usenames,dvipsnames]{color}
}
\makeatother
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={Devan Allen McGranahan (USDA-Agricultural Research Service, Livestock and Range Research
Station, Miles City, Montana, USA) and Megan Zopfi (Department of Biology, University of North Dakota, Grand Forks, North
Dakota, USA) and Kathryn Yurkonis (Department of Biology, University of North Dakota, Grand Forks, North
Dakota, USA)},
             pdfkeywords = {pandoc, r markdown, knitr},  
            pdftitle={Test},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

% Add an option for endnotes. -----

% This will better treat References as a section when using natbib
% https://tex.stackexchange.com/questions/49962/bibliography-title-fontsize-problem-with-bibtex-and-the-natbib-package

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\newtheorem{hypothesis}{Hypothesis}

\begin{document}

% \textsf{\textbf{This is sans-serif bold text.}}
% \textbf{\textsf{This is bold sans-serif text.}}

% \maketitle

{% \usefont{T1}{pnc}{m}{n}
\setlength{\parindent}{0pt}
\thispagestyle{plain}
{%\fontsize{18}{20}\selectfont\raggedright
\maketitle  % title \par

}

{
   \vskip 13.5pt\relax \normalsize\fontsize{11}{12} 
   \MakeUppercase{Devan Allen McGranahan}, \small{USDA-Agricultural Research Service, Livestock and Range Research
Station, Miles City, Montana, USA}   \par \vskip -3.5pt \MakeUppercase{Megan Zopfi}, \small{Department of Biology, University of North Dakota, Grand Forks, North
Dakota, USA}   \par \vskip -3.5pt \MakeUppercase{Kathryn Yurkonis}, \small{Department of Biology, University of North Dakota, Grand Forks, North
Dakota, USA}

}

}

\begin{abstract}

%    \hbox{\vrule height .2pt width 39.14pc}

    \vskip 8.5pt % \small

\noindent \small{Lorem ipsum dolor sit amet, consectetur adipiscing elit. Donec sit amet
libero justo. Pellentesque eget nibh ex. Aliquam tincidunt egestas
lectus id ullamcorper. Proin tellus orci, posuere sed cursus at,
bibendum ac odio. Nam consequat non ante eget aliquam. Nulla facilisis
tincidunt elit. Nunc hendrerit pellentesque quam, eu imperdiet ipsum
porttitor ut. Interdum et malesuada fames ac ante ipsum primis in
faucibus. Suspendisse potenti. Duis vitae nibh mauris. Duis nec sem sit
amet ante dictum mattis. Suspendisse diam velit, maximus eget commodo
at, faucibus et nisi. Ut a pellentesque eros, sit amet suscipit eros.
Nunc tincidunt quis risus suscipit vestibulum. Quisque eu fringilla
massa.}

\vskip 8.5pt \noindent \emph{Keywords}: pandoc, r markdown, knitr \par

%    \hbox{\vrule height .2pt width 39.14pc}

\end{abstract}

\vskip -8.5pt

 % removetitleabstract

\setlength{\parindent}{16pt}
\setlength{\parskip}{0pt}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Fire is awesome.

\end{document}

### 2021-08-06 paper/analysis.Rmd
finish first draft

--- 2021-03-01 writeups/FuelsWeatherFirebehaviorFigures.Rmd
@@
     keep_md: false
-    toc: true
+    toc: false
     toc_float: true
@@
 knitr::opts_chunk$set(message = FALSE, warning=FALSE, 
-               echo=FALSE, eval=TRUE, fig.path='figures/', dev=c('png', 'pdf')) 
+                      echo=FALSE, eval=TRUE) 
 ```
@@
   AnalysisData <- read_csv(fp)
+
+  wd = 'C:/Users/devan.mcgranahan/GithubProjects/SpatialFireBehavior/'
 # Extra scripts
@@
 We used the multiple imputation method in the **mice** package to fill in missing values. 
-The method creates 5 datasets, each with different but reasonable values for the missing data based on patterns in existing data, and performs the regression analysis on each. 
+The method creates multiple datasets, each with different but reasonable values for the missing data based on patterns in existing data, and performs the regression analysis on each. 
 These results are pooled into a single set of results.
@@
                     mutate(across(location:array, as.factor)) %>%
-                     mice(m=5, seed = 23109, print=F)
+                     mice(m=100, seed = 23109, print=F)
 # Reduce mids object to tibble for multivariate analysis    
@@
     ros_imp <- with(imp_sc, suppressMessages(
-                lme4::glmer(ros ~ RH + WindSpeed + 
-                              FuelMoisture + tHa + 
+                lme4::glmer(ros ~ VPD + tHa + RH +
+                              FuelMoisture + WindSpeed + 
                               (1|location/block/year/plot), 
@@
                             control=lme4::glmerControl(optimizer="bobyqa", 
-                                                       optCtrl=list(maxfun=100000)) )) )
+                                                       optCtrl=list(maxfun=500000)) )) )
     ros_imp_pooled <- pool(ros_imp)$pooled  %>%
@@
 canopy_imp <- with(imp_sc, suppressMessages(
-              lme4::glmer(MaxC ~ RH + WindSpeed +  
-                            FuelMoisture + tHa + 
+              lme4::glmer(MaxC ~ VPD + tHa +  RH +
+                              FuelMoisture + WindSpeed +
                             (1|location/block/year/plot), 
@@
   soil_imp <- with(imp_sc, suppressMessages(
-                  lme4::glmer(log(SoilMaxC+1) ~ RH + WindSpeed +  
-                      FuelMoisture + tHa + 
+                  lme4::glmer(log(SoilMaxC+1) ~ VPD + tHa +  RH +
+                              FuelMoisture + WindSpeed +
                       (1|location/block/year/plot), 
@@
                              RH = 'Relative humidity', 
-                             FuelMoisture = 'Fuel moisture')) 
+                             FuelMoisture = 'Fuel moisture', 
+                             VPD = 'Vapor Pressure Deficit')) 
+save(response_CIs, file = paste0(wd, 'r objects/response_CIs.Rdata'))
 ```
@@
     as_tibble()
+  save(pca_scores, file = paste0(wd, 'r objects/pca_scores.Rdata'))
 ```

### 2021-08-06 paper/analysis2.Rmd
finish first draft

--- 2021-08-06 paper/analysis.Rmd
@@
 
+```{r}
+ AnalysisData %>% 
+          select(-LAI, -JD) %>%
+  ggplot(aes(x=tHa)) + theme_bw(16) + 
+              geom_histogram(aes(y=..density..),      
+                 binwidth=0.1,
+                 colour="black",
+                 fill="lightgreen")
+
+AnalysisData %>%
+  select(SoilMaxC) %>%
+  quantile(c(0.25, 0.5, .59, 0.6, 0.75), na.rm = T)
+```
+
 ```{r imp, eval=TRUE}
@@
           select(-LAI, -JD) %>%
+          mutate(ros = ifelse(ros >= 12, NA, ros), 
+                 tHa = ifelse(tHa >= 4, NA, tHa)) %>%
           mutate_at(vars(AirTemp:tHa), ~as.numeric(scale(., center=F))) %>%
                     mutate(across(location:array, as.factor)) %>%
-                     mice(m=100, seed = 23109, print=F)
+                     mice(m=50, seed = 23109, print=F)
+
 # Reduce mids object to tibble for multivariate analysis    
@@
 
-### Model specification
-
-Mixed-effect regression models fit with `lme4::glmer` and constructed as follows:
-
-$$y \approx RH + WS + FM + FL + (1|location/year/plot),~ 
-\text{family = gamma, link = log}$$
-
-where
-
-$RH$ = Relative humidity,
-
-$WS$ = Average wind speed,
-
-$FM$ = Fuel moisture (from clipped samples),
-
-$FL$ = Fuel load (t/ha from clipped samples)
-
-Weather variables $RH$ and $WS$ were taken from NDAWN stations for the hour in which the fire behavior observation occurred.
-
-All variables were scaled to a common range prior to analysis, and units for each predictor variable are not reported.
-
-Response variables included
-
-* Maximum temperature (C) 15 cm above soil surface (mean of three thermocouples)
-* Maximum temperature (C) at soil surface (single thermocouple)
-* Rate of spread (m/s) through three points of a 1 m equilateral triangle
-
-Thermocouple data recorded with the open-source [FeatherFlame](https://diyfirescience.info/portfolio/work2/) system.
-
-### Multivariate analysis
-
-Principal Components Analysis (PCA) on fire behavior responses fit with `vegan::rda`. 
-Post-hoc group (location) and gradient (fire weather) analysis with `vegan::envfit`.
-
-An additional composite variable, vapor pressure deficit (VPD), was calculated as
-
-$\text{VPD} = e - e_s$
-
-where
-
-$e = 6.11 \cdot 10^{\frac{7.5 \cdot \text{Dewpoint}}{237.3 + \text{Dewpoint}}}$
-
-and
-
-$e_s = 6.11 \cdot 10^{\frac{7.5 \cdot \text{Air temperature}}{237.3 + \text{Air temperature}}}$
-
 # Results
@@
     select(-LAI, -JD, -MaxWindSpeed) %>%
+    mutate(ros = ifelse(ros >= 10, NA, ros)) %>%
     pivot_longer(cols = AirTemp:tHa, 
@@
   # Fit model 
-    ros_imp <- with(imp_sc, suppressMessages(
-                lme4::glmer(ros ~ VPD + tHa + RH +
+# ros_VPD <- with(imp_sc, suppressMessages(
+#               lme4::glmer(ros ~ VPD + RH + tHa +
+#                                 FuelMoisture + WindSpeed +
+#                                 (1|location/block/year/plot), 
+#                           family=Gamma(link = "log"), 
+#                           control=lme4::glmerControl(optimizer="bobyqa", 
+#                                                      optCtrl=list(maxfun=100000)) )) )
+
+ros_RH <- with(imp_sc, suppressMessages(
+              lme4::glmer(ros ~ RH + tHa +
                                 FuelMoisture + WindSpeed +
@@
                           control=lme4::glmerControl(optimizer="bobyqa", 
-                                                       optCtrl=list(maxfun=500000)) )) )
-    ros_imp_pooled <- pool(ros_imp)$pooled  %>%
-                          as_tibble()
+                                                     optCtrl=list(maxfun=100000)) )) )
+D1(ros_VPD, ros_RH) # Different
+
   # Get terms
@@
       full_join(
-      summary(pool(ros_imp)) %>% 
+      summary(pool(ros_RH)) %>% 
         as_tibble() %>%
@@
 
-      confint.mipo(pool(ros_imp)) %>%
+      confint.mipo(pool(ros_RH)) %>%
         as_tibble() %>%
@@
 # Fit model 
-canopy_imp <- with(imp_sc, suppressMessages(
-              lme4::glmer(MaxC ~ VPD + tHa +  RH +
+# canopy_VPD <- with(imp_sc, suppressMessages(
+#               lme4::glmer(MaxC ~ VPD + RH + tHa +
+#                               FuelMoisture + WindSpeed +
+#                             (1|location/block/year/plot), 
+#                           family=Gamma(link = "log"), 
+#                           control=lme4::glmerControl(optimizer="bobyqa", 
+#                                                      optCtrl=list(maxfun=100000)) )) )
+
+canopy_RH <- with(imp_sc, suppressMessages(
+              lme4::glmer(MaxC ~ RH + tHa +
                               FuelMoisture + WindSpeed +
@@
 
-  canopy_imp_pooled <- pool(canopy_imp)$pooled  %>%
+D1(canopy_VPD, canopy_RH) # not different
+
+  canopy_RH_pooled <- pool(canopy_RH)$pooled  %>%
                       as_tibble()
@@
     full_join(
-      summary(pool(canopy_imp)) %>% 
+      summary(pool(canopy_RH)) %>% 
         as_tibble() %>%
@@
 
-      confint.mipo(pool(canopy_imp)) %>%
+      confint.mipo(pool(canopy_RH)) %>%
         as_tibble() %>%
@@
 # Fit model 
-  soil_imp <- with(imp_sc, suppressMessages(
-                  lme4::glmer(log(SoilMaxC+1) ~ VPD + tHa +  RH +
+  # soil_VPD <- with(imp_sc, suppressMessages(
+  #                 lme4::glmer(log(SoilMaxC+1) ~ VPD + RH + tHa +
+  #                             FuelMoisture + WindSpeed +
+  #                             (1|location/block/year/plot), 
+  #                   family=Gamma(link = "log"), 
+  #                   control=lme4::glmerControl(optimizer="bobyqa", 
+  #                                              optCtrl=list(maxfun=100000)) )) )
+
+  soil_RH <- with(imp_sc, suppressMessages(
+                  lme4::glmer(log(SoilMaxC+1) ~ RH + tHa +
                               FuelMoisture + WindSpeed +
@@
 
-soil_imp_pooled <- pool(soil_imp)$pooled  %>%
+  D1(soil_VPD, soil_RH) # Not different
+
+soil_RH_pooled <- pool(soil_RH)$pooled  %>%
   as_tibble()
@@
   full_join(
-    summary(pool(soil_imp)) %>% 
+    summary(pool(soil_RH)) %>% 
       as_tibble() %>%
@@
 
-    confint.mipo(pool(soil_imp)) %>%
+    confint.mipo(pool(soil_RH)) %>%
       as_tibble() %>%
@@
                              RH = 'Relative humidity', 
-                             FuelMoisture = 'Fuel moisture', 
-                             VPD = 'Vapor Pressure Deficit')) 
+                             FuelMoisture = 'Fuel moisture')) 
 save(response_CIs, file = paste0(wd, 'r objects/response_CIs.Rdata'))
@@
          P = p.value) %>%
-    select(term, t, P) %>%
+    select(term, df, t, P) %>%
   mutate(across(t:P, ~round(., 2))) %>%
@@
 pca_fc <-  envfit( fb_pca ~ location, imp_raw, 
-                   choices = c(1:3), 
+                   choices = c(1:2), 
                    strata = imp_raw$year, 

### 2022-01-25 paper/FigureGenerator.Rmd
begin revisions

@@
                         MaxC = 'Canopy temp (°C)', 
-                        WindSpeed = 'Wind speed (m/s)', 
+                        WindSpeed = 'Wind speed (km/h)', 
                         ros = 'Rate of spread (m/s)', 

### 2022-01-25 paper/GrasslandFireBehaviorNGP.Rmd
begin revisions

@@
 title: "Weather and fuel as modulators of grassland fire behavior in the northern Great Plains"
-author: "Devan Allen McGranahan^A,C^, Megan E.
+author: 'Devan Allen McGranahan^A,C^, Megan E.
 Zopfi^B^, Kathryn A.
 Yurkonis^B^ ^A^ USDA Agricultural Research Service, Livestock and Range Research Laboratory, Miles City, Montana, USA ^B^ Department of Biology, University of North Dakota, Grand Forks, North Dakota, USA ^C^ Corresponding author.
-Email: Devan.McGranahan@usda.gov"
+Email: Devan.McGranahan@usda.gov'
 output:
@@
 
-Fuel and fire weather conditions interact to affect wildland fire behavior, but little is known about how these variables affect fire behavior in the northern Great Plains of North America.
-Data that are available consist mostly of reports based on temperature-time data from thermocouples, and neither consider rate of spread or statistically test the influence of fire environmental variables.
-We measured fuel load and fuel moisture ahead of prescribed fires in southwestern and central North Dakota, USA, and used a unique multi-channel thermocouple array to measure rate of spread, soil surface temperature, and flame temperature in the plant canopy, which we compared with fire weather data taken from nearby weather stations.
-Canopy temperatures averaged 225^o^C during spring burns in central North Dakota and 250^o^C during fall burns in southwestern North Dakota. 
-Surface temperatures averaged just above 100^o^C, although 50% of observations were 60^o^C or less. 
-Regression analysis indicated that wind speed drove faster rates of spread while higher fuel loads and lower fuel moisture produced higher canopy temperatures. 
+Fuel and weather interact to affect wildland fire behavior, but little is known about associations between these variables in the northern Great Plains of North America. 
+Few studies consider rate of spread or statistically test the influence of fire environmental variables.
+We measured fuel load and moisture ahead of prescribed fires in North Dakota, USA, and used a multi-channel thermocouple array to measure rate of spread, soil surface temperature, and flame temperature, which we compared with fire weather data.
+Aboveground flame temperatures averaged 225$^\circ$C during spring burns in and 250$^\circ$C during fall burns.
+Surface temperatures averaged just above 100$^\circ$C, although 50% of observations were 60$^\circ$C or less.
+Regression analysis indicated that wind speed drove faster rates of spread while higher fuel loads and lower overall fuelbed moisture produced higher flame temperatures.
 None of our measurements explained variability in soil surface temperature, likely because ground-level heating remained low.
-We highlight the differential responses among fire behavior metrics to different components of the wildland fire environment. 
-These results can help wildland fire managers better match burn conditions to desired outcomes and fire ecologists match measurements to specific ecological responses. 
-We also discuss knowledge gaps that remain in understanding how land management and ignition patterns interact to effect variability in fire behavior, and what fire scientists need to learn about the relationships between atmospheric moisture and fuel curing in grassland systems.
+We highlight differential responses among fire behavior metrics to different components of the wildland fire environment.
+Results help wildland fire managers match burn conditions to desired outcomes, and fire ecologists match measurements to specific ecological responses.
+We also discuss remaining knowledge gaps in how management and ignition patterns interact to create variability in fire behavior.
 
@@
 Wind speed has long been recognized as a primary driver of fire behavior, especially in well-cured grassland fuels [@whittaker1961; @cheney1995; @kidnie2015].
-Two measures of atmospheric moisture content—relative humidity and vapor pressure deficit—are associated with fire growth [@evett2008; @reid2010; @sedano2014].
+Two measures of atmospheric moisture content—relative humidity and vapor pressure deficit—are also associated with fire growth [@evett2008; @reid2010; @sedano2014].
 
-*Biotic factors* relate principally to the vegetation biomass available for combustion. 
-Overall, fire spread and energy release rates increase as more fuel is available to burn; the structure and arrangement of plant biomass is also important. 
-Greater fuel load attributable to longer time-since-fire increased fire temperature; spatial variability in fuel load and patchy distribution of fine fuels contributed to spatial variability in fire behavior [@patten1984; @gibson1990; @gomes2020].
+*Biotic factors* relate principally to the amount and nature of plant biomass available for combustion. 
+Overall, fire spread and energy release rates increase as more fuel is available to burn; the structure and arrangement of vegetation is also important. 
+Greater fuel load attributable to longer time-since-fire increases fire temperature, and spatial variability in fuel load and patchy distribution of fine fuels in turn drive variability in fire behavior [@patten1984; @gibson1990; @gomes2020].
 Furthermore, fine-leaved grasses burn more completely and hotter than an equal mass of forbs [@wragg2018].
@@
 For decades, fire ecologists have measured fire behavior as *flame temperature* via various methods, including arrays of temperature-sensitive paints [e.g., @whittaker1961; @smith1966; @bailey1980] or by recording air temperature as a flame front passes over a thermocouple connected to a datalogger [e.g., @strong2013; @russell2015].
+
 Despite its popularity among fire ecologists, temperature alone is a poor response variable fraught by several issues in collecting and interpreting thermocouple data [see review by @mcgranahan2020]. 
-Firstly, a considerable amount of variability in temperature attributable to sensor placement relative to both the ground and the fire. 
+Firstly, a considerable amount of variability in temperature is attributable to sensor placement relative to both the ground and the fire. 
 There is neither a standard for placing thermocouple probes in the wildland fire environment nor consistency in vertical temperature profiles. 
 Most observations of surface fire temperature profiles describe an inverse, linear relationship between height and temperature [@smith1966; @patten1984; @archibold2003], although @ramsay1996 found the highest temperatures at the top of a 1 m profile and the lowest temperatures at 30 cm, while @frost1987 and @bailey1980 present evidence that the highest temperatures occur midway up the profile.
-At least some of this variability might be due to differences in surface vs. canopy temperature among head and back fires [@trollope1978]. 
-Secondly, many factors that contribute to variability in temperatures recorded by thermocouples are attributable to the nature of the sensor rather than the nature of the fire. 
+At least some of this variability might be due to differences in surface vs. flame temperature among head and back fires [@trollope1978].
+
+Secondly, many factors that contribute to variability in temperatures recorded by thermocouples are attributable to the nature of the sensor rather than the nature of the fire [e.g., @walker1968]. 
 Thus, reporting 'device temperatures' alone impedes comparisons between studies; @bova2008 present a standard calibration of thermocouple probes, while @mcgranahan2021b simply uses the timestamps of peak heating across an array of several thermocouples to calculate rate of spread.
@@
 In the North American Great Plains, most reports of grassland fire behavior consist of temperatures derived from thermocouples, and there are few data on rate of spread. 
-Soil surface temperatures in South Dakota tallgrass prairie ranged from 200-500^o^C during spring burns, and were greatest under lower fuel loads [@ohrtman2015]; fires in Saskatchewan mixed grass prairie exceeded 300^o^C 5-10cm above the soil surface in spring, summer, and fall [@archibold2003].
-Mean temperatures in experimental burns in eastern Montana ranged from 172-222^o^C in the summer to 253^o^C in the spring [@strong2013; @russell2015]. 
+Soil surface temperatures in South Dakota tallgrass prairie ranged from 200-500$^\circ$C during spring burns, and were greatest under lower fuel loads [@ohrtman2015]; fires in Saskatchewan mixed grass prairie exceeded 300$^\circ$C 5-10 cm above the soil surface in spring, summer, and fall [@archibold2003].
+Mean temperatures in experimental burns in eastern Montana ranged from 172-222$^\circ$C in the summer to 253$^\circ$C in the spring [@strong2013; @russell2015]. 
 To our knowledge, no studies on grassland fire in the northern Great Plains region has explicitly tested the effect of fire weather on fire behavior.
 
-Our objectives were to (1) describe the range of variability in three measures of fire behavior—rate of spread, canopy temperature, and soil surface temperature—during prescribed burns in typical fuelbeds of the northern US Great Plains, and (2) explain variability in fire behavior in terms of abiotic and biotic conditions.
-Our analysis emphasizes the differential effects of environmental variables among the three responses, and the multidimensional relationship between these different measures of grassland fire behavior.
+Our objectives were to (1) describe the range of variability in three measures of fire behavior—rate of spread, soil surface temperature, and flame temperature 15 cm above soil surface—during prescribed burns in typical fuelbeds of the northern US Great Plains, and (2) explain variability in fire behavior in terms of abiotic and biotic conditions.
+Our analysis emphasizes the differential effects of environmental variables among the three responses, and the multidimensional relationship among these different measures of grassland fire behavior.
 
@@
 
-We sampled 25 prescribed fires at two locations in central and southwestern North Dakota, USA. 
+We sampled 25 prescribed fires at two locations in central and southwestern North Dakota, USA (Maps in Supplemental Information Figure 1). 
 At both locations, sampled grasslands are included in a patch-burn grazing study that requires a portion of each experimental unit to be burned each year [@spiess2020]. 
@@
 Located in a mixed-grass prairie ecoregion, this location has a rolling topography and receives an average of 468 mm annual precipitation. 
-Vegetation is mixed-grass prairie invaded by exotic, C3 grasses; stands are dominated by *Pascopyrum smithii*, *Nassella viridula*, *Poa pratensis*, *Bromus inermis*, *Koeleria macrantha*, *Artemisia* spp., *Solidago* spp., and clumps of *Glycyrrhiza lepidota* and *Symphoricarpos occidentalis*.
+Vegetation is mixed-grass prairie invaded by introduced, C$_3$ grasses; stands are dominated by *Pascopyrum smithii*, *Nassella viridula*, *Poa pratensis*, *Bromus inermis*, *Koeleria macrantha*, *Artemisia* spp., *Solidago* spp., and clumps of *Glycyrrhiza lepidota* and *Symphoricarpos occidentalis*.
 
@@
 Topography is consistently flat.
-Located in a shortgrass prairie ecoregion, these pastures are dominated by introduced C3 grasses *Thinopyrum intermedium*, *Bromus inermis*, *Agropyron cristatum*, and *Poa pratensis*, along with the non-native legume *Medicago sativa*.
+Located in a shortgrass prairie ecoregion, these pastures are dominated by introduced C$_3$ grasses *Thinopyrum intermedium*, *Bromus inermis*, *Agropyron cristatum*, and *Poa pratensis*, along with the non-native legume *Medicago sativa*.
 
@@
 
-We measured each fire with 27 a set of 9, 1 m equilateral triangle plots arranged in a nested fashion such that three 10 m triangles, each containing 3, 1m plots, were placed 100 m apart to form a total plot area of 0.433 ha with 27 sample points positioned at the centroid of each burn unit.
+We measured each fire with a set of 9, 1 m equilateral triangle plots arranged in a nested fashion such that three 10 m triangles, each containing 3, 1m plots, were placed 100 m apart to form a total plot area of 0.433 ha with 27 sample points positioned at the centroid of each burn unit (A detailed schematic of this layout is presented in Supplemental Information Figure 2).
 This fractal design is modified from the Sierpinksi triangle described by @dorrough2007 and applied to measuring wildland fire spread by @mcgranahan2021b.
@@
 We clipped and collected all fuels in a 25 × 25 cm quadrat positioned 0.5 m away from each 1 m triangle vertex; the three measurements per plot were averaged prior to analysis. 
-Fuel samples were initially placed in airtight plastic bags to retain moisture, and then weighed, dried to constant mass at 60^o^ for 48 hours, and reweighed. 
-These data were used to calculate percent fuel moisture content (expressed on a dry-weight basis) and fuel load (kg/ha) for each plot (n = 9 subsamples for each fire).
+Fuel samples were initially placed in airtight plastic bags to retain moisture, and then weighed, dried to constant mass at 60$^\circ$C for 48 hours, and reweighed. 
+These data were used to calculate percent fuel moisture content (expressed on a dry-weight basis) and fuel load (kg m$^{2}$) for each plot (n = 9 subsamples for each fire).
 
-**Fire behavior data** were recorded as temperature (^o^) associated with the advancing flame front at each of the 27 points arranged in 9, 1 m triangles at the center of each burn unit.
+**Fire behavior data** were recorded as temperature ($^\circ$C) associated with the advancing flame front at each of the 27 points arranged in 9, 1 m triangles at the center of each burn unit.
 Data were recorded with the open-source FeatherFlame thermocouple datalogger system [@mcgranahan2021b].
-Briefly, the FeatherFlame system reads overbraided, ceramic fiber-insulated K-type thermocouples (Omega, Norwalk, CT) connected to an Arduino-based datalogger assembled from Adafruit Feather breakout boards (M0 Adalogger, datalogging shield, and OLED display; Adafruit Industries, LLC, New York City, NY) and housed inside water-resistant Pelican (Pelican Products, Inc, Torrance, California) cases.
+Briefly, the FeatherFlame system reads overbraided, ceramic fiber-insulated K-type thermocouples (Omega, Norwalk, CT) connected to an Arduino-based datalogger assembled from Adafruit Feather breakout boards (M0 Adalogger, datalogging shield, and OLED display; Adafruit Industries, LLC, New York City, NY) and housed inside water-resistant Pelican cases (Pelican Products, Inc, Torrance, California).
 The low cost of open-source systems make multiple units more affordable than proprietary data loggers with no sacrifice in data quality [@mcgranahan2021].
 
-At each 1-m triangle—the individual observational unit in the nested plot design—we used four thermocouples to a single FeatherFlame datalogger.
+At each 1-m triangle—the individual observational unit in the nested plot design—we used four thermocouples connected to a single FeatherFlame datalogger.
 Three thermocouples measured flame temperature 15 cm above the soil surface at each 1 m vertex while a fourth thermocouple recorded soil surface temperature at a representative point within the 1 m array.
+Beads of the thermocouple probes extended at least 3 cm from the supporting apparatus, to which the insulated lead was attached with wire. 
 The soil thermocouple was placed on mineral ground, perpendicular to the soil surface, below plant litter. 
@@
 
-For each fire event, we determined the time and value of the maximum temperature (^o^) as the flame front encountered each thermocouple. 
-We calculated the Maximum flame temperature (^o^) 15 cm above soil surface for each plot as the mean of the thermocouple readings from its vertices. 
-We calculated the rate of spread (m/s) of the flame front as it passed through each plot using the maximum temperature timestamps following equations 1 and 2 from @simard1984.
+For each fire event, we determined the time and value of the maximum temperature ($^\circ$C) as the flame front encountered each thermocouple. 
+We calculated the Maximum flame temperature ($^\circ$C) 15 cm above soil surface for each plot as the mean of the thermocouple readings from its vertices. 
+We calculated the rate of spread (m min$^{\text{-}1}$) of the flame front as it passed through each plot using the maximum temperature timestamps following equations from @simard1984, which are presented in full in Supplemental Information.
+.
 
 **Fire weather data** were obtained after each fire from records made available by the North Dakota Agriculture Weather Network, the statewide mesonet system with sensor arrays at both experimental stations.
-We downloaded hourly relative humidity (%), dew point (^o^), air temperature (^o^), and average wind speed (km/h).
-From these data we calculated atmospheric vapor pressure (e) and saturation vapor pressure (es) and used these quantities to determine the vapor pressure deficit (VPD = e - es) for the hour in which each fire behavior observation occurred.
+We downloaded hourly relative humidity (%), dew point ($^\circ$C), air temperature ($^\circ$C), and average wind speed (m s$^{\text{-}1}$).
+From these data we calculated atmospheric vapor pressure (e) and saturation vapor pressure (e$_s$) and used these quantities to determine the vapor pressure deficit (VPD = e - e$_s$) for the hour in which each fire behavior observation occurred.
 These data capture hourly trends in weather at the meso-gamma scale [2-20 km; @orlanski1975], and are reliably connected to our fire behavior measurements via time stamps provided by the dataloggers. 
@@
 
-## Data Analysis
+## Data analysis
 
 Prior to analysis, we ensured statistical power across 167 observational units by using multiple imputation to interpolate missing datapoints, as missing field data occurred for three rate of spread samples (2% of total), 29 fuel load values (17%), and 46 soil surface temperature values (27%).
-We used the multiple imputation method in the *mice* package 
-[@vanbuuren2011] in the R statistical environment [@rct2020] to fill in these missing values. 
-The procedure simulated 100 datasets with different, but reasonable, values for the missing data based on patterns in the existing data. 
-We then scaled all variables to a common range within each imputed dataset, performed regression analysis on each *mice*-generated dataset, and report composite statistical results pooled from the results of the 100 individual regression models.
+We used the multiple imputation method in the *mice* package [@vanbuuren2011] in the R statistical environment [@rct2020] to fill in these missing values. 
+The procedure simulated 50 datasets with different, but reasonable, values for the missing data based on patterns in the existing data. 
+We then scaled all variables to a common range within each imputed dataset, performed regression analysis on each *mice*-generated dataset, and report composite statistical results pooled from the results of the 50 individual regression models.
 
-**Multivariate analysis**
-We first conducted a multivariate analysis to assess composite relationships among the fuel, weather, and fire behavior responses. 
+**Multivariate analysis.---**We first conducted a multivariate analysis to assess composite relationships among the fuel, weather, and fire behavior responses. 
 We used Principal Components Analysis (PCA) fit with the 'rda' function in the *vegan* package for R [@oksanen2017].
@@
 
-**Regression analysis** 
-We assessed weather and fuel effects on three fire behavior response variables: Maximum flame temperature (^o^) 15 cm above the soil surface (mean of three thermocouples), Maximum soil surface temperature (^o^; single thermocouple), and Rate of spread (m/s) through each 1 m equilateral triangle plot.
+**Regression analysis.---**We assessed weather and fuel effects on three fire behavior response variables: Maximum flame temperature ($^\circ$C) 15 cm above the soil surface (mean of three thermocouples), maximum soil surface temperature ($^\circ$C; single thermocouple), and rate of spread (m min$^{\text{-}1}$) through each 1 m equilateral triangle plot.
 Because all three response variables were best modeled with a gamma distribution, we fit generalized linear mixed-effect regression models for each response with the 'glmer' function from the *lme4* package in R [@bates2015].
 Fixed effects consisted of weather and fuel variables, as described above. 
-The random-effect term was constructed to account for spatial non-independence within locations and nested variance within sample plots, and the effect of repeated measurements within each location. 
-Due to concerns about colinearity between relative humidity and vapor pressure deficit because they are derived from the same variables, vapor pressure deficit was excluded from regression analysis for all three response variables.
+The random-effect term was constructed to account for random variance among locations, spatial non-independence within locations and nested variance within sample plots, and the effect of repeated measurements within each location.
+Due to concerns about collinearity between relative humidity and vapor pressure deficit because they are derived from the same variables, vapor pressure deficit was excluded from regression analysis for all three response variables.
 
@@
 The first two axes of the Principal Components Analysis (Fig. 2) explained 86% of overall variance in the fire behavior dataset. 
-The first axis (52% variance explained) was most strongly associated with canopy temperature and rate of spread, while the second axis was more strongly associated with soil surface temperature. 
-Dew point was marginally related (p < 0.05) and inversely associated with canopy temperatures and rate of spread dew point had marginally significant association with variation in fire behavior.
+The first axis (52% variance explained) was most strongly associated with flame temperature and rate of spread, while the second axis was more strongly associated with soil surface temperature. 
+Dew point was marginally related (p < 0.05) and inversely associated with flame temperatures and rate of spread.
 
 Fires at both locations were characterized by considerable variability among sub-plots. 
-Within the plant canopy, half of the fires (13) exceeded 325^o^C, but only four of those fires had >50% of individual sample plots within the burns reach an average of 325^o^C. 
+Above the soil surface, half of the fires (13) exceeded 325$^\circ$C, but only four of those fires had >50% of individual sample plots within the burns reach an average of 325$^\circ$C. 
 Sparse fuels meant that fire did not spread to some individual plots in some burns despite strip ignitions.
-Among plots that burned, less than half exceeded 100^o^C at the soil surface (49 of 121 plots).
-A majority of the 21 fires had at least one plot exceed 100ºC at the soil surface (18 of 21 fires), seven had over half of the plots exceed 100^o^C at the soil surface and only one fire had all plots reach this (a spring Central Grasslands fire).
-Only 10 fires reached 325^o^C on the soil surface, with most of these fires reaching this point in only one plot and never in more than half of the plots.
+Among plots that burned, less than half exceeded 100$^\circ$C at the soil surface (49 of 121 plots).
+A majority of the 21 fires had at least one plot exceed 100$^\circ$C at the soil surface (18 of 21 fires), seven had over half of the plots exceed 100$^\circ$C at the soil surface and only one fire had all plots reach this (a spring Central Grasslands fire).
+Only 10 fires reached 325$^\circ$C on the soil surface, with most of these fires reaching this point in only one plot and never in more than half of the plots.
 
@@
 Fires spread faster with higher wind speeds (t = 2.92, P < 0.01), but no other variable had a statistically-significant association with rate of spread (Table 1). 
-Temperatures in the plant canopy (15 cm above the soil surface) increased as fuel load increased (t = 2.82, P = 0.01) and decreased as fuel moisture increased (t = -2.16, P = 0.04). 
+Aboveground temperatures increased as fuel load increased (t = 2.82, P = 0.01) and decreased as fuel moisture increased (t = -2.16, P = 0.04). 
 No fuel or weather variable included here had statistically-significant associations with soil surface temperature (Table 1).
@@
 These data directly support the safe and effective application of prescribed fire in the region. 
-Some results are straightforward and consistent with decades of fire safety science; e.g., faster rates of fire spread associated with higher wind speed and lower relative humidity. 
-Other results add nuance to an ecological understanding of how fire behavior relates to fire effects—e.g., factors like fuel load and fuel moisture were related with canopy temperature but not soil surface temperature, which suggests that direct effects on belowground plant tissue and soil biota are not correlated with aboveground heating and fire spread.
+Some results are straightforward and consistent with decades of fire safety science; e.g., faster rates of fire spread are associated with higher wind speed and lower relative humidity. 
+Other results add nuance to an ecological understanding of how fire behavior relates to fire effects—e.g., factors like fuel load and overall fuelbed moisture were related with flame temperature but not soil surface temperature, which suggests that direct effects on belowground plant tissue and soil biota are not correlated with aboveground heating and fire spread.
 
@@
 Mean temperatures recorded in this study are consistent with other reports from northern rangelands. 
-In central Alberta grassland, @bailey1980 observed that surface temperatures varied between 110^o^C and 165^o^C for backfires and headfires, respectively, and headfires averaged 200^o^C 15 cm above the ground; temperatures generally tracked with fuel load.
-Surface fires through jack pine barrens in Ontario had a similar range as ours: 140-545^o^C [@smith1966]. 
-In our study, average 15-cm temperatures were 225^o^C during spring burns in central North Dakota and 250^o^C during fall burns in southwestern North Dakota; surface temperatures at both locations generally averaged just above 100^o^C (Fig.
+In central Alberta grassland, @bailey1980 observed that surface temperatures varied between 110$^\circ$C and 165$^\circ$C for backfires and headfires, respectively, and headfires averaged 200$^\circ$C 15 cm above the ground; temperatures generally tracked with fuel load.
+Surface fires through jack pine barrens in Ontario had a similar range as ours: 140-545$^\circ$C [@smith1966]. 
+In our study, average 15-cm temperatures were 225$^\circ$C during spring burns in central North Dakota and 250$^\circ$C during fall burns in southwestern North Dakota; surface temperatures at both locations generally averaged just above 100$^\circ$C (Fig.
 1).
 
-Discrepencies between our data and others reported in the region are consistent with what would be expected when differences in the fire environment are considered.
-For example, @ohrtman2015 reported a wide range of maximum temperatures at the soil surface---150-500^o^C---that was generally explained by variability in annual productivity and clipping frequency. 
-Our fires were also cooler than those reported by @archibold2003 in Saskatchewan: using the mid-point of observations made at 10 cm and 20 cm as a comparison, spring fires reached 314^o^C and fall fires reached 298^o^C.
+Discrepancies between our data and others reported in the region are consistent with what would be expected when differences in the fire environment are considered.
+For example, @ohrtman2015 reported a wide range of maximum temperatures at the soil surface---150-500$^\circ$C---that was generally explained by variability in annual productivity and clipping frequency.
+Our fires were also cooler than those reported by @archibold2003 in Saskatchewan: using the mid-point of observations made at 10 cm and 20 cm as a comparison, spring fires reached 314$^\circ$C and fall fires reached 298$^\circ$C.
 But @archibold2003 also reported substantially lower fuel moisture in each season and they had approximately three times the fuel load, likely due to an absence of grazing management on the remnant prairie.
+A previous study reported similar results---temperatures approaching 500$^\circ$C when fuel loads averaged 2.8-4.5 t$\cdot$ha$^{\text{-}1}$ [@archibold1998]. 
 With greater variability in fuel load and fuel moisture, we might also expect to see these factors have greater influence on aboveground flame temperatures. 
-For example, in Colorado, @augustine2014 observed a strong linear relationship between fuel load and  temperatures ranging from 60-200^o^C, but their fuel load also ranged from 0.2 to 1.2 t/ha.
+For example, in Colorado, @augustine2014 observed a strong linear relationship between fuel load and  temperatures ranging from 60-200$^\circ$C, but their fuel load also ranged from 0.2 to 1.2 t$\cdot$ha$^{\text{-}1}$.
 
@@
 For example, our data match the pattern from Australian grassland, where @cheney1993 found a statistical relationship between rate of spread and both wind speed and fuel moisture, but not total fuel load.
-Their fires traveled at rates similar to the vast majority of ours (1-2 m/s), with slightly higher  fuel loads and slightly lower relative humidity. 
+Their fires traveled at rates similar to the vast majority of ours (1-2 m$\cdot$s$^{\text{-}1}$), with slightly higher fuel loads and slightly lower relative humidity. 
 Wind was by far their most important variable, followed by dead fuel moisture, consistent with patterns in our data. 
-Likewise, @cruz2015 observed fire spread rates up to 2 m/s in fully-cured grassland fuels with fuel loads at the upper end of our samples.  
+Likewise, @cruz2015 observed fire spread rates up to 2 m$\cdot$s$^{\text{-}1}$ in fully-cured grassland fuels with fuel loads at the upper end of our samples.  
 In comparing our results to these studies, it appears the trigonometric method of calculating rate of spread through the thermocouple array [@simard1984] overestimates rate of spread, but there is no indication that using this measure of spread as a scaled response variable in regression models presents error in interpreting the relative effect of fuel and weather predictors.
@@
 One would expect, then, that back fires would have more opportunity to burn down through even thick *P. pratensis* litter to mineral soil. 
-Unfortunately, our results offer little insight into what fuel or weather variables enhance litter combustion, likely because most of our fires never got very hot at the soil surface---50% of our observations were less than 60^o^C, and 60% less than 100^0^C (Fig.
+Unfortunately, our results offer little insight into what fuel or weather variables enhance litter combustion, likely because most of our fires never got very hot at the soil surface---50% of our observations were less than 60$^\circ$C, and 60% less than 100$^\circ$C (Fig.
 1).
 Nor can we differentiate the direction of fire spread with the current trigonometry applied to the triangular thermocouple arrays [@simard1984], although it would theoretically be possible to compare spread direction to wind direction if the latter data were available at a fine enough scale.
-But in these pastures, the functional difference between head and back fires might be moot.
+
+The functional difference between head and back fires in the fuelbeds reported here are likely. 
 Because our fuels were often sparse, sometimes only marginally cured, and prescriptions precluded taking advantage of higher wind or lower relative humidity to mitigate fuel limitations, we often employed substantial interior ignitions using strip, point, flanking, and spiral patterns that sent flame fronts towards our sensors in all possible directions at different times.
-Differences in ignition patterns create additional spatial variability in fire behavior and burn severity [@williams2015]. 
+While @williams2015 did show that different ignition patterns created additional spatial variability in fire behavior, the effect of shorter line ignitions and spot ignitions was to mitigate the high severity of wildfire and long line ignitions in highly-flammable spinifex fuelbeds.
+In our case, we had to manipulate ignition pattern just to get fire to carry.
 Our data are certainly useful in describing the variability in fire behavior across these burns, but do not inform the relationship between ignition pattern and fire behavior. 
-Thus, future research on fire behavior in the northern Great Plains should use experimental plots with consistent fuelbeds to explicitly compare head and back fires set via line ignitions, akin to the experimental burning program described by @cruz2015 in Australia.
+Thus, future research on fire behavior in the northern Great Plains should (1) use experimental plots with consistent fuelbeds to explicitly compare head and back fires set via line ignitions, akin to the experimental burning program described by @cruz2015 in Australia, and (2) attempt to at least address severity, if not fire behavior, in wildfire scenarios via remote sensing and/or modelling.
 
 A more detailed analysis of fuelbed effects on fire behavior also ought to separate fuels into live and dead components, and consider the moisture content of each along with fuel load ratios. 
-We report here the fine fuel moisture content of the entire fuelbed, consistent with descriptive, post hoc statistical approaches to describing fire behavior [@trollope1985; @mcgranahan2016; @cruz2015].
+Explicitly measuring litter moisture might also be valuable, especially when variability in belowground heating is expected to influence first-order fire effects. 
+We report here the overall fuel moisture content of the entire fuelbed, consistent with descriptive, post hoc statistical approaches to describing fire behavior [@trollope1978; @trollope1985; @bidwell1992; @trollope2002; @mcgranahan2016].
 But predictive fire behavior models accommodate inputs for live and dead fuel categories [@scott2005], and @kidnie2015a found that four categories of live, dead, and senescent fuels best represented differences in grassland fuel moisture scenarios.
-Although parsing live and dead fuel moisture in the present analysis would probably not better explain variability in our dataset, it would likely contribute to better predictions of fire behavior relative to management objectives if information on fuel moisture content were available prior to ignition.
+With this in mind, @cruz2015 employed a hybrid approach in which fuel moisture was measured for the various components, from which a weighted average was used as a predictor variable in regression analysis.
+
+Often, time and resource constraints preclude the separation of fine plant material by live and dead class, and overall fuelbed moisture content is the best available data for managers. 
+Although parsing live and dead fuel moisture in the present analysis would probably not better explain variability in our dataset, it would likely contribute to better predictions of fire behavior relative to management objectives if information on overall fuelbed moisture content were available prior to ignition.
 Unfortunately, the standard clipping and drying method is not compatible with providing day-of fuel moisture data, and visual assessments based on color tend to over-predict curedness [@kidnie2015a]. 
@@
 In fact, @srock2018 suggest vapor pressure deficit might be a better measure of atmospheric moisture content for fire predictions, but the Hot-Dry-Windy index they developed to incorporate vapor pressure deficit operates at synoptic scales beyond the spatial extent and operational periods of prescribed burns.
-Given that substantial changes in atmospheric moisture changes in recent decades are expected to strengthen over the 21^st^ century [@seager2015; @ficklin2017], understanding how these dynamics affect fire behavior will be an essential component of managing resilent fire regimes.
+Given that substantial changes in atmospheric moisture changes in recent decades are expected to strengthen over the 21^st^ century [@seager2015; @ficklin2017], understanding how these dynamics affect fire behavior will be an essential component of managing resilient fire regimes.
+
+This study is novel in that it examines the fire environment at a spatial scale consistent with land management in the region using realistic ignition scenarios. 
+To our knowledge, no other study in the northern Great Plains has reported the behavior of fires larger than experimental plots. 
+Integrating research into management almost invariably requires trade-offs; two already discussed above include (1) measuring only the overall moisture content of the entire fuelbed, being precluded from parsing fuel into live, dead, and litter components, and (2) measuring two-dimensional rate of spread of fire fronts within the burn unit without being able to associate them with wind direction.
+But these are the respective conditions under which prescribed fire managers in the region decide whether to burn, and ensure fire spread objectives are met. 
+Research conducted at the spatial scales at which management occurs helps managers trust the transfer of knowledge from studies to working landscapes [@sayre2005; @cacciapaglia2012].
+Critically, all of our sensors were greater than 50 m from the initial fire line, identified in simulations as a
 
@@
 
-We appreciate support from the North Dakota State Agricultural Experiment Station, including K. Sedivec at Central Grasslands REC and C. Schauer and B. Geaumont at Hettinger REC. 
+We appreciate support from the North Dakota State Agricultural Experiment Station, including K. Sedivec at the Central Grasslands REC and C. Schauer and B. Geaumont at the Hettinger REC. 
 We recognize funding from the University of North Dakota Department of Biology, USDA-NIFA Hatch project number ND02393, and USDA-NIFA AFRI award number 2018-67020-27856. 
@@
 J. Spiess, J. Cutter, M. Lakey, B. Poling, and A. Steele assisted with sensor deployment and data collection. 
+We appreciate the helpful comments contributed by two anonymous reviewers.
+
+The data that support this study will be shared upon reasonable request to the corresponding author.
+
+The authors declare no conflicts of interest.
 

### 2022-01-25 paper/SupplementaryInfo.Rmd
begin revisions

---
title: Weather and fuel as modulators of grassland fire behavior in the northern Great Plains
subtitle: Supplementary information
date: '`r Sys.Date()`'
author: DA McGranahan, ME Zopfi, KA Yurkonis
header-includes:
  - \usepackage{amsmath}
  - \renewcommand{\familydefault}{\sfdefault}
  - \usepackage{hyperref}
output:
  pdf_document:
    highlight: tango
---

```{r setup, echo=FALSE, warning=FALSE, message = FALSE, results='hide'}
knitr::opts_chunk$set(message = FALSE, warning=FALSE, 
                      echo=FALSE, eval=FALSE)
```

# Study locations

```{r map, eval = TRUE, out.width='100%', fig.align='center', fig.cap='Main map: Study locations (triangles) within two EPA level 3 ecoregions in North Dakota. Inset: State of North Dakota (dark blue) within the Great Plains (light blue) with respect to the continental United States. \\label{fig:map}'}
knitr::include_graphics('figures/map.png')
```
\clearpage

# Data collection

For even more information about the FeatherFlame datalogger system, please see:

* McGranahan DA (2021) FeatherFlame: An Arduino-based thermocouple datalogging system to record wildland fire flame temperatures \emph{in agris}. Rangeland Ecology \& Management 76, 43–47 [DOI: 10.1016/j.rama.2021.01.008](https://doi.org/10.1016/j.rama.2021.01.008)

* [diyfirescience.info](https://diyfirescience.info)

```{r scheme, eval=TRUE, out.width='100%', fig.align='center', fig.cap='Schematic representation of the Sierpinski Triangle used to deploy 27 thermocouples across 9, 1 m equilateral triangles. Total plot area = 0.433 ha. \\label{fig:triangle}'}
knitr::include_graphics('figures/SierpinskiTriangle.png')
```

Simard et al. (1982) describe how rate of spread $r$ through an equilateral triangle with sides of length $D$ can be determined from the arrival times of the flame front at each point in the triangle sequentially\textemdash $t_1$, $t_2$, and $t_3$.

If $t_1 \neq t_2$,

\begin{equation} \label{eq:simard1}
\theta = tan\textsuperscript{-1} \left( \dfrac{2t_3 - t_2 - t_1}{\sqrt{3}\cdot (t_2 - t_1)}\right) 
\end{equation}

and rate of spread $r$ is

\begin{equation} \label{eq:simard2}
r = \dfrac{D\cdot cos\theta}{t_2 - t_1}
\end{equation}

otherwise, if $t_1 = t_2$, $\theta = 90$ and rate of spread $r$ is

\begin{equation} \label{eq:simard3}
r = D (\sqrt{3}/2)/({t_3 - t_1})
\end{equation}

\clearpage

```{r example_script}
### S E T U P 
##
# Additional packages required for analysis
  pacman::p_load(tidyverse, readr, mice, broom.mixed, vegan, lubridate)
# Additional script available via GitHub
  source('https://raw.githubusercontent.com/cran/mice/master/R/mipo.R')
#
##
### D A T A   P R E P A R A T I O N
##
# Load raw data directly from GitHub
  fp = 'https://raw.githubusercontent.com/devanmcg/SpatialFireBehavior/main'
#
# Data wrangling
#
 AllData <-  
    read_csv(paste0(fp, "/data/fromMZ/CompiledData2.csv")) %>%
    filter(location != "OAK") %>%
      mutate(date = as.Date(date, format = "%m/%d/%Y"),
               L = str_remove(location, "REC"), 
               B = str_sub(block, 1,3), 
               Ps = str_replace(pasture, "[.]", ""), 
               Ps = str_sub(Ps, 1,2), 
               patch = str_replace(patch, "[.]", ""),
               y = format(date, "%y")) %>%
        unite("FireCode", c(L,B,Ps,patch,y), sep=".") %>%
    mutate(time = str_remove(MaxTempTime, "[.]+[0-9]"))%>%
    unite(timestamp, c(date, time), sep = " ") %>%
    mutate(timestamp = as.POSIXct(timestamp, format = "%Y-%m-%d %H:%M:%S")) %>%
        select(FireCode, timestamp, plot, array, TC, MaxC, 
               AirTemp, RH, dpC, WindSpeed, 
               LAI, FMC, KgHa) 
# Isolate soil surface temperature (TC 4)
  SoilTemp <-
    filter(AllData, TC == 4) %>%
      select(FireCode, plot, array, MaxC) %>%
        rename(SoilC = MaxC)
# Summarize array-level data 
  DataMeans <- 
    AllData %>%
      filter(TC %in% c('1', '2', '3')) %>% 
      select(-timestamp) %>%
      pivot_longer(cols = c(MaxC:KgHa), 
                   names_to = "var",
                   values_to = "value") %>%
      group_by(FireCode, plot, array, var) %>%
        summarize(Mean = mean(value) ) %>%
      ungroup() %>%
      pivot_wider(names_from = var, 
                  values_from = Mean) 
# Calculate Vapor Pressure Deficit
  DataMeans <- 
    DataMeans %>%
      mutate(e  = 6.11*(10^((7.5*dpC)/(237.3+dpC))), 
             es = 6.11*(10^((7.5*AirTemp)/(237.3+AirTemp))), 
             VPD = es - e) %>%
      select(-e, -es)
# Calculate rate of spread by arrival time of flame front at sensors 
  D = 1   # Distance between thermocouples (m)
  ROS <- 
    AllData %>%
      filter(TC %in% c('1', '2', '3')) %>%
      mutate(timestamp = format(timestamp, "%H:%M:%S"), 
             ArrivalTime = seconds(hms(timestamp)) ) %>%
    select(FireCode, plot, array, ArrivalTime) %>%
    group_by(FireCode, plot, array) %>%
    arrange(ArrivalTime, .by_group = TRUE) %>% 
    mutate(position = order(order(ArrivalTime, decreasing=FALSE)), 
           position = recode(position, "1"="a", "2"="b", "3"="c"), 
           ArrivalTime = as.numeric(ArrivalTime) /60 ) %>%
    spread(position, ArrivalTime)  %>%
    ungroup %>% 
    # Apply equations from Simard et al. (1984)
    mutate( theta_rad = atan((2*c - b - a) / (sqrt(3)*(b - a))), 
            ros = case_when(
              a == b ~ (sqrt(3) / 2) / (c - a) , 
              a != b ~  (D*cos(theta_rad) / (b - a) ) 
            )) %>%
    select(-a, -b, -c, -theta_rad)
#
# Create final tibble for analysis 
#
  AnalysisData  <- 
    full_join(DataMeans, ROS) %>%
              left_join(SoilTemp) %>%
      filter( ros <= 40,       # remove outliers
              MaxC >= 40) %>%  # ditto
      rename(FuelMoisture = FMC, 
             SoilMaxC = SoilC) %>%
        mutate(FuelMoisture = ifelse(FuelMoisture >= 0, 
                                      FuelMoisture, NA), 
               FuelMoisture = FuelMoisture * 100) %>%
      separate(FireCode, into = c("location", "block", "pasture", 
                           "patch", "year"), 
               remove = F)
#
# Imputing missing values with mice package
#
  # Calculate imputed datasets on scaled data
    imp_sc <- AnalysisData %>% 
            select(-LAI, -JD) %>%
            mutate(ros = ifelse(ros >= 12, NA, ros),    # remove outliers
                   tHa = ifelse(tHa >= 4, NA, tHa)) %>% # ditto
            mutate_at(vars(AirTemp:tHa), ~as.numeric(scale(., center=F))) %>%
                      mutate(across(location:array, as.factor)) %>%
                       mice(m=50, seed = 23109, print=F)
#
##
###  S T A T I S T I C A L   M O D E L S
##
#
# Mixed-effect regression models on imputed datasets
#
# Rate of spread
#   
  # Fit model 
    ros_RH <- 
      with(imp_sc, suppressMessages(
            lme4::glmer(ros ~ RH + tHa +
                              FuelMoisture + WindSpeed +
                              (1|location/block/year/plot), 
                        family=Gamma(link = "log"), 
                        control=lme4::glmerControl(optimizer="bobyqa", 
                                      optCtrl=list(maxfun=100000)) )) )
  # Get terms
    ros_terms <- 
      full_join(
      summary(pool(ros_RH)) %>% 
        as_tibble() %>%
        rownames_to_column("row"), 
      confint.mipo(pool(ros_RH)) %>%
        as_tibble() %>%
        rownames_to_column("row") ) 
#    
# Maximum canopy temperature  
#
  # Fit model
    canopy_RH <- 
      with(imp_sc, suppressMessages(
            lme4::glmer(MaxC ~ RH + tHa +
                            FuelMoisture + WindSpeed +
                          (1|location/block/year/plot), 
                        family=Gamma(link = "log"), 
                        control=lme4::glmerControl(optimizer="bobyqa", 
                                      optCtrl=list(maxfun=100000)) )) )
  # Get terms
    canopy_terms <- 
      full_join(
        summary(pool(canopy_RH)) %>% 
          as_tibble() %>%
          rownames_to_column("row"), 
        confint.mipo(pool(canopy_RH)) %>%
          as_tibble() %>%
          rownames_to_column("row") ) 
#
# Maximum soil surface temperature  
#
  # Fit model 
    soil_RH <- 
      with(imp_sc, suppressMessages(
            lme4::glmer(log(SoilMaxC+1) ~ RH + tHa +
                        FuelMoisture + WindSpeed +
                        (1|location/block/year/plot), 
              family=Gamma(link = "log"), 
              control=lme4::glmerControl(optimizer="bobyqa", 
                            optCtrl=list(maxfun=100000)) )) )
  # Get terms
    soil_terms <- 
      full_join(
        summary(pool(soil_RH)) %>% 
          as_tibble() %>%
          rownames_to_column("row"), 
        confint.mipo(pool(soil_RH)) %>%
          as_tibble() %>%
          rownames_to_column("row") ) 
#
# Multivariate analysis
#
  # Reduce mids object to tibble   
    imp_raw <- 
      complete(imp_sc, 'long') %>%
      as_tibble() %>%
      unite("TreeID", c(location, block, pasture,
                        year, plot, array), sep = ".") %>%
      select(-patch, -.id, -.imp, -FireCode) %>%
      pivot_longer(names_to = "response", 
                   values_to = "values", 
                   -TreeID) %>%
      group_by(TreeID, response) %>%
      summarize(value = median(values)) %>%
      ungroup() %>%
      pivot_wider(names_from = response, 
                  values_from = value) %>%
      separate(TreeID, c("location", "block", "pasture",
                         "year", "plot", "array")) %>% 
      mutate(across(location:array, as.factor))
  # Fire behavior PCA 
    fb_d <- 
      imp_raw %>%
      select(MaxC, ros, SoilMaxC) 
    fb_pca <- rda(fb_d ~ 1, 'euc', scale = T)
  # Test differences between locations 
    envfit(fb_pca ~ location, imp_raw, 
           choices = c(1:2), 
           strata = imp_raw$year, 
           199)$factors
# Test fire weather against PCA
   envfit(fb_pca ~ MaxWindSpeed+AirTemp+dpC+RH+VPD, 
          data = imp_raw, 
          choices = c(1:3), 
          strata = imp_raw$location)
```

# Script

```{r script, ref.label=knitr::all_labels(label %in% c('example_script') ), echo=TRUE, eval=FALSE}
```

### 2022-01-25 r script/DataAndDistributions.R

@@ -7,13 +7,19 @@ GGally::ggpairs(AnalysisData, columns = 9:17)
 # Distribution
 
 AnalysisData %>% 
-  ggplot(aes(x=ros)) + theme_bw(14) + 
+  ggplot(aes(x=ros/60)) + theme_bw(14) + 
   geom_histogram(aes(y=..density..),      
-                 binwidth=0.5,
+                 binwidth=0.01,
                  colour="black", 
                  fill="lightgreen") +
   geom_density(alpha=0.2, 
                fill="#FF6666") 
+
+AnalysisData %>%
+  filter(ros >= 2.5) %>%
+  select(FireCode, plot, array, MaxC, ros) %>%
+  arrange(desc(ros)) %>%
+  write_csv('HighestROS.csv')
 
 AnalysisData %>% 
   ggplot(aes(x=MaxC)) + theme_bw(14) + 

### 2022-01-25 r script/NorthDakotaMapping.R

@@ -1,0 +1,115 @@
+pacman::p_load(tidyverse, sf, geonames, 
+               grid, gridExtra)
+source('https://raw.githubusercontent.com/devanmcg/rangeR/master/R/CustomGGplotThemes.R')
+
+epa <- read_sf('S:/DevanMcG/GIS/SpatialData/US/EPAecoregions/L3', 
+               'us_eco_l3_state_boundaries')
+aea <- st_crs(epa)
+
+load('C:/Users/devan.mcgranahan/GithubProjects/GreatPlainsFire/gis/Robjects/us_sf.Rdata')
+us_sf <- us_sf %>%
+          st_transform(aea) %>%
+            st_simplify()
+
+us_states <- 
+  epa %>%
+  select(STATE_NAME) %>%
+  mutate(area = st_area(.)) %>%
+  group_by(STATE_NAME) %>%
+  summarize(area = sum(area)) %>%
+  ungroup() %>%
+  select(-area)  %>%
+  st_simplify()
+
+nd <- filter(us_states, STATE_NAME == "North Dakota")
+
+nd_l3 <- st_crop(gp_l3, nd) %>%
+  filter(L2 != 'Temperate Prairies') %>%
+  mutate(area = st_area(.)) %>%
+  group_by(L3) %>%
+  arrange(desc(area)) %>%
+  slice(1)
+
+# Find map points 
+pacman::p_load(geonames)
+options(geonamesUsername="devanmcg")
+
+points <- tibble(feature=c("city", "capital", "rec", "rec"),
+                 name=c("Fargo", "Bismarck", "Hettinger", "Streeter"), 
+                 state=c("ND")) 
+map_points <- 
+  points %>%
+  split(.$name) %>%
+  purrr::map( ~ GNsearch(name_equals = .x$name, 
+                         country = "US", 
+                         adminCode1=.x$state, 
+                         featureClass="P")) %>%
+  map_dfr(~ (.))  %>%
+  full_join(points) %>%
+  mutate(state = adminName1, 
+         long=as.numeric(lng), 
+         lat = as.numeric(lat)) %>%
+  select(feature, name, state, long, lat) %>%
+  st_as_sf(coords = c('long', 'lat'), 
+           crs = 4326) %>%
+  st_transform(aea) %>%
+  mutate(name = case_when(
+    name == "Hettinger" ~ "Hettinger REC", 
+    name == "Streeter" ~ "Central Grasslands REC", 
+    TRUE ~ name
+  ))
+
+mapping <- lst(us = us_states, 
[…]

### 2022-01-25 r script/RevisionAnalysis.R

@@ -1,0 +1,92 @@
+pacman::p_load(tidyverse, readr, mice, broom.mixed, vegan, lubridate)
+# Additional script available via GitHub
+source('https://raw.githubusercontent.com/cran/mice/master/R/mipo.R')
+#
+##
+### D A T A   P R E P A R A T I O N
+##
+# Load raw data directly from GitHub
+fp = 'https://raw.githubusercontent.com/devanmcg/SpatialFireBehavior/main'
+#
+# Data wrangling
+#
+AllData <-  
+  read_csv(paste0(fp, "/data/fromMZ/CompiledData2.csv")) %>%
+  filter(location != "OAK") %>%
+  mutate(date = as.Date(date, format = "%m/%d/%Y"),
+         L = str_remove(location, "REC"), 
+         B = str_sub(block, 1,3), 
+         Ps = str_replace(pasture, "[.]", ""), 
+         Ps = str_sub(Ps, 1,2), 
+         patch = str_replace(patch, "[.]", ""),
+         y = format(date, "%y")) %>%
+  unite("FireCode", c(L,B,Ps,patch,y), sep=".") %>%
+  mutate(time = str_remove(MaxTempTime, "[.]+[0-9]"))%>%
+  unite(timestamp, c(date, time), sep = " ") %>%
+  mutate(timestamp = as.POSIXct(timestamp, format = "%Y-%m-%d %H:%M:%S")) %>%
+  select(FireCode, timestamp, plot, array, TC, MaxC, 
+         AirTemp, RH, dpC, WindSpeed, 
+         LAI, FMC, KgHa) 
+# Isolate soil surface temperature (TC 4)
+SoilTemp <-
+  filter(AllData, TC == 4) %>%
+  select(FireCode, plot, array, MaxC) %>%
+  rename(SoilC = MaxC)
+# Summarize array-level data 
+DataMeans <- 
+  AllData %>%
+  filter(TC %in% c('1', '2', '3')) %>% 
+  select(-timestamp) %>%
+  pivot_longer(cols = c(MaxC:KgHa), 
+               names_to = "var",
+               values_to = "value") %>%
+  group_by(FireCode, plot, array, var) %>%
+  summarize(Mean = mean(value) ) %>%
+  ungroup() %>%
+  pivot_wider(names_from = var, 
+              values_from = Mean) 
+# Calculate Vapor Pressure Deficit
+DataMeans <- 
+  DataMeans %>%
[…]

### 2022-01-27 paper/FigureGenerator_r1.Rmd
begin revisions
First draft of first revision

--- 2021-08-06 paper/FigureGenerator.Rmd
@@
 output:
+  #html_document
   word_document:
     reference_docx: word_template.docx
-    keep_md: yes
 ---
@@
 violin_cap <- f_nums(name = "violin_plots", 
-                   caption = "Distribution of weather, fuel, and fire behavior data for fires in southwestern North Dakota (Hettinger, dark maroon) and central North Dakota (Central Grasslands, light blue) sampled from 2017 to 2019.
-Summary statistics include median (horizontal gray lines) and means (triangles).
-")
+                   caption = "Distribution of weather, fuel, and fire behavior data for fires in southwestern North Dakota (Hettinger, dark maroon) and central North Dakota (Central Grasslands, light blue).
+Horizontal gray lines denote 25%, 50% (median) and 75% quantiles; triangles are arithmetic means.
+Means and standard deviation are also reported in Supplemental Information Table 1.")
 ```
@@
 `r f_nums('violin_plots')`
-```{r data_summary_gg, fig.cap = violin_cap, fig.height=7.5, fig.width=6}
+```{r data_summary_gg, fig.cap = violin_cap, fig.height=8, fig.width=6}
 LongData <- 
@@
            tHa = ifelse(tHa >=4, NA, tHa)) %>%
+    # convert units 
+    mutate(WindSpeed = WindSpeed * 0.28,  # km/hr to m/s
+           tHa = tHa * 0.1) %>%           # t/ha to kg/m^s
     pivot_longer(cols = AirTemp:tHa, 
@@
            var = recode(var, 
-                        AirTemp = 'Air temperature (°C)', 
-                        RH = 'Relative humidity (%)', 
-                        dpC = 'Dew point (°C)', 
-                        FuelMoisture = 'Fuel moisture (%)', 
-                        tHa = 'Fuel load (t/ha)',
-                        MaxC = 'Canopy temp (°C)', 
-                        WindSpeed = 'Wind speed (m/s)', 
-                        ros = 'Rate of spread (m/s)', 
-                        SoilMaxC = 'Soil surface temp (°C)'), 
+                        AirTemp = 'Air~temperature~(degree*C)', 
+                        RH = 'Relative~humidity~(\'%\')', 
+                        dpC = 'Dew~point~(degree*C)', 
+                        FuelMoisture = 'Total~fuel~moisture~(\'%\')', 
+                        tHa = 'Fuel~load~(kg~m^-2)',
+                        MaxC = 'Flame~temp~(degree*C)', 
+                        WindSpeed = 'Wind~speed~(m~s^-1)', 
+                        ros = 'Spread~rate~(m~min^-1)', 
+                        SoilMaxC = 'Soil~surface~temp~(degree*C)'), 
            location = recode(location, 
@@
              fill = location)) + theme_bw(14) +
-    geom_violin(draw_quantiles = 0.5, 
+    geom_violin(draw_quantiles = c(0.25, 0.5, 0.75), 
                 color = "darkgrey", 
@@
        title = "Weather") + 
-  facet_wrap( ~ var, scales = 'free') +
+  facet_wrap( ~ var, scales = 'free', labeller=label_parsed) +
   scale_fill_manual(name = "Location", 
@@
              fill = location)) + theme_bw(14) +
-        geom_violin(draw_quantiles = 0.5, 
+        geom_violin(draw_quantiles = c(0.25, 0.5, 0.75), 
                 color = "darkgrey", 
@@
        title = "Fuel") + 
-  facet_wrap( ~ var, scales = 'free') +
+  facet_wrap( ~ var, scales = 'free', labeller=label_parsed) +
   scale_fill_manual(name = "Location", 
@@
                fill = location)) + theme_bw(14) +
-          geom_violin(draw_quantiles = 0.5, 
+          geom_violin(draw_quantiles = c(0.25, 0.5, 0.75), 
                 color = "darkgrey", 
@@
          title = "Fire behavior") + 
-    facet_wrap( ~ var, scales = 'free') +
+    facet_wrap( ~ var, scales = 'free', labeller=label_parsed) +
     scale_fill_manual(name = "Location", 
@@
 
+\newpage
+
 ```{r }
 pca_cap <- f_nums(name = "pca", 
-                   caption = "Principal Components Analysis of fire behavior data (response variables in blue; rate of spread (ROS), temperature above surface (canopy ºC), and temperature at soil surface (soil ºC) for prescribed burns on rangeland at Hettinger (H), in southwestern North Dakota, and Central Grasslands (CG), in central North Dakota.
+                   caption = "Principal Components Analysis of fire behavior data (response variables in blue; rate of spread (ROS), temperature above surface (flame ºC), and temperature at soil surface (soil ºC) for prescribed burns on rangeland at Hettinger (H), in southwestern North Dakota, and Central Grasslands (CG), in central North Dakota.
 No difference between locations (P = 0.11).
@@
 ```{r pca_gg, fig.height=4, fig.width=6.5, fig.cap=pca_cap }
+pca_scores$species <-
+  pca_scores$species %>%
+    mutate(response = str_replace_all(response, 'canopy', 'flame')) 
 # ggplot PCA
@@
 
+\newpage
+
 ```{r }
 CIs_cap <- f_nums(name = "CI_plots", 
-                   caption = "Regression coefficients and 95% confidence intervals for fuel and weather terms from models for maximum temperature at 15 cm above the soil surface (canopy), maximum temperature at the soil surface (soil) and rate of spread.
+                   caption = "Regression coefficients and 95% confidence intervals for fuel and weather terms from models for maximum temperature at 15 cm above the soil surface (flame), maximum temperature at the soil surface (soil), and rate of spread.
 ")
@@
 `r f_nums('CI_plots')`
-```{r CI_gg, fig.cap = CIs_cap, fig.width=8, fig.height = 5}
+```{r CI_gg, fig.cap = CIs_cap, fig.width=8, fig.height = 4}
 # Plot confidence intervals
@@
   mutate(response =case_when(
-                            response == "Temp. above surface" ~ "°C, canopy", 
+                            response == "Temp. above surface" ~ "°C, flame", 
                             response == "Temp. at surface" ~ "°C, soil", 

### 2022-01-27 paper/GrasslandFireBehaviorNGP.Rmd
First draft of first revision

@@
 Surface temperatures averaged just above 100$^\circ$C, although 50% of observations were 60$^\circ$C or less.
+Fires spread at an average of 2.5 m min$^{\text{-}1}$. 
 Regression analysis indicated that wind speed drove faster rates of spread while higher fuel loads and lower overall fuelbed moisture produced higher flame temperatures.
@@
 *Biotic factors* relate principally to the amount and nature of plant biomass available for combustion. 
-Overall, fire spread and energy release rates increase as more fuel is available to burn; the structure and arrangement of vegetation is also important. 
+Overall, energy release rates increase as more fuel is available to burn; the structure and arrangement of vegetation is also important. 
 Greater fuel load attributable to longer time-since-fire increases fire temperature, and spatial variability in fuel load and patchy distribution of fine fuels in turn drive variability in fire behavior [@patten1984; @gibson1990; @gomes2020].
@@
 Finally, fuel moisture content is an especially important driver outside of the highly-cured context of wildfire seasons [@sparling1966; @kidnie2015]. 
-Together, variability in flammability traits and curing rates among species that comprise grassland fuelbeds contributes to variability in fire behavior [@mcgranahan2016; @cardoso2018].
+Together, variability in flammability traits and curing rates among species that comprise grassland fuelbeds contributes to variability in fire behavior [@cruz2015; @kidnie2015; @mcgranahan2016; @cardoso2018].
 
@@
 Our objectives were to (1) describe the range of variability in three measures of fire behavior—rate of spread, soil surface temperature, and flame temperature 15 cm above soil surface—during prescribed burns in typical fuelbeds of the northern US Great Plains, and (2) explain variability in fire behavior in terms of abiotic and biotic conditions.
-Our analysis emphasizes the differential effects of environmental variables among the three responses, and the multidimensional relationship among these different measures of grassland fire behavior.
+Our analysis emphasizes the differential effects of environmental variables among the three responses, and the multidimensional relationship among these different measures of grassland fire behavior during prescribed burns conducted at scales consistent with land management in the region.
 
@@
 At both locations, sampled grasslands are included in a patch-burn grazing study that requires a portion of each experimental unit to be burned each year [@spiess2020]. 
-Fire sizes ranged from 8-16 ha. 
+The majority of the burn units were 16 ha, while a small set were 8 ha. 
 Typical ignition patterns consisted of downwind backing fires followed by either ring ignition and primarily head fire spread, when fuels were conducive; when fuels were sparse or higher-moisture, flanking fires and strip ignitions were employed as necessary to ensure fire spread through the entire burn unit.
@@
 This fractal design is modified from the Sierpinksi triangle described by @dorrough2007 and applied to measuring wildland fire spread by @mcgranahan2021b.
-Although the logarithmically-scaled nested design was intended for geospatial analysis of point-level data, for our analyses here we calculate averages from the finest (1 m) scale.
+Although the logarithmically-scaled nested design was intended for geospatial analysis of point-level data, for our analyses here we calculate averages from the finest (1 m) scale, consistent with the method of locating multiple microplots within larger burned areas to characterize spatial variability within fires [@fernandes2000].
 
@@
 **Fire behavior data** were recorded as temperature ($^\circ$C) associated with the advancing flame front at each of the 27 points arranged in 9, 1 m triangles at the center of each burn unit.
-Data were recorded with the open-source FeatherFlame thermocouple datalogger system [@mcgranahan2021b].
-Briefly, the FeatherFlame system reads overbraided, ceramic fiber-insulated K-type thermocouples (Omega, Norwalk, CT) connected to an Arduino-based datalogger assembled from Adafruit Feather breakout boards (M0 Adalogger, datalogging shield, and OLED display; Adafruit Industries, LLC, New York City, NY) and housed inside water-resistant Pelican cases (Pelican Products, Inc, Torrance, California).
+Data were recorded with the open-source FeatherFlame thermocouple datalogger system [@mcgranahan2021b], logging at 1.5 Hz.
+Briefly, the FeatherFlame system reads overbraided, ceramic fiber-insulated, 20-gauge K-type thermocouples (Omega, Norwalk, CT) connected to an Arduino-based datalogger assembled from Adafruit Feather breakout boards (M0 Adalogger, datalogging shield, and OLED display; Adafruit Industries, LLC, New York City, NY) and housed inside water-resistant Pelican cases (Pelican Products, Inc, Torrance, California).
 The low cost of open-source systems make multiple units more affordable than proprietary data loggers with no sacrifice in data quality [@mcgranahan2021].
@@
 The soil thermocouple was placed on mineral ground, perpendicular to the soil surface, below plant litter.
-Dataloggers recorded thermocouple temperatures at 1.5 Hz.
 
 For each fire event, we determined the time and value of the maximum temperature ($^\circ$C) as the flame front encountered each thermocouple. 
-We calculated the Maximum flame temperature ($^\circ$C) 15 cm above soil surface for each plot as the mean of the thermocouple readings from its vertices. 
+We calculated the maximum flame temperature ($^\circ$C) 15 cm above soil surface for each plot as the mean of the thermocouple readings from its vertices. 
 We calculated the rate of spread (m min$^{\text{-}1}$) of the flame front as it passed through each plot using the maximum temperature timestamps following equations from @simard1984, which are presented in full in Supplemental Information.
-.
 
@@
 Sparse fuels meant that fire did not spread to some individual plots in some burns despite strip ignitions.
-Among plots that burned, less than half exceeded 100$^\circ$C at the soil surface (49 of 121 plots).
-A majority of the 21 fires had at least one plot exceed 100$^\circ$C at the soil surface (18 of 21 fires), seven had over half of the plots exceed 100$^\circ$C at the soil surface and only one fire had all plots reach this (a spring Central Grasslands fire).
-Only 10 fires reached 325$^\circ$C on the soil surface, with most of these fires reaching this point in only one plot and never in more than half of the plots.
+Among plots that burned, fewer than half exceeded 100$^\circ$C at the soil surface (49 of 121 plots).
+A majority of the fires (18) had at least one plot exceed 100$^\circ$C at the soil surface, seven had over half of the plots exceed 100$^\circ$C at the soil surface, and only one fire had all plots reach this (a spring Central Grasslands fire).
+Only 10 fires reached or exceeded 325$^\circ$C on the soil surface, with most of these fires reaching this point in only one plot and never in more than half of the plots.
 
@@
 Fires spread faster with higher wind speeds (t = 2.92, P < 0.01), but no other variable had a statistically-significant association with rate of spread (Table 1). 
-Aboveground temperatures increased as fuel load increased (t = 2.82, P = 0.01) and decreased as fuel moisture increased (t = -2.16, P = 0.04). 
+Aboveground flame temperatures increased as fuel load increased (t = 2.82, P = 0.01) and decreased as fuel moisture increased (t = -2.16, P = 0.04). 
 No fuel or weather variable included here had statistically-significant associations with soil surface temperature (Table 1).
@@
 But @archibold2003 also reported substantially lower fuel moisture in each season and they had approximately three times the fuel load, likely due to an absence of grazing management on the remnant prairie.
-A previous study reported similar results---temperatures approaching 500$^\circ$C when fuel loads averaged 2.8-4.5 t$\cdot$ha$^{\text{-}1}$ [@archibold1998]. 
+A previous study reported similar results---temperatures approaching 500$^\circ$C when fuel loads averaged 2.8-4.5 t ha$^{\text{-}1}$ [@archibold1998]. 
 With greater variability in fuel load and fuel moisture, we might also expect to see these factors have greater influence on aboveground flame temperatures. 
-For example, in Colorado, @augustine2014 observed a strong linear relationship between fuel load and  temperatures ranging from 60-200$^\circ$C, but their fuel load also ranged from 0.2 to 1.2 t$\cdot$ha$^{\text{-}1}$.
+For example, in Colorado, @augustine2014 observed a strong linear relationship between fuel load and  temperatures ranging from 60-200$^\circ$C, but their fuel load also ranged from 0.2 to 1.2 t ha$^{\text{-}1}$.
 
-Although one must look beyond the northern Great Plains for comparable data on fire spread, our results are consistent with international reports. 
-For example, our data match the pattern from Australian grassland, where @cheney1993 found a statistical relationship between rate of spread and both wind speed and fuel moisture, but not total fuel load.
-Their fires traveled at rates similar to the vast majority of ours (1-2 m$\cdot$s$^{\text{-}1}$), with slightly higher fuel loads and slightly lower relative humidity. 
-Wind was by far their most important variable, followed by dead fuel moisture, consistent with patterns in our data. 
-Likewise, @cruz2015 observed fire spread rates up to 2 m$\cdot$s$^{\text{-}1}$ in fully-cured grassland fuels with fuel loads at the upper end of our samples.  
-In comparing our results to these studies, it appears the trigonometric method of calculating rate of spread through the thermocouple array [@simard1984] overestimates rate of spread, but there is no indication that using this measure of spread as a scaled response variable in regression models presents error in interpreting the relative effect of fuel and weather predictors.
+Fires at both of our locations spread much more slowly than most reports from other grasslands, due primarily to high fuelbed moisture content and little opportunity for mitigation by wind or dry air (Fig.
+1).
+Perhaps the most variability in fire spread was reported by @sneeuwjagt1977 from prescribed grass fires in California and Washington, where rates of spread ranged from 0.2-61 m min$^{\text{-}1}$.
+From grassland fires in South Africa and Kansas, USA, @trollope2002 reported average spread rates of 24 and 32 m min$^{\text{-}1}$, respectively, for head fires and 0.12 and 0.14 m min$^{\text{-}1}$, respectively, for back fires.
+In a tallgrass prairie in Texas, @clements2019 recorded fire spreading between 72-150 m min$^{\text{-}1}$ with the wind and 48 m min$^{\text{-}1}$ for flanking fires.
+Likewise, fires through cured grass fuelbeds in Australia spread much more rapidly than we observed---up to 18-180 m min$^{\text{-}1}$ [@cheney1993; @cheney1995; @cruz2015]. 
+Through partially-cured stands, though, @cruz2015 found spread rates dropped to 3-44 m min$^{\text{-}1}$, approaching our location averages of 2.1 and 3.1 m min$^{\text{-}1}$. 
+Consistent with our finding that only wind speed had a statistically-significant effect on increasing rate of spread (Fig.
+3), @cheney1993 found that wind was by far the most important variable to spread rate.
 
@@
 
-The functional difference between head and back fires in the fuelbeds reported here are likely. 
+The functional difference between head and back fires in the fuelbeds reported here is likely moot.
 Because our fuels were often sparse, sometimes only marginally cured, and prescriptions precluded taking advantage of higher wind or lower relative humidity to mitigate fuel limitations, we often employed substantial interior ignitions using strip, point, flanking, and spiral patterns that sent flame fronts towards our sensors in all possible directions at different times.
@@
 With this in mind, @cruz2015 employed a hybrid approach in which fuel moisture was measured for the various components, from which a weighted average was used as a predictor variable in regression analysis.
+They subsequently found that overall degree of curing, not simply live fuel moisture, was the most important variable in explaining the dampening effect of fuel moisture on fire behavior.
 
@@
 Research conducted at the spatial scales at which management occurs helps managers trust the transfer of knowledge from studies to working landscapes [@sayre2005; @cacciapaglia2012].
-Critically, all of our sensors were greater than 50 m from the initial fire line, identified in simulations as a
+For example, all of our fire behavior measurements were made more than 50 m from the initial fire line, the distance identified in simulations and used in wildland fire science to allow flame fronts to achieve a quasi steady state in spread rate [@fernandes2000; @sutherland2020], which is obviously precluded in studies that employ small plots.
 

### 2022-01-27 paper/SupplementaryInfo.Rmd
First draft of first revision

@@
   - \usepackage{hyperref}
+  - \usepackage{booktabs}
 output:
@@
 
-```{r map, eval = TRUE, out.width='100%', fig.align='center', fig.cap='Main map: Study locations (triangles) within two EPA level 3 ecoregions in North Dakota.
+```{r eval = TRUE, out.width='100%', fig.align='center', fig.cap='Main map: Study locations (triangles) within two EPA level 3 ecoregions in North Dakota.
 Inset: State of North Dakota (dark blue) within the Great Plains (light blue) with respect to the continental United States.
@@
 \end{equation}
+
+# Additional results
+
+## Fuel, weather, and fire behavior summaries
+
+```{r results='asis', eval = TRUE}
+pacman::p_load(tidyverse)
+    read_csv(url('https://raw.githubusercontent.com/devanmcg/SpatialFireBehavior/main/data/AnalysisDataKY.csv')) %>% 
+    select(-LAI, -JD, -MaxWindSpeed) %>%
+    mutate(ros = ifelse(ros >= 10, NA, ros), 
+           tHa = ifelse(tHa >=4, NA, tHa)) %>%
+    # convert units 
+    mutate(WindSpeed = WindSpeed * 0.28,  # km/hr to m/s
+           tHa = tHa * 0.1) %>%           # t/ha to kg/m^s
+    pivot_longer(cols = AirTemp:tHa, 
+                 names_to = 'var', 
+                 values_to = 'value') %>%
+    mutate(type = case_when(
+                    var %in% c('MaxC', 
+                               'SoilMaxC', 
+                               'ros') ~ 'Fire behavior', 
+                    var %in% c('tHa',
+                               'FuelMoisture') ~ 'Fuel', 
+                    TRUE ~ 'Weather' ), 
+           type = factor(type, levels = c('Fuel', 'Weather', 'Fire behavior')),
+           var = recode(var, 
+                        AirTemp = 'Air temperature ($^\\circ$C)', 
+                        RH = 'Relative humidity (\\%)', 
+                        dpC = 'Dew point ($^\\circ$C)', 
+                        FuelMoisture = 'Total fuel moisture (\\%)', 
+                        tHa = 'Fuel load (kg m$^{-2}$)',
+                        MaxC = 'Flame temp ($^\\circ$C)', 
+                        WindSpeed = 'Wind speed (m s$^{-1}$)', 
+                        ros = 'Rate of spread (m min$^{-1}$)', 
+                        VPD = 'Vapor pressure deficit',
+                        SoilMaxC = 'Soil surface temp ($^\\circ$C)'), 
+           location = recode(location, 
+                             H = "Hettinger", 
+                             CG = "Central Grasslands")) %>%
+  group_by(location, type, var) %>%
+    summarize(Mean = round(mean(value, na.rm = T), 1), 
+              SD = round(sd(value, na.rm = T), 2),
+              .groups = 'drop') %>%
+  mutate(value = paste0(Mean, ' ± ', SD)) %>%
+  rename(Variable = var) %>%
+      select(-Mean, -SD, -type) %>%
+  pivot_wider(names_from = location, 
+              values_from = value) %>%
+      xtable::xtable(caption = "Summary of fuel, weather, and fire behavior data collected from 25 fires in two locations in North Dakota.
+Fires in Hettinger were conducted in autumn, while those at Central Grasslands were conducted in spring.") %>%
+      print(include.rownames=FALSE, 
+            comment=FALSE, 
+            sanitize.text.function=function(x){x})
+```
 
@@
            position = recode(position, "1"="a", "2"="b", "3"="c"), 
-           ArrivalTime = as.numeric(ArrivalTime) /60 ) %>%
+           ArrivalTime = as.numeric(ArrivalTime) /60 ) %>% # converts to m/min!
     spread(position, ArrivalTime)  %>%

### 2022-01-31 paper/FigureGenerator_r1.Rmd
create structured abstract

@@
 Horizontal gray lines denote 25%, 50% (median) and 75% quantiles; triangles are arithmetic means.
-Means and standard deviation are also reported in Supplemental Information Table 1.")
+Means and standard deviation are also reported in Supplemental Information Table 1.
+VPD = Vapor pressure deficit.")
 ```
@@
 `r f_nums('violin_plots')`
-```{r data_summary_gg, fig.cap = violin_cap, fig.height=8, fig.width=6}
+```{r data_summary_gg, fig.cap = violin_cap, fig.height=8.25, fig.width=6}
 LongData <- 

### 2022-08-24 paper/springer/ReviewResponse/ReviewResponse1.tex
start Env Man R1

\documentclass[parskip=half]{scrartcl}
\usepackage{booktabs}
\usepackage[margin=2.5cm]{geometry}
\usepackage[dvipsnames]{xcolor}

\title{\Large Weather and fuel as modulators of grassland fire behavior in the northern Great Plains}
\subtitle{Response to Review}
\author{ }
\date{}

% \pagenumbering{gobble}
\raggedright

\newcommand{\AR}[1]
	{\color{PineGreen}AR: #1\color{black} \par }

\begin{document}

\maketitle

\vspace{-2cm}

\AR{We appreciate the opportunity to revise our submission to \emph{Environmental Management}.

Major points of revision include: 
\begin{itemize}
\item   
\item 
\item

\end{itemize}

We trust the editor will find these revisions address the important reviewer concerns. 
Author responses (AR:) to each reviewer comment follow inline below, as colored text. }

\section*{Associate Editor}

The topic of this manuscript falls within the scope of the journal. Two reviewers have evaluated the manuscript - although they see potentially interesting results, they have also identified critical issues that need to be addressed. These issues are related to the limited availability of soil data, which is required to better understand the soil processes regulating the observed treatment effects, and to the excessive speculation, leading to claims not supported by the data. In light of their comments, which I support, I recommend major revisions.

\AR{We appreciate the review and thoughtful considerations from both editor and reviewers.}

\section*{Reviewer 1}

The manuscript addresses fire behaviour in prescribed burns carried out in the northern Great Plains of the US. The authors concentrated their efforts on rate of spread and fire temperatures as measured by thermocouples and they analysed the variation as a function of weather-related variables. The work suffers from a number of shortcomings:
-	The futile measurement of temperatures, which are not really able to inform about fire behaviour or fire effects and thus do not have  correspondence with standard fire behavior characteristics;
-	Flame characteristics, which are valuable from the point of view of managers, were not monitored. Likewise, fireline intensity was not calculated (but it seems it could easily be derived, as the authors have RoS and fuel load – in this system nearly all of the elevated biomass is expected to burn, even if it’s not totally cured). However, FLI estimates would suffer from the same problems as RoS, see below;
-	Fuel moisture and fuel load do not distinguish between components;
-	Calculated RoS deviates from the standard, i.e. distinction between forward and backward spread rate, as it integrates variation in wind direction and, apparently, interacting flame fronts because of combined ignition patterns;
-	Fire behavior analyses are based on weather collected far from the fires

It’s interesting that the authors seem well aware of all these deficiencies. Were they aware of them before planning the work?  I understand that the study reflects a management setting. However, with all the above issues, I don’t see how it can contribute to improved prescribed burning management, let alone increased fire behaviour understanding in this system.

\AR{We appreciate the reviewer's attention to detail and clear understanding of the wildland fire environment, and the challenges in properly characterizing it.  
The reviewer is correct in observing that we are well aware of the advantages and disadvantages of various approaches to measuring fire behavior. }

Specific comments

L8. Flame size is also very often used by fire scientists.

L13-L20. Important to mention that moisture-related atmospheric variables are important because they control dead fuel moisture content.

L22. To be accurate, energy release rate not always increase with available fuel. What always increase is the quantity of released energy.

L25. Although temperature varies within flame, all flames and all fires have the same temperature, from the tiniest candle flame to the highest-intensity crown fire. So, what potentially increases with time since fire is flame size and fireline intensity.

L30. Not sure what is meant with “outside the highly-cured context”. The effect of fuel moisture content is relevant throughout the entire range of other variables, e.g. wind speed.

L37. For decades fire ecologists have been wrong in this endeavour, and it was not for lack of warning, see for e.g. Van Wagner \& Methven in the 1970s. The value of temperatures measured during fires is very limited in relation to fire behavior or effects.

L79. It’s hard to reconcile the objectives of the study with the long (but not exhaustive) previous text explaining the problems with measuring ``fire temperatures''. Instead of temperatures why did the authors not measured flame length or height or calculated fireline intensity?

L162. Please provide information regarding the capability of the thermocouples to detect peak temperatures, namely the type of thermocouple and its diameter.

L191. So, fire weather records were systematically taken on-site or not? If they were, this should be described, and the results presented – weather stations are too far away to be useful to describe wind conditions during the fires.

L246-248. These results, like in many other studies, are very indicative of the problems associated with flame temperature measurement. Most of the values measured and reported are below the temperature of flaming combustion!

L259. The lack of association between RoS and weather does not surprise me, given that it was collected far away. Supplement Table 1 should be included with the main text and plots of RoS versus wind speed and other weather variables (and fuel moisture) could be included as supplementary material.

Fig. 3 and Table 1 should not appear before Discussion.

L274-277. This would be expected because fire spread in the elevated fuels and residual flaming or smouldering combustion in the ground is not expected owing to the nature of the fuel bed.

L360. This is clearly a relevant limitation of the RoS methodology and one that precludes characterizing the fires in terms of the benchmark (headfire RoS) or comparison with most experiments.

L384. So, why was not this done, as it is standard practice? And why was curing not assessed?

\section*{Reviewer 2}

This manuscript describes the relationship between fuel bed, weather, and three fire behavior metrics in low-intensity prescribed fires in two sites in North Dakota. The manuscript is generally clearly written, though I provide many suggestions for improvement below.

INTRODUCTION
First paragraph would benefit from clear statement of link between ``prescribed fire'' and ``wildland fire''. Also, the introduction reads like a textbook until the second-to-last paragraph. State in the first paragraph where you’re going with all this terminology – lack of information for the Northern Great Plains (NGP) and why it’s needed. Although the title and abstract indicate that this paper is about grasslands, some terms used in the Introduction would make more sense if its stated up front that all this information is about grassland wildland fire.

24: Greater fuel load is also attributable to vegetation composition, productivity of system (soil fertility, climate, etc.), time since/intensity of grazing, and a whole lot of other stuff.

30-32:  ``highly-cured context of wildfire seasons''?  See first comment – explanation of ``curing'' might be useful for some readers.

63-66:  Does rate of spread measure intensity or energy flux? If not, how does it make moot the third issue of temperature as a fire behavior metric? If so, make this connection clearer. This is critical given the emphasis on rate of spread in this manuscript.

76-78:  Do the papers cited in this paragraph test the effect of fuel beds on fire behavior in the NGP? Mention of that would be nice here, given the emphasis on fuel in the title, abstract, and Introduction.

METHODS

Consistency is needed in this section and throughout the manuscript in terms used for different scales of measurement. This becomes important in the results when it’s unclear what scale is being referenced. For example, in the first paragraph of the ``Data collection'' subsection, the 1-m triangles seem to be referenced as both ``plots'' and ``microplots''. Since all the data analysis and results seem to be using the 1-m data, stick with just one term. Unless there’s something different between plots and microplots.
163: Why 15 cm? What is the biological or physical significance of that height?
184, Figure 1, and elsewhere: Units need to be provided for VPD. 
214: Here or in the description of sites/fires, state how many years over which data were collected.
222-223: It would be nice for the reader to briefly remind them here exactly what the fuel and weather variables described above are.
227-229: The data (Figure 1) suggest that there was a greater difference in VPD than RH between the two sites. Justify (physically, biologically) using RH rather than VPD. 
RESULTS
Figure 1 caption: Add detail that Hettinger fires occurred in fall and Central grasslands fires occurred in spring.
Figure 2 caption:  Suggest changing to ``response variables in boxes'' (My eye barely registers the blue text compared to the light blue data points indicating CG vs. H). Caption also needs to explain what RH, DP, and VPD are.
235-237: Are you indicating ``drier air'' by higher RH? How is it different from ``more evaporative'' conditions? Please clarify in this paragraph and/or in methods when describing these weather variables.
246: Figure 1 indicates that only 25\% of the fires exceeded 325 degrees C. Clarify in Figure 1 caption the scale of the data presented are from, and clarify here what scale you are talking about when referencing ``fires''.
Table 1: Formatting went awry in t/df column.
DISCUSSION
288: ``will require collecting''
293-294: Confusing wording: Suggest changing to ``...surface temperatures were 110 deg C and 165 deg C for backfires and headfires, respectively...''
312-313, 318: Is this second Archibold study also 15-cm temperature, or surface? Also, use the same units for fuel load as your data (kg/m2) to make comparison to your results easier.
358-359: Figure 1 shows ~25\% of observations <60 deg C and <50\% <100 deg C. If I’m interpreting Figure 1 incorrectly, the caption of Figure 1 needs modifying to make interpretation easier.

\end{document}

### 2022-08-24 paper/FigureGenerator_r1.Rmd
start Env Man R1

@@
 output:
-  #html_document
-  word_document:
-    reference_docx: word_template.docx
+  pdf_document:
+     keep_tex: yes
 ---
@@
 knitr::opts_chunk$set(message = FALSE, warning=FALSE, 
-                      echo=FALSE, eval=TRUE)
+                      echo=FALSE, eval=TRUE, fig.path = './springer/', dev = c('png', 'pdf'))
 
@@
 
-<!-- Table 1: Results of generalized linear mixed effect regression models testing three measure of fire behavior against four potential predictor variables.
+Table 1: Results of generalized linear mixed effect regression models testing three measure of fire behavior against four potential predictor variables.
 Statistics reflect pooled results of 50 imputed datasets using the *mice* package in R; see Methods.
 Vapor pressure deficit included for Rate of spread only due to statistically-significant difference between GLMM regression results that included VPD compared to RH alone (Wald = 5.32, P = 0.02), while temperature models had no such difference.
--->
 
-<!-- |Response           |Model term             |       t df  |      P| -->
-<!-- |:------------------|:----------------------|------------:|------:| -->
-<!-- |Rate of spread     |                       |             |       | -->
-<!-- |                   |Wind speed             |  2.92 108.0 | < 0.01| -->
-<!-- |                   |Vapor pressure deficit | -2.31 108.1 |   0.02| -->
-<!-- |                   |Relative humidity      | -1.66 80.8  |   0.10| -->
-<!-- |                   |Fuel load              |  1.16 49.9  |   0.25| -->
-<!-- |                   |Fuel moisture          | -0.68 62.9  |   0.50| -->
-<!-- |Canopy temperature |                       |             |       | -->
-<!-- |                   |Fuel load              |  2.82 54.2  |   0.01| -->
-<!-- |                   |Fuel moisture          | -2.16 40.9  |   0.04| -->
-<!-- |                   |Relative humidity      | -1.19 120.4 |   0.24| -->
-<!-- |                   |Wind speed             |  0.02 132.4 |   0.99| -->
-<!-- |Surface temperature|                       |             |       | -->
-<!-- |                   |Relative humidity      | -1.19 74.8  |   0.24| -->
-<!-- |                   |Fuel load              | -0.48 20.1  |   0.64| -->
-<!-- |                   |Fuel moisture          | -0.47 40.1  |   0.64| -->
-<!-- |                   |Wind speed             |  0.06 49.5  |   0.95| -->
+|Response           |Model term             |       t df  |      P|
+|:------------------|:----------------------|------------:|------:|
+|Rate of spread     |                       |             |       |
+|                   |Wind speed             |  2.92 108.0 | < 0.01|
+|                   |Vapor pressure deficit | -2.31 108.1 |   0.02|
+|                   |Relative humidity      | -1.66 80.8  |   0.10|
+|                   |Fuel load              |  1.16 49.9  |   0.25|
+|                   |Fuel moisture          | -0.68 62.9  |   0.50|
+|Canopy temperature |                       |             |       |
+|                   |Fuel load              |  2.82 54.2  |   0.01|
+|                   |Fuel moisture          | -2.16 40.9  |   0.04|
+|                   |Relative humidity      | -1.19 120.4 |   0.24|
+|                   |Wind speed             |  0.02 132.4 |   0.99|
+|Surface temperature|                       |             |       |
+|                   |Relative humidity      | -1.19 74.8  |   0.24|
+|                   |Fuel load              | -0.48 20.1  |   0.64|
+|                   |Fuel moisture          | -0.47 40.1  |   0.64|
+|                   |Wind speed             |  0.06 49.5  |   0.95|
 
@@
            var = recode(var, 
-                        AirTemp = 'Air~temperature~(degree*C)', 
-                        RH = 'Relative~humidity~(\'%\')', 
-                        dpC = 'Dew~point~(degree*C)', 
-                        FuelMoisture = 'Total~fuel~moisture~(\'%\')', 
-                        tHa = 'Fuel~load~(kg~m^-2)',
-                        MaxC = 'Flame~temp~(degree*C)', 
-                        WindSpeed = 'Wind~speed~(m~s^-1)', 
-                        ros = 'Spread~rate~(m~min^-1)', 
-                        SoilMaxC = 'Soil~surface~temp~(degree*C)'), 
+                        AirTemp = 'Air temperature (°C)', 
+                        RH = 'Relative humidity (%)', 
+                        dpC = 'Dew point (°C)', 
+                        FuelMoisture = 'Total fuel moisture (%)', 
+                        tHa = 'Fuel load (kg/sq m)',
+                        MaxC = 'Flame temp (°C)', 
+                        WindSpeed = 'Wind speed (m/s)', 
+                        ros = 'Spread rate (m/min)', 
+                        SoilMaxC = 'Soil surface temp (°C)'), 
            location = recode(location, 
@@
        title = "Weather") + 
-  facet_wrap( ~ var, scales = 'free', labeller=label_parsed) +
+  facet_wrap( ~ var, scales = 'free') +
   scale_fill_manual(name = "Location", 
@@
        title = "Fuel") + 
-  facet_wrap( ~ var, scales = 'free', labeller=label_parsed) +
+  facet_wrap( ~ var, scales = 'free') +
   scale_fill_manual(name = "Location", 
@@
          title = "Fire behavior") + 
-    facet_wrap( ~ var, scales = 'free', labeller=label_parsed) +
+    facet_wrap( ~ var, scales = 'free') +
     scale_fill_manual(name = "Location", 

### 2022-08-24 paper/FigureGenerator_r1.tex
start Env Man R1

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\author{}
\date{\vspace{-2.5em}}

\begin{document}

Table 1: Results of generalized linear mixed effect regression models
testing three measure of fire behavior against four potential predictor
variables. Statistics reflect pooled results of 50 imputed datasets
using the \emph{mice} package in R; see Methods. Vapor pressure deficit
included for Rate of spread only due to statistically-significant
difference between GLMM regression results that included VPD compared to
RH alone (Wald = 5.32, P = 0.02), while temperature models had no such
difference.

\begin{longtable}[]{@{}llrr@{}}
\toprule
Response & Model term & t df & P \\
\midrule
\endhead
Rate of spread & & & \\
& Wind speed & 2.92 108.0 & \textless{} 0.01 \\
& Vapor pressure deficit & -2.31 108.1 & 0.02 \\
& Relative humidity & -1.66 80.8 & 0.10 \\
& Fuel load & 1.16 49.9 & 0.25 \\
& Fuel moisture & -0.68 62.9 & 0.50 \\
Canopy temperature & & & \\
& Fuel load & 2.82 54.2 & 0.01 \\
& Fuel moisture & -2.16 40.9 & 0.04 \\
& Relative humidity & -1.19 120.4 & 0.24 \\
& Wind speed & 0.02 132.4 & 0.99 \\
Surface temperature & & & \\
& Relative humidity & -1.19 74.8 & 0.24 \\
& Fuel load & -0.48 20.1 & 0.64 \\
& Fuel moisture & -0.47 40.1 & 0.64 \\
& Wind speed & 0.06 49.5 & 0.95 \\
\bottomrule
\end{longtable}

Figure 1: Distribution of weather, fuel, and fire behavior data for
fires in southwestern North Dakota (Hettinger, dark maroon) and central
North Dakota (Central Grasslands, light blue). Horizontal gray lines
denote 25\%, 50\% (median) and 75\% quantiles; triangles are arithmetic
means. Means and standard deviation are also reported in Supplemental
Information Table 1. VPD = Vapor pressure deficit.
\includegraphics{./springer/data_summary_gg-1.png}

\newpage

Figure 2: Principal Components Analysis of fire behavior data (response
variables in blue; rate of spread (ROS), temperature above surface
(flame ºC), and temperature at soil surface (soil ºC) for prescribed
burns on rangeland at Hettinger (H), in southwestern North Dakota, and
Central Grasslands (CG), in central North Dakota. No difference between
locations (P = 0.11). Total variance explained in these two axes = 86\%.
\includegraphics{./springer/pca_gg-1.png}

\newpage

Figure 3: Regression coefficients and 95\% confidence intervals for fuel
and weather terms from models for maximum temperature at 15 cm above the
soil surface (flame), maximum temperature at the soil surface (soil),
and rate of spread. \includegraphics{./springer/CI_gg-1.png}

\end{document}

### 2022-08-24 paper/GrasslandFireBehaviorNGP.Rmd
start Env Man R1

@@
 ---
-title: "Weather and fuel as modulators of grassland fire behavior in the northern Great Plains"
-author: 'Devan Allen McGranahan^A,C^, Megan E.
-Zopfi^B^, Kathryn A.
-Yurkonis^B^ ^A^ USDA Agricultural Research Service, Livestock and Range Research Laboratory, Miles City, Montana, USA ^B^ Department of Biology, University of North Dakota, Grand Forks, North Dakota, USA ^C^ Corresponding author.
-Email: Devan.McGranahan@usda.gov'
+title: Weather and fuel as modulators of grassland fire behavior in the northern Great
+  Plains
+author: 'Devan Allen McGranahan^A,C^, Megan E. Zopfi^B^, Kathryn A. Yurkonis^B^ ^A^
+  USDA Agricultural Research Service, Livestock and Range Research Laboratory, Miles
+  City, Montana, USA ^B^ Department of Biology, University of North Dakota, Grand
+  Forks, North Dakota, USA ^C^ Corresponding author. Email: Devan.McGranahan@usda.gov'
 output:
+  pdf_document: 
+     citation_package: natbib
+     keep_tex: yes
   word_document:
@@
 
+**Background:** 
 Fuel and weather interact to affect wildland fire behavior, but little is known about associations between these variables in the northern Great Plains of North America. 
-Few studies consider rate of spread or statistically test the influence of fire environmental variables.
-We measured fuel load and moisture ahead of prescribed fires in North Dakota, USA, and used a multi-channel thermocouple array to measure rate of spread, soil surface temperature, and flame temperature, which we compared with fire weather data.
-Aboveground flame temperatures averaged 225$^\circ$C during spring burns in and 250$^\circ$C during fall burns.
-Surface temperatures averaged just above 100$^\circ$C, although 50% of observations were 60$^\circ$C or less.
-Fires spread at an average of 2.5 m min$^{\text{-}1}$. 
-Regression analysis indicated that wind speed drove faster rates of spread while higher fuel loads and lower overall fuelbed moisture produced higher flame temperatures.
-None of our measurements explained variability in soil surface temperature, likely because ground-level heating remained low.
-We highlight differential responses among fire behavior metrics to different components of the wildland fire environment.
-Results help wildland fire managers match burn conditions to desired outcomes, and fire ecologists match measurements to specific ecological responses.
-We also discuss remaining knowledge gaps in how management and ignition patterns interact to create variability in fire behavior.
+Few studies consider rate of spread or statistically test the influence of fuel and weather.
+
+**Methods:** 
+We measured overall fuel load and moisture ahead of prescribed fires in North Dakota, USA, and used a thermocouple array to measure rate of spread, soil surface temperature, and aboveground flame temperature, to compare with fire weather data.
+
+**Key results:** 
+Flame temperatures averaged 225$^\circ$C during spring burns and 250$^\circ$C during fall burns, and were generally higher with greater fuel loads and lower overall fuelbed moisture. 
+Surface temperatures averaged ~100$^\circ$C, although 50% of observations were ≤60$^\circ$C. 
+Fires spread at an average of 2.5 m min$^{\text{-}1}$, increasing with wind speed.
+
+**Conclusions:** 
+Prescribed fire in northern Great Plains working rangeland spreads slowly and effects low soil surface temperatures, often limited by high fuelbed moisture. 
+Fire behavior measurements respond differently to variability in fuel and weather. Belowground heating is likely minimal.
+
+**Implications:** 
+Ecologists, consider which fire behavior measurements best relate to fire effects. 
+Managers, consider weather and ignition pattern mitigations when fuels constrain desired fire behavior to ensure effective burns.
 
@@
 
-More than simply the result of combustion of vegetation, fire behavior in wildland environments is multi-faceted, with different components producing different effects on the surrounding environment and organisms within.
+More than simply the result of combustion of vegetation, wildland fire behavior is multi-faceted, with different components producing different effects on the surrounding environment and organisms within.
 Most wildland fire scientists describe fire behavior in terms of *rate of spread*—how quickly a flame front moves through a fuelbed—and *intensity*—a suite of measurements of how much energy is released by combustion, often expressed as a rate of energy release over time [@mcgranahan2021a].
 
-Wildland fire behavior is controlled by interactions between several abiotic and biotic factors, and understanding them is critical to safe and effective wildland fire management [@benson2009]. 
+Wildland fire behavior is controlled by interactions among several abiotic and biotic factors, and understanding them is critical to safe and effective wildland fire management [@benson2009]. 
 *Abiotic factors* include those determined by the physical environment, such as wind speed and atmospheric moisture content. 
@@
 *Biotic factors* relate principally to the amount and nature of plant biomass available for combustion. 
-Overall, energy release rates increase as more fuel is available to burn; the structure and arrangement of vegetation is also important. 
+Overall, energy release rates increase as more fuel is available to burn. 
+The structure and arrangement of vegetation is also important. 
 Greater fuel load attributable to longer time-since-fire increases fire temperature, and spatial variability in fuel load and patchy distribution of fine fuels in turn drive variability in fire behavior [@patten1984; @gibson1990; @gomes2020].
@@
 
-Our objectives were to (1) describe the range of variability in three measures of fire behavior—rate of spread, soil surface temperature, and flame temperature 15 cm above soil surface—during prescribed burns in typical fuelbeds of the northern US Great Plains, and (2) explain variability in fire behavior in terms of abiotic and biotic conditions.
-Our analysis emphasizes the differential effects of environmental variables among the three responses, and the multidimensional relationship among these different measures of grassland fire behavior during prescribed burns conducted at scales consistent with land management in the region.
+Our objectives were to (1) describe the range of variability in three measures of fire behavior—rate of spread, soil surface temperature, and flame temperature 15 cm above soil surface—during prescribed burns in typical fuelbeds of the northern US Great Plains, and (2) explain variability in fire behavior in terms of abiotic and biotic factors.
+Our analysis emphasizes the differential effects of environmental variables among the three measures of fire behavior, and the multidimensional relationship among these responses during prescribed burns conducted at scales consistent with land management in the region.
 
@@
 
-We measured each fire with a set of 9, 1 m equilateral triangle plots arranged in a nested fashion such that three 10 m triangles, each containing 3, 1m plots, were placed 100 m apart to form a total plot area of 0.433 ha with 27 sample points positioned at the centroid of each burn unit (A detailed schematic of this layout is presented in Supplemental Information Figure 2).
+We measured each fire with a set of 9, 1-m equilateral triangle plots arranged in a nested fashion such that three 10 m triangles, each containing 3, 1-m plots, were placed 100 m apart to form a total plot area of 0.433 ha with 27 sample points positioned at the centroid of each burn unit (a schematic of this layout is presented in Supplemental Information Figure 2).
 This fractal design is modified from the Sierpinksi triangle described by @dorrough2007 and applied to measuring wildland fire spread by @mcgranahan2021b.
-Although the logarithmically-scaled nested design was intended for geospatial analysis of point-level data, for our analyses here we calculate averages from the finest (1 m) scale, consistent with the method of locating multiple microplots within larger burned areas to characterize spatial variability within fires [@fernandes2000].
+Although the logarithmically-scaled nested design was intended for geospatial analysis of point-level data, for our analyses here we calculate averages from the finest (1 m microplot) scale, consistent with the method of locating multiple microplots within larger burned areas to characterize spatial variability within fires [@fernandes2000].
 
@@
 
-**Fire behavior data** were recorded as temperature ($^\circ$C) associated with the advancing flame front at each of the 27 points arranged in 9, 1 m triangles at the center of each burn unit.
+**Fire behavior data** were recorded as temperature ($^\circ$C) associated with the advancing flame front at each of the 27 points arranged in 9, 1-m triangular microplots at the center of each burn unit.
 Data were recorded with the open-source FeatherFlame thermocouple datalogger system [@mcgranahan2021b], logging at 1.5 Hz.
-Briefly, the FeatherFlame system reads overbraided, ceramic fiber-insulated, 20-gauge K-type thermocouples (Omega, Norwalk, CT) connected to an Arduino-based datalogger assembled from Adafruit Feather breakout boards (M0 Adalogger, datalogging shield, and OLED display; Adafruit Industries, LLC, New York City, NY) and housed inside water-resistant Pelican cases (Pelican Products, Inc, Torrance, California).
+The FeatherFlame system reads overbraided, ceramic fiber-insulated, 20-gauge K-type thermocouples (Omega, Norwalk, CT) connected to an Arduino-based datalogger assembled from Adafruit Feather breakout boards (M0 Adalogger, datalogging shield, and OLED display; Adafruit Industries, LLC, New York City, NY) and housed inside water-resistant Pelican cases (Pelican Products, Inc, Torrance, California).
 The low cost of open-source systems make multiple units more affordable than proprietary data loggers with no sacrifice in data quality [@mcgranahan2021]. 
+More details on the datalogger system are available in Supplemental Information.
 
-At each 1-m triangle—the individual observational unit in the nested plot design—we used four thermocouples connected to a single FeatherFlame datalogger.
+At each 1 m triangular microplot—the observational unit in the nested plot design—we used four thermocouples connected to a single FeatherFlame datalogger.
 Three thermocouples measured flame temperature 15 cm above the soil surface at each 1 m vertex while a fourth thermocouple recorded soil surface temperature at a representative point within the 1 m array.
@@
 For each fire event, we determined the time and value of the maximum temperature ($^\circ$C) as the flame front encountered each thermocouple. 
-We calculated the maximum flame temperature ($^\circ$C) 15 cm above soil surface for each plot as the mean of the thermocouple readings from its vertices. 
-We calculated the rate of spread (m min$^{\text{-}1}$) of the flame front as it passed through each plot using the maximum temperature timestamps following equations from @simard1984, which are presented in full in Supplemental Information.
+We calculated the maximum flame temperature ($^\circ$C) 15 cm above soil surface for each microplot as the mean of the thermocouple readings from its 3 vertices. 
+We calculated the rate of spread (m min$^{\text{-}1}$) of the flame front as it passed through each microplot using the maximum temperature timestamps as arrival times following equations from @simard1984, which are presented in full in Supplemental Information.
 
@@
 
-Prior to analysis, we ensured statistical power across 167 observational units by using multiple imputation to interpolate missing datapoints, as missing field data occurred for three rate of spread samples (2% of total), 29 fuel load values (17%), and 46 soil surface temperature values (27%).
+Prior to analysis, we ensured statistical power across 167 observational units by using multiple imputation to interpolate missing datapoints, as missing field data occurred for three rate of spread samples (2% of total), 29 fuel load values (17%), and 46 soil surface temperature values (27%) due to logistical and time constraints during the operational burn periods.
 We used the multiple imputation method in the *mice* package [@vanbuuren2011] in the R statistical environment [@rct2020] to fill in these missing values. 
@@
 
-**Regression analysis.---**We assessed weather and fuel effects on three fire behavior response variables: Maximum flame temperature ($^\circ$C) 15 cm above the soil surface (mean of three thermocouples), maximum soil surface temperature ($^\circ$C; single thermocouple), and rate of spread (m min$^{\text{-}1}$) through each 1 m equilateral triangle plot.
+**Regression analysis.---**We assessed weather and fuel effects on three fire behavior response variables: Maximum flame temperature ($^\circ$C) 15 cm above the soil surface (mean of three thermocouples), maximum soil surface temperature ($^\circ$C; single thermocouple), and rate of spread (m min$^{\text{-}1}$) through each 1 m trianglular microplot.
 Because all three response variables were best modeled with a gamma distribution, we fit generalized linear mixed-effect regression models for each response with the 'glmer' function from the *lme4* package in R [@bates2015].
@@
 In central Alberta grassland, @bailey1980 observed that surface temperatures varied between 110$^\circ$C and 165$^\circ$C for backfires and headfires, respectively, and headfires averaged 200$^\circ$C 15 cm above the ground; temperatures generally tracked with fuel load.
-Surface fires through jack pine barrens in Ontario had a similar range as ours: 140-545$^\circ$C [@smith1966]. 
-In our study, average 15-cm temperatures were 225$^\circ$C during spring burns in central North Dakota and 250$^\circ$C during fall burns in southwestern North Dakota; surface temperatures at both locations generally averaged just above 100$^\circ$C (Fig.
+Surface fires through jack pine barrens in Ontario had a similar range as ours [140-545$^\circ$C, @smith1966]. 
+In our study, mean 15-cm temperatures were 225$^\circ$C during spring burns in central North Dakota and 250$^\circ$C during fall burns in southwestern North Dakota; surface temperatures at both locations generally averaged just above 100$^\circ$C (Fig.
 1).
 
-Discrepancies between our data and others reported in the region are consistent with what would be expected when differences in the fire environment are considered.
-For example, @ohrtman2015 reported a wide range of maximum temperatures at the soil surface---150-500$^\circ$C---that was generally explained by variability in annual productivity and clipping frequency.
-Our fires were also cooler than those reported by @archibold2003 in Saskatchewan: using the mid-point of observations made at 10 cm and 20 cm as a comparison, spring fires reached 314$^\circ$C and fall fires reached 298$^\circ$C.
-But @archibold2003 also reported substantially lower fuel moisture in each season and they had approximately three times the fuel load, likely due to an absence of grazing management on the remnant prairie.
+Discrepancies between our data and others from the region are consistent with what would be expected when differences in the fire environment are considered.
+For example, @ohrtman2015 reported a wide range of maximum temperatures at the soil surface---150-500$^\circ$C---that was generally explained by variability in annual productivity and clipping frequency, which altered fuel load.
+Our fires were also cooler than those reported by @archibold2003 in Saskatchewan: using the mid-point of observations made at 10 cm and 20 cm as a comparison to our 15-cm values, spring fires reached 314$^\circ$C and fall fires reached 298$^\circ$C.
+But @archibold2003 also reported substantially lower fuel moisture in each season and they had approximately three times the fuel load, likely due to an absence of grazing on the remnant prairie. 
 A previous study reported similar results---temperatures approaching 500$^\circ$C when fuel loads averaged 2.8-4.5 t ha$^{\text{-}1}$ [@archibold1998]. 
@@
 
-Fires at both of our locations spread much more slowly than most reports from other grasslands, due primarily to high fuelbed moisture content and little opportunity for mitigation by wind or dry air (Fig.
+Fires at both of our locations spread much more slowly than most reports from other grasslands, due primarily to high fuelbed moisture content and little opportunity for mitigation by wind or lower atmospheric moisture (Fig.
 1).
@@
 Research conducted at the spatial scales at which management occurs helps managers trust the transfer of knowledge from studies to working landscapes [@sayre2005; @cacciapaglia2012].
-For example, all of our fire behavior measurements were made more than 50 m from the initial fire line, the distance identified in simulations and used in wildland fire science to allow flame fronts to achieve a quasi steady state in spread rate [@fernandes2000; @sutherland2020], which is obviously precluded in studies that employ small plots.
+For example, all of our fire behavior measurements were made more than 50 m from the initial fire line, the distance identified in simulations and used in wildland fire science to allow flame fronts to achieve a quasi-steady state in spread rate [@fernandes2000; @sutherland2020], which is obviously precluded in studies that employ small plots.
 

### 2022-08-24 paper/SupplementaryInfo.Rmd
start Env Man R1

@@
     highlight: tango
+    keep_tex: yes
 ---

### 2022-08-24 paper/springer/CoverLetter/CoverLetter_R1.tex
start Env Man R1

\documentclass[parskip=half, 
			   fontsize=11pt,
			   paper=a4]				
{scrartcl}
\input{SetupOptions} % contains scrletter

\newcommand{\journal}{\emph{Environmental Management}}

\vspace{-5em}

\begin{document}
	\begin{letter}{Editor\\
		\journal }
\setlength{\parindent}{10pt}

\opening{Dear Sir or Madam,}

Please consider the attached submission for publication in \journal.

This submission is motivated by the positive experience my colleague and I recently enjoyed publishing in \journal~(DOI \href{https://doi.org/10.1007/s00267-022-01659-y}{10.1007/s00267-022-01659-y}). 
I have decided to move on with this work after a lack of response from the Editor in Chief at the journal to which we initially submitted, following mishandling by an Associate Editor. 
After our initial review, the paper was understandably rejected and a resubmission requested, due to the extent of the reviewer comments. 
Each were productive, and we responded completely (these responses are attached). 
However, despite no reviewer criticism on the resubmitted manuscript, the (presumably new?) AE assigned to the resubmission rejected the manuscript on the grounds that our approach to measuring two-dimensional rate of spread violated assumptions that are in fact limited to one-dimensional rate of spread. 
But recently, Dr. Mark Finney\textemdash almost certainly the leading wildland fire scientist in the US, if not the world\textemdash led a textbook in which his team described the superiority of 2-D measurements over 1-D measurements. 
We brought this to the EiC's attention in our appeal of the AE decision, but have gone several weeks without a response.

As these data are of great interest to our stakeholders, we need to press on with publication. 
I think the scope and audience of \journal~is a good fit for us. 
We have explicitly referenced Finney et al's textbook into our Methods section to make this point clear. 
While I understand \journal~has no formal Portable Peer-Review process, I hope the editorial staff might take this context into account and consider an expedited review process.

We look forward to working with the staff of \journal~to ensure our manuscript meets the standards of the journal.

\vspace{-3em} 
\closing{} %eg. Regards
%\cc{another dude}
\end{letter}

\end{document}

### 2022-08-24 paper/springer/GrasslandFireBehaviorNGP_R1.tex
start Env Man R1

\documentclass[referee, 
		     sn-basic]{sn-jnl}

\usepackage{amssymb}
\usepackage{lineno}
\usepackage{booktabs}

\newcommand{\degC}{$^\circ$C}

% Springer stuff 
\jyear{2021}%
\raggedbottom
\unnumbered% uncomment this for unnumbered level heads

\begin{document}
\raggedright

\title[Weather, fuel, and grassland fire behavior]{Weather and fuel as modulators of grassland fire behavior in the northern Great Plains}

\author*[1]{\fnm{Devan Allen}  \sur{McGranahan}}\email{Devan.McGranahan@usda.gov}
\author[2]{\fnm{Megan E.} \sur{ Zopfi}}\email{Megan.Zopfi@und.edu }
\author[2]{\fnm{Kathryn A.} \sur{ Yurkonis}}\email{Kathryn.Yurkonis@und.edu}

\affil[1]{\orgname{USDA Agricultural Research Service}, 
	     \orgdiv{Livestock \& Range Research Laboratory},
              \orgaddress{\street{243 Ft. Keogh Rd.}, 
            		     \city{Miles City},
            		     \postcode{59301}, 
           		     \state{Montana},
            		     \country{USA}}}
\affil[2]{\orgname{University of North Dakota}, 
            \orgdiv{Department of Biology},
            \orgaddress{\street{10 Cornell St.}, 
            		   \city{Grand Forks},
            		   \postcode{58202}, 
           		   \state{North Dakota},
            		   \country{USA}}}

\abstract{Fuel and weather interact to affect wildland fire behavior, but little is known about associations between these variables
in the northern Great Plains of North America. Few studies consider rate of spread or statistically test the influence of fuel and weather.
We measured overall fuel load and moisture ahead of prescribed fires in North Dakota, USA, and used a thermocouple array to
measure two-dimensional rate of spread, soil surface temperature, and aboveground flame temperature, to compare with fire weather data.
Flame temperatures averaged 225\degC~during spring burns and 250\degC~during fall burns, and were generally higher with greater fuel loads and lower overall fuelbed moisture.
Surface temperatures averaged $\approx$100\degC, although 50\% of observations were $\leq$60\degC. 
Fires spread at an average of 2.5 m min\textsuperscript{-1}, increasing with wind speed.
As such, prescribed fire in northern Great Plains working rangeland appear to spread slowly and effect low soil surface temperatures, often limited by high fuelbed moisture. 
Fire behavior measurements respond differently to variability in fuel and weather. 
Belowground heating is likely minimal.
We suggest ecologists ought to consider which fire behavior measurements best relate to fire effects, and managers consider weather and ignition pattern mitigations when fuels constrain desired fire behavior to ensure effective burns.}

\keywords{Grassland fire ecology and management, Prescribed fire, Rangeland fire management, Robust wildland fire science, Wildland fire science in working landscapes}

\maketitle
\begin{linenumbers}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Prescribed fire is used widely for ecosystem management around the world \citep{weir2022}. 
More than simply the result of combustion of vegetation, wildland fire behavior is multi-faceted, with different components producing different effects on the surrounding environment and organisms within. 
Most wildland fire scientists describe fire behavior in terms of \emph{rate of spread}---how quickly a flame front moves through a fuelbed---and \emph{intensity}---a suite of measurements of how much energy is released by combustion, often expressed as a rate of energy release over time \citep{mcgranahan2021a}.

Wildland fire behavior is controlled by interactions among several
abiotic and biotic factors, and understanding them is critical to safe
and effective wildland fire management \citep{benson2009}. \emph{Abiotic
factors} include those determined by the physical environment, such as
wind speed and atmospheric moisture content. Wind speed has long been
recognized as a primary driver of fire behavior, especially in
well-cured grassland fuels
\citep{whittaker1961, cheney1995, kidnie2015}. Two measures of
atmospheric moisture content---relative humidity and vapor pressure
deficit---are also associated with fire growth
\citep{evett2008, reid2010, sedano2014}.

\emph{Biotic factors} relate principally to the amount and nature of
plant biomass available for combustion. Overall, energy release rates
increase as more fuel is available to burn. The structure and
arrangement of vegetation is also important. Greater fuel load
attributable to longer time-since-fire increases fire temperature, and
spatial variability in fuel load and patchy distribution of fine fuels
in turn drive variability in fire behavior
\citep{patten1984, gibson1990, gomes2020}. Furthermore, fine-leaved
grasses burn more completely and hotter than an equal mass of forbs
\citep{wragg2018}. Finally, fuel moisture content is an especially
important driver outside of the highly-cured context of wildfire seasons
\citep{sparling1966, kidnie2015}. Together, variability in flammability
traits and curing rates among species that comprise grassland fuelbeds
contributes to variability in fire behavior
\citep{cruz2015, kidnie2015, mcgranahan2016, cardoso2018}.

How and where within the wildland fire environment fire behavior
measurements are made matters a great deal to assessing fire effects.
For decades, fire ecologists have measured fire behavior as \emph{flame
temperature} via various methods, including arrays of
temperature-sensitive paints
\citep[e.g.,][]{whittaker1961, smith1966, bailey1980} or by recording
air temperature as a flame front passes over a thermocouple connected to
a datalogger \citep[e.g.,][]{strong2013, russell2015}.

Despite its popularity among fire ecologists, temperature alone is a
poor response variable fraught by several issues in collecting and
interpreting thermocouple data \citep[see review by][]{mcgranahan2020}.
Firstly, a considerable amount of variability in temperature is
attributable to sensor placement relative to both the ground and the
fire. There is neither a standard for placing thermocouple probes in the
wildland fire environment nor consistency in vertical temperature
profiles. Most observations of surface fire temperature profiles
describe an inverse, linear relationship between height and temperature
\citep{smith1966, patten1984, archibold2003}, although
\citet{ramsay1996} found the highest temperatures at the top of a 1 m
profile and the lowest temperatures at 30 cm, while \citet{frost1987}
and \citet{bailey1980} present evidence that the highest temperatures
occur midway up the profile. At least some of this variability might be
due to differences in surface vs.~flame temperature among head and back
fires \citep{trollope1978}.

Secondly, many factors that contribute to variability in temperatures
recorded by thermocouples are attributable to the nature of the sensor
rather than the nature of the fire \citep[e.g.,][]{walker1968}. Thus,
reporting `device temperatures' alone impedes comparisons between
studies; \citet{bova2008} present a standard calibration of thermocouple
probes, while \citet{mcgranahan2021b} simply uses the timestamps of peak
heating across an array of several thermocouples to calculate rate of
spread. Using rate of spread as the response variable makes moot the
third issue with temperature as a fire behavior metric: temperature of
the media around a probe is a poor proxy for the thermal experience of
an organism. Measures of intensity or energy flux are more biologically
relevant \citep{kremens2012, smith2016}.

In the North American Great Plains, most reports of grassland fire
behavior consist of temperatures derived from thermocouples, and there
are few data on rate of spread. Soil surface temperatures in South
Dakota tallgrass prairie ranged from 200-500\(^\circ\)C during spring
burns, and were greatest under lower fuel loads \citep{ohrtman2015};
fires in Saskatchewan mixed grass prairie exceeded 300\(^\circ\)C 5-10
cm above the soil surface in spring, summer, and fall
\citep{archibold2003}. Mean temperatures in experimental burns in
eastern Montana ranged from 172-222\(^\circ\)C in the summer to 253\(^\circ\)C in the spring \citep{strong2013, russell2015}. To our
knowledge, no studies on grassland fire in the northern Great Plains
region has explicitly tested the effect of fire weather on fire
behavior.

Our objectives were to (1) describe the range of variability in three
measures of fire behavior---rate of spread, soil surface temperature,
and flame temperature 15 cm above soil surface---during prescribed burns
in typical fuelbeds of the northern US Great Plains, and (2) explain
variability in fire behavior in terms of abiotic and biotic factors.
Our analysis emphasizes the differential effects of environmental variables among the three measures of fire behavior, and the multidimensional relationship among these responses during prescribed burns conducted at scales consistent with land management in the region.

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\hypertarget{study-locations}{%
\subsection{Study locations}\label{study-locations}}

We sampled 25 prescribed fires at two locations in central and
southwestern North Dakota, USA (Maps in Supplemental Information Figure
1). At both locations, sampled grasslands are included in a patch-burn
grazing study that requires a portion of each experimental unit to be
burned each year \citep{spiess2020}. The majority of the burn units were
16 ha, while a small set were 8 ha. Typical ignition patterns consisted
of downwind backing fires followed by either ring ignition and primarily
head fire spread, when fuels were conducive; when fuels were sparse or
higher-moisture, flanking fires and strip ignitions were employed as
necessary to ensure fire spread through the entire burn unit.

In central North Dakota, we sampled 15 spring (May) fires at the North
Dakota State University Central Grassland Research Extension Center near
Streeter, ND (46.718686 N, 99.448521 W). Burned grasslands at this
location are divided into two 260 ha blocks with four, 65 ha pastures
each in which either an 8- or 16-ha patch is burned each spring. Located
in a mixed-grass prairie ecoregion, this location has a rolling
topography and receives an average of 468 mm annual precipitation.
Vegetation is mixed-grass prairie invaded by introduced, C\(_3\)
grasses; stands are dominated by \emph{Pascopyrum smithii},
\emph{Nassella viridula}, \emph{Poa pratensis}, \emph{Bromus inermis},
\emph{Koeleria macrantha}, \emph{Artemisia} spp., \emph{Solidago} spp.,
and clumps of \emph{Glycyrrhiza lepidota} and \emph{Symphoricarpos
occidentalis}.

In southwestern North Dakota, we sampled ten, 16-ha fall (October) fires
in two blocks at the North Dakota State University Hettinger Research
Extension Center, Hettinger, ND (46.004443 N, 100.646477 W) with mean
annual precipitation of 380 mm. Topography is consistently flat. Located
in a shortgrass prairie ecoregion, these pastures are dominated by
introduced C\(_3\) grasses \emph{Thinopyrum intermedium}, \emph{Bromus
inermis}, \emph{Agropyron cristatum}, and \emph{Poa pratensis}, along
with the non-native legume \emph{Medicago sativa}.

\hypertarget{data-collection}{%
\subsection{Data collection}\label{data-collection}}

We measured each fire with a set of 9, 1-m equilateral triangle plots
arranged in a nested fashion such that three 10 m triangles, each
containing 3, 1-m plots, were placed 100 m apart to form a total plot
area of 0.433 ha with 27 sample points positioned at the centroid of
each burn unit (a schematic of this layout is presented in Supplemental
Information Figure 2). This fractal design is modified from the
Sierpinksi triangle described by \citet{dorrough2007} and applied to
measuring wildland fire spread by \citet{mcgranahan2021b}. Although the
logarithmically-scaled nested design was intended for geospatial
analysis of point-level data, for our analyses here we calculate
averages from the finest (1 m microplot) scale, consistent with the
method of locating multiple microplots within larger burned areas to
characterize spatial variability within fires \citep{fernandes2000}.

\textbf{Fuel data} were collected no more than three hours prior to fire
ignition. We clipped and collected all fuels in a 25 × 25 cm quadrat
positioned 0.5 m away from each 1 m triangle vertex; the three
measurements per plot were averaged prior to analysis. Fuel samples were
initially placed in airtight plastic bags to retain moisture, and then
weighed, dried to constant mass at 60\(^\circ\)C for 48 hours, and
reweighed. These data were used to calculate percent fuel moisture
content (expressed on a dry-weight basis) and fuel load (kg m\(^{2}\))
for each plot (n = 9 subsamples for each fire).

\textbf{Fire behavior data} were recorded as temperature (\(^\circ\)C) associated with the advancing flame front at each of the 27 points arranged in 9, 1-m triangular microplots at the center of each burn
unit. Data were recorded with the open-source FeatherFlame thermocouple
datalogger system \citep{mcgranahan2021b}, logging at 1.5 Hz. The
FeatherFlame system reads overbraided, ceramic fiber-insulated, 20-gauge
K-type thermocouples (Omega, Norwalk, CT) connected to an Arduino-based
datalogger assembled from Adafruit Feather breakout boards (M0
Adalogger, datalogging shield, and OLED display; Adafruit Industries,
LLC, New York City, NY) and housed inside water-resistant Pelican cases
(Pelican Products, Inc, Torrance, California). 
The low cost of open-source systems make multiple units more affordable than proprietary data loggers with no sacrifice in data quality \citep{mcgranahan2021}.
More details on the datalogger system are available in Supplemental Information.

Another advantage of the multi-channel datalogger is the opportunity to calculate \emph{two-dimensional rate of spread} from the arrival time of the flame front at vertices of a thermocouple array logged to a common timestamp. 
While one-dimensional rate of spread requires direct observation of a flame front moving perpendicular to a vector of fixed points, two-dimensional rate of spread is ideal for larger-scale fires in which direct observation is infeasible and the exact angle of approach for oncoming flame fronts is not known \citep{finney2021}. 
Here, we calculate 2-D rate of spread through the 1 m triangular microplots using trigonometric equations from \citet{simard1984}.

At each 1 m triangular microplot---the observational unit in the nested plot design---we used four thermocouples connected to a single FeatherFlame datalogger. 
Three thermocouples measured flame temperature 15 cm above the soil surface at each 1 m vertex while a fourth thermocouple recorded soil surface temperature at a representative point within the 1 m array. 
Beads of the thermocouple probes extended at least 3 cm from the supporting apparatus, to which the insulated lead was attached with wire. 
The soil thermocouple was placed on mineral ground, perpendicular to the soil surface, below plant litter.

For each fire event, we determined the time and value of the maximum temperature (\degC) as the flame front encountered each thermocouple. 
We calculated the maximum flame temperature (\degC) 15 cm above soil surface for each microplot as the mean of the thermocouple readings from its 3 vertices. 
We calculated the rate of spread (m min\textsuperscript{-1}) of the flame front as it passed through
each microplot using the maximum temperature timestamps as arrival times following equations from \citet{simard1984}, which are presented in full in Supplemental Information.

\textbf{Fire weather data} were obtained after each fire from records
made available by the North Dakota Agriculture Weather Network, the
statewide mesonet system with sensor arrays at both experimental
stations. We downloaded hourly relative humidity (\%), dew point
(\degC), air temperature (\degC), and average wind speed (m s\textsuperscript{-1}). From these data we calculated atmospheric vapor pressure ($e$) and saturation vapor pressure ($e_s$) and used these
quantities to determine the vapor pressure deficit ($VPD = e - e_s$)
for the hour in which each fire behavior observation occurred. These
data capture hourly trends in weather at the meso-gamma scale \citep[2-20 km;][]{orlanski1975}, and are reliably connected to our fire behavior measurements via time stamps provided by the dataloggers. 
The Hettinger mesonet array is 3--8 km from burned pastures, while the Central Grasslands mesonet array is 1--7.5 km from burned pastures. We
found a high degree of consistency between these meso-scale data and
fire weather records made on the fireline during operational periods,
and the open rangeland physiognomy with flat to rolling terrain
precludes substantial microsite differences in weather between these
records and the conditions at each fire behavior sample point.

\hypertarget{data-analysis}{%
\subsection{Data analysis}\label{data-analysis}}

Prior to analysis, we ensured statistical power across 167 observational
units by using multiple imputation to interpolate missing datapoints, as
missing field data occurred for three rate of spread samples (2\% of
total), 29 fuel load values (17\%), and 46 soil surface temperature
values (27\%) due to logistical and time constraints during the
operational burn periods. We used the multiple imputation method in the
\emph{mice} package \citep{vanbuuren2011} in the \textsf{R} statistical environment \citep{rct2020} to fill in these missing values. 
The procedure simulated 50 datasets with different, but reasonable, values for the missing data based on patterns in the existing data. 
We then scaled all variables to a common range within each imputed dataset, performed regression analysis on each \emph{mice}-generated dataset, and report composite statistical results pooled from the results of the 50 individual regression models.

\textbf{Multivariate analysis.---}We first conducted a multivariate analysis to assess composite relationships among the fuel, weather, and fire behavior responses. 
We used Principal Components Analysis (PCA) fit with the \texttt{rda} function in the \emph{vegan} package for \textsf{R} \citep{oksanen2017}. 
We performed post-hoc group (location) and gradient (fire weather) analysis with the \emph{vegan} \texttt{envfit} function, stratified by year.

\textbf{Regression analysis.---}We assessed weather and fuel effects on three fire behavior response variables: Maximum flame temperature (\degC) 15 cm above the soil surface (mean of three thermocouples), maximum soil surface temperature (\degC; single thermocouple), and rate of spread (m min\textsuperscript{-1}) through each 1 m triangular microplot. 
Because all three response variables were best modeled with a gamma distribution, we fit generalized linear mixed-effect regression models for each response with the \texttt{glmer} function from the \emph{lme4} package in \textsf{R} \citep{bates2015}. 
Fixed effects consisted of weather and fuel variables, as described above. 
The
random-effect term was constructed to account for random variance among
locations, spatial non-independence within locations and nested variance
within sample plots, and the effect of repeated measurements within each
location. Due to concerns about collinearity between relative humidity
and vapor pressure deficit because they are derived from the same
variables, vapor pressure deficit was excluded from regression analysis
for all three response variables.

\hypertarget{results}{%
\section{Results}\label{results}}

\begin{figure}[!h]
\includegraphics[width = 1\columnwidth]{data_summary_gg-1.pdf}
\caption{Distribution of weather, fuel, and fire behavior data for fires in southwestern North Dakota (Hettinger, dark maroon) and central North Dakota (Central Grasslands, light blue). 
Horizontal gray lines denote 25\%, 50\% (median) and 75\% quantiles; triangles are arithmetic means. 
Means and standard deviation are also reported in Supplemental Information Table 1. 
VPD = Vapor pressure deficit.}
\label{DataSummary}
\end{figure}

Most measures of fuel, fire weather, and fire behavior showed
considerable variability within each location, although rates of spread
were generally low (Fig.~\ref{DataSummary}). 
Principal Components Analysis indicated fire behavior patterns were consistent across locations ($P$ = 0.11), although the fall season in Hettinger---our semi-arid location in southwestern North Dakota---tended to have drier air and hotter fires (Fig.~\ref{PCA}). 
Spring fires in the Central Grasslands were conducted under warmer and more evaporative (VPD) conditions than fall fires at
Hettinger. The first two axes of the Principal Components Analysis (Fig.
2) explained 86\% of overall variance in the fire behavior dataset. The
first axis (52\% variance explained) was most strongly associated with
flame temperature and rate of spread, while the second axis was more
strongly associated with soil surface temperature. Dew point was
marginally related ($P~<$ 0.05) and inversely associated with
flame temperatures and rate of spread.

\begin{figure} 
\includegraphics[width = 1\columnwidth]{pca_gg-1.pdf}
\caption{Principal Components Analysis of fire behavior data (response
variables in blue; rate of spread (ROS), temperature above surface
(flame \degC), and temperature at soil surface (soil \degC) for prescribed
burns on rangeland at Hettinger (H), in southwestern North Dakota, and
Central Grasslands (CG), in central North Dakota. No difference between
locations ($P$ = 0.11). 
Total variance explained in these two axes = 86\%.}
\label{PCA}
\end{figure}

Fires at both locations were characterized by considerable variability
among sub-plots. Above the soil surface, half of the fires (13) exceeded
325\(^\circ\)C, but only four of those fires had \textgreater50\% of
individual sample plots within the burns reach an average of
325\(^\circ\)C. Sparse fuels meant that fire did not spread to some
individual plots in some burns despite strip ignitions. Among plots that
burned, fewer than half exceeded 100\(^\circ\)C at the soil surface (49
of 121 plots). A majority of the fires (18) had at least one plot exceed
100\(^\circ\)C at the soil surface, seven had over half of the plots
exceed 100\(^\circ\)C at the soil surface, and only one fire had all
plots reach this (a spring Central Grasslands fire). Only 10 fires
reached or exceeded 325\(^\circ\)C on the soil surface, with most of
these fires reaching this point in only one plot and never in more than
half of the plots.

\begin{figure}
\includegraphics[width = 1\columnwidth]{CI_gg-1.pdf}
\caption{Regression coefficients and 95\% confidence intervals for fuel
and weather terms from models for maximum temperature at 15 cm above the
soil surface (flame), maximum temperature at the soil surface (soil),
and rate of spread.}
\label{CIs}
\end{figure}

The effects of fuel and fire weather predictor variables varied across the three response variables (Fig.~\ref{CIs}). 
Fires spread faster with higher wind speeds ($t$ = 2.92, $P~<$ 0.01), but no other variable had a statistically-significant association with rate of spread (Table~\ref{RegResults}).
Aboveground flame temperatures increased as fuel load increased $(t$ = 2.82, $P$ = 0.01) and decreased as fuel moisture increased ($t$ = -2.16, $P$ = 0.04). 
No fuel or weather variable included here had statistically-significant associations with soil surface temperature
(Table~\ref{RegResults}).

\begin{table}
\caption{Results of generalized linear mixed effect regression models testing three measure of fire behavior against four potential predictor variables. 
Statistics reflect pooled results of 50 imputed datasets using the \emph{mice} package in \textsf{R},  see Methods. Vapor pressure deficit
included for Rate of spread only due to statistically-significant
difference between GLMM regression results that included VPD compared to
RH alone (Wald = 5.32, $P$ = 0.02), while temperature models had no such
difference. \label{RegResults} }
\centering 
\begin{tabular}{llcc}
\toprule
Response & Model term & $t_{df}$ & P \\
\midrule

Rate of spread & & & \\
& Wind speed & $2.92_{108.0}$ & \textless{} 0.01 \\
& Vapor pressure deficit & -2.31 108.1 & 0.02 \\
& Relative humidity & -1.66 80.8 & 0.10 \\
& Fuel load & 1.16 49.9 & 0.25 \\
& Fuel moisture & -0.68 62.9 & 0.50 \\
Canopy temperature & & & \\
& Fuel load & 2.82 54.2 & 0.01 \\
& Fuel moisture & -2.16 40.9 & 0.04 \\
& Relative humidity & -1.19 120.4 & 0.24 \\
& Wind speed & 0.02 132.4 & 0.99 \\
Surface temperature & & & \\
& Relative humidity & -1.19 74.8 & 0.24 \\
& Fuel load & -0.48 20.1 & 0.64 \\
& Fuel moisture & -0.47 40.1 & 0.64 \\
& Wind speed & 0.06 49.5 & 0.95 \\
\bottomrule
\end{tabular}
\end{table}

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

In our comparison of three measurements of fire behavior---rate of spread and maximum temperature recorded on the soil surface and 15 cm
above the soil surface---against fuel and fire weather variables, we
found considerable variability in which predictor variables were
associated with different measures of fire behavior. These data directly
support the safe and effective application of prescribed fire in the
region. Some results are straightforward and consistent with decades of
fire safety science; e.g., faster rates of fire spread are associated
with higher wind speed and lower relative humidity. Other results add
nuance to an ecological understanding of how fire behavior relates to
fire effects---e.g., factors like fuel load and overall fuelbed moisture
were related with flame temperature but not soil surface temperature,
which suggests that direct effects on belowground plant tissue and soil
biota are not correlated with aboveground heating and fire spread.

To our knowledge, this is the first study from the northern Great Plains
to scrutinize the factors that influence fire behavior, and the first to
combine reports of fire spread and temperature data from thermocouples.
Most published research on fire spread in the Great Plains is derived
from computer simulations
\citep{mcgranahan2013, overholt2014, yurkonis2019}. The few field
studies from the region mostly report temperature data from
thermocouples and rarely incorporate fuel and fire weather data into the
analysis; when such information is provided, it is typically included in
the study description, not as data. Given the high degree of variability
in the wildland fire environment, a mechanistic understanding of
grassland fire dynamics will require collect fuel, fire weather, and
fire behavior data in a spatially and temporally consistent manner to
facilitate statistical analyses of their relationships
\citep{hiers2020, mcgranahan2018}.

Mean temperatures recorded in this study are consistent with other
reports from northern rangelands. In central Alberta grassland,
\citet{bailey1980} observed that surface temperatures varied between
110\(^\circ\)C and 165\(^\circ\)C for backfires and headfires,
respectively, and headfires averaged 200\(^\circ\)C 15 cm above the
ground; temperatures generally tracked with fuel load. Surface fires
through jack pine barrens in Ontario had a similar range as ours
\citep[140-545\(^\circ\)C,][]{smith1966}. In our study, mean 15-cm
temperatures were 225\(^\circ\)C during spring burns in central North
Dakota and 250\(^\circ\)C during fall burns in southwestern North
Dakota; surface temperatures at both locations generally averaged just
above 100\(^\circ\)C (Fig. 1).

Discrepancies between our data and others from the region are consistent
with what would be expected when differences in the fire environment are
considered. For example, \citet{ohrtman2015} reported a wide range of
maximum temperatures at the soil surface---150-500\(^\circ\)C---that was
generally explained by variability in annual productivity and clipping
frequency, which altered fuel load. Our fires were also cooler than
those reported by \citet{archibold2003} in Saskatchewan: using the
mid-point of observations made at 10 cm and 20 cm as a comparison to our
15-cm values, spring fires reached 314\(^\circ\)C and fall fires reached
298\(^\circ\)C. But \citet{archibold2003} also reported substantially
lower fuel moisture in each season and they had approximately three
times the fuel load, likely due to an absence of grazing on the remnant
prairie. A previous study reported similar results---temperatures
approaching 500\(^\circ\)C when fuel loads averaged 2.8-4.5 t
ha\(^{\text{-}1}\) \citep{archibold1998}. With greater variability in
fuel load and fuel moisture, we might also expect to see these factors
have greater influence on aboveground flame temperatures. For example,
in Colorado, \citet{augustine2014} observed a strong linear relationship
between fuel load and temperatures ranging from 60-200\(^\circ\)C, but
their fuel load also ranged from 0.2 to 1.2 t ha\(^{\text{-}1}\).

Fires at both of our locations spread much more slowly than most reports
from other grasslands, due primarily to high fuelbed moisture content
and little opportunity for mitigation by wind or lower atmospheric
moisture (Fig. 1). Perhaps the most variability in fire spread was
reported by \citet{sneeuwjagt1977} from prescribed grass fires in
California and Washington, where rates of spread ranged from 0.2-61 m
min\(^{\text{-}1}\). From grassland fires in South Africa and Kansas,
USA, \citet{trollope2002} reported average spread rates of 24 and 32 m
min\(^{\text{-}1}\), respectively, for head fires and 0.12 and 0.14 m
min\(^{\text{-}1}\), respectively, for back fires. In a tallgrass
prairie in Texas, \citet{clements2019} recorded fire spreading between
72-150 m min\(^{\text{-}1}\) with the wind and 48 m min\(^{\text{-}1}\)
for flanking fires. Likewise, fires through cured grass fuelbeds in
Australia spread much more rapidly than we observed---up to 18-180 m
min\(^{\text{-}1}\) \citep{cheney1993, cheney1995, cruz2015}. Through
partially-cured stands, though, \citet{cruz2015} found spread rates
dropped to 3-44 m min\(^{\text{-}1}\), approaching our location averages
of 2.1 and 3.1 m min\(^{\text{-}1}\). Consistent with our finding that
only wind speed had a statistically-significant effect on increasing
rate of spread (Fig. 3), \citet{cheney1993} found that wind was by far
the most important variable to spread rate.

Interestingly, \citet{cheney1993} found that fires spread faster in
undisturbed pastures compared to those that had been cut, which they
attribute to differences in fuel structure (height, bulk density) rather
than fuel load. This might have implications for fire behavior in our
region, where invasive species like \emph{Poa pratensis} generally
increase aboveground plant biomass but do so by adding thick dense
litter at the soil surface, rather than standing dead fuel in the plant
canopy \citep{gasch2020}. While difficult to tease apart statistically
in the present data, many burn units in our mesic location in central
North Dakota were dominated by \emph{P. pratensis} and indeed, that
location tended to have higher fuel loads and lower rates of spread
(Fig.~\ref{DataSummary}), consistent with simulations of fire spread through those
\emph{P. pratensis}-dominated prairies \citep{yurkonis2019}.

Much is made of the difference in fire behavior between head and back
fires in the fire ecology literature, and while we expect these
differences translate to different fire effects in our system, making
distinctions between fire types is difficult in both our data and our
management. \citet{trollope1978} emphasized that while head fires move
faster and generally release more energy, back fires effect greater
heating at the ground level. One would expect, then, that back fires
would have more opportunity to burn down through even thick \emph{P.
pratensis} litter to mineral soil. Unfortunately, our results offer
little insight into what fuel or weather variables enhance litter
combustion, likely because most of our fires never got very hot at the
soil surface---50\% of our observations were less than 60\(^\circ\)C,
and 60\% less than 100\(^\circ\)C (Fig.~\ref{DataSummary}). 
Nor can we differentiate the direction of fire spread with the current trigonometry applied to the
triangular thermocouple arrays \citep{simard1984}, although it would
theoretically be possible to compare spread direction to wind direction
if the latter data were available at a fine enough scale.

The functional difference between head and back fires in the fuelbeds
reported here is likely moot. Because our fuels were often sparse,
sometimes only marginally cured, and prescriptions precluded taking
advantage of higher wind or lower relative humidity to mitigate fuel
limitations, we often employed substantial interior ignitions using
strip, point, flanking, and spiral patterns that sent flame fronts
towards our sensors in all possible directions at different times. While
\citet{williams2015} did show that different ignition patterns created
additional spatial variability in fire behavior, the effect of shorter
line ignitions and spot ignitions was to mitigate the high severity of
wildfire and long line ignitions in highly-flammable spinifex fuelbeds.
In our case, we had to manipulate ignition pattern just to get fire to
carry. Our data are certainly useful in describing the variability in
fire behavior across these burns, but do not inform the relationship
between ignition pattern and fire behavior. Thus, future research on
fire behavior in the northern Great Plains should (1) use experimental
plots with consistent fuelbeds to explicitly compare head and back fires
set via line ignitions, akin to the experimental burning program
described by \citet{cruz2015} in Australia, and (2) attempt to at least
address severity, if not fire behavior, in wildfire scenarios via remote
sensing and/or modelling.

A more detailed analysis of fuelbed effects on fire behavior also ought
to separate fuels into live and dead components, and consider the
moisture content of each along with fuel load ratios. Explicitly
measuring litter moisture might also be valuable, especially when
variability in belowground heating is expected to influence first-order
fire effects. We report here the overall fuel moisture content of the
entire fuelbed, consistent with descriptive, post hoc statistical
approaches to describing fire behavior
\citep{trollope1978, trollope1985, bidwell1992, trollope2002, mcgranahan2016}.
But predictive fire behavior models accommodate inputs for live and dead
fuel categories \citep{scott2005}, and \citet{kidnie2015a} found that
four categories of live, dead, and senescent fuels best represented
differences in grassland fuel moisture scenarios. With this in mind,
\citet{cruz2015} employed a hybrid approach in which fuel moisture was
measured for the various components, from which a weighted average was
used as a predictor variable in regression analysis. They subsequently
found that overall degree of curing, not simply live fuel moisture, was
the most important variable in explaining the dampening effect of fuel
moisture on fire behavior.

Often, time and resource constraints preclude the separation of fine
plant material by live and dead class, and overall fuelbed moisture
content is the best available data for managers. Although parsing live
and dead fuel moisture in the present analysis would probably not better
explain variability in our dataset, it would likely contribute to better
predictions of fire behavior relative to management objectives if
information on overall fuelbed moisture content were available prior to
ignition. Unfortunately, the standard clipping and drying method is not
compatible with providing day-of fuel moisture data, and visual
assessments based on color tend to over-predict curedness
\citep{kidnie2015a}. However, electronic devices can provide accurate
and instantaneous measurements of grassland fuel moisture
\citep{mcgranahan2019}.

Research must also address the influence of atmospheric moisture
conditions on prescribed fire behavior. Several broad-scale, post-hoc 
analyses of wildfire conditions conclude that atmospheric moisture is an
important driver of burned area \citep{evett2008, reid2010, sedano2014}.
But experiments that explicitly test the immediate effect of relative
humidity on fire behavior report no appreciable effect on surface fire
temperatures or rate of spread \citep{sparling1966, trollope1985}. It is
likely that atmospheric moisture plays a larger role in modulating fuel
moisture content prior to combustion than affecting instantaneous fire
behavior itself---consider how fire behavior models take fuel moisture
as a parameter and not relative humidity, but include relative humidity
as an input to determine fuel moisture content
\citep{rothermel1983, cruz2016}.

The most appropriate measure of atmospheric moisture content might also
be unresolved. We focused our analysis here on relative humidity because
it is so common in fire behavior models and fire weather forecasts. But
vapor pressure deficit has also been identified as an important driver
of fire spread and intensity \citep{gomes2020}. In fact,
\citet{srock2018} suggest vapor pressure deficit might be a better
measure of atmospheric moisture content for fire predictions, but the
Hot-Dry-Windy index they developed to incorporate vapor pressure deficit
operates at synoptic scales beyond the spatial extent and operational
periods of prescribed burns. Given that substantial changes in
atmospheric moisture changes in recent decades are expected to
strengthen over the 21\textsuperscript{st} century
\citep{seager2015, ficklin2017}, understanding how these dynamics affect
fire behavior will be an essential component of managing resilient fire
regimes.

This study is novel in that it examines the fire environment at a spatial scale consistent with land management in the region using
realistic ignition scenarios. 
To our knowledge, no other study in the northern Great Plains has reported the behavior of fires larger than experimental plots. Integrating research into management almost invariably requires trade-offs; two already discussed above include (1)
measuring only the overall moisture content of the entire fuelbed, being
precluded from parsing fuel into live, dead, and litter components, and
(2) measuring two-dimensional rate of spread of fire fronts within the
burn unit without being able to associate them with wind direction. But
these are the respective conditions under which prescribed fire managers
in the region decide whether to burn, and ensure fire spread objectives
are met. Research conducted at the spatial scales at which management
occurs helps managers trust the transfer of knowledge from studies to
working landscapes \citep{sayre2005, cacciapaglia2012}. For example, all
of our fire behavior measurements were made more than 50 m from the
initial fire line, the distance identified in simulations and used in
wildland fire science to allow flame fronts to achieve a quasi-steady
state in spread rate \citep{fernandes2000, sutherland2020}, which is
obviously precluded in studies that employ small plots.

\backmatter

\bmhead{Supplementary Information}

Supplementary Information containing additional information on the study location, rate of spread calculations, and \textsf{R} script is available in the online version of the paper.

\bmhead{Acknowledgments}

We appreciate support from the North Dakota State Agricultural Experiment Station, including K. Sedivec at the Central Grasslands REC and C. Schauer and B. Geaumont at the Hettinger REC. 
We recognize funding from the University of North Dakota Department of Biology, USDA-NIFA Hatch project number ND02393, and USDA-NIFA AFRI award number 2018-67020-27856. 
We appreciate the technical assistance of several North Dakota State University graduate students and faculty for assistance with prescribed burning. 
L. LaFond, C. LaFond, and E. Wahl assisted with datalogger assembly. J. Spiess, J. Cutter, M. Lakey, B. Poling, and A. Steele assisted with sensor deployment and data collection.

 \bibliography{FireBehaviorBib.bib}

\end{linenumbers}
\end{document}

### 2023-10-12 paper/springer/CoverLetter/CoverLetter_R2.tex
surface update

--- 2022-08-24 paper/springer/CoverLetter/CoverLetter_R1.tex
@@
 
-\vspace{-5em}
+%\vspace{-5em}
 
@@
 
-Please consider the attached submission for publication in \journal.
-
-This submission is motivated by the positive experience my colleague and I recently enjoyed publishing in \journal~(DOI \href{https://doi.org/10.1007/s00267-022-01659-y}{10.1007/s00267-022-01659-y}). 
-I have decided to move on with this work after a lack of response from the Editor in Chief at the journal to which we initially submitted, following mishandling by an Associate Editor. 
-After our initial review, the paper was understandably rejected and a resubmission requested, due to the extent of the reviewer comments. 
-Each were productive, and we responded completely (these responses are attached). 
-However, despite no reviewer criticism on the resubmitted manuscript, the (presumably new?) AE assigned to the resubmission rejected the manuscript on the grounds that our approach to measuring two-dimensional rate of spread violated assumptions that are in fact limited to one-dimensional rate of spread.
-But recently, Dr.
-Mark Finney\textemdash almost certainly the leading wildland fire scientist in the US, if not the world\textemdash led a textbook in which his team described the superiority of 2-D measurements over 1-D measurements.
-We brought this to the EiC's attention in our appeal of the AE decision, but have gone several weeks without a response.
-
-As these data are of great interest to our stakeholders, we need to press on with publication. 
-I think the scope and audience of \journal~is a good fit for us. 
-We have explicitly referenced Finney et al's textbook into our Methods section to make this point clear. 
-While I understand \journal~has no formal Portable Peer-Review process, I hope the editorial staff might take this context into account and consider an expedited review process.
+We appreciate the opportunity to revise our submission for publication in \journal.
+We've taken the editor's comments about further review seriously. 
+We trust the editor will find that each of the second reviewer's comments have been fully replied to in this second round of review. 
+Unfortunately the new major revisions period of 14 days is prohibitively short, and our attention in the previous review was focused on the detailed critique of the first reviewer. 
+In this latest round, we have attended to all of the second reviewer's points and appreciate the additional detail the reviewer provided.
 

### 2023-10-12 paper/springer/GrasslandFireBehaviorNGP_R2.tex
surface update

--- 2022-08-24 paper/springer/GrasslandFireBehaviorNGP_R1.tex
@@
 More than simply the result of combustion of vegetation, wildland fire behavior is multi-faceted, with different components producing different effects on the surrounding environment and organisms within.
-Most wildland fire scientists describe fire behavior in terms of \emph{rate of spread}---how quickly a flame front moves through a fuelbed---and \emph{intensity}---a suite of measurements of how much energy is released by combustion, often expressed as a rate of energy release over time \citep{mcgranahan2021a}.
+Unlike wildfire events, in which incident commanders are most interested in understanding fire behavior to ensure safe suppression operations, often during extremely dry conditions, prescribed fire managers seek to understand fire behavior to achieve specific vegetation  objectives under a range of weather and fuel moisture conditions.
+In either context, wildland fire scientists describe fire behavior in terms of \emph{rate of spread}---how quickly a flame front moves through a fuelbed---and \emph{intensity}---a suite of measurements of how much energy is released by combustion, often expressed as a rate of energy release over time \citep{mcgranahan2021a}.
 
@@
 increase as more fuel is available to burn. The structure and
-arrangement of vegetation is also important. Greater fuel load
-attributable to longer time-since-fire increases fire temperature, and
+arrangement of vegetation is also important. Greater fuel load increases fire temperature, and
 spatial variability in fuel load and patchy distribution of fine fuels
@@
 grasses burn more completely and hotter than an equal mass of forbs
-\citep{wragg2018}. Finally, fuel moisture content is an especially
-important driver outside of the highly-cured context of wildfire seasons
-\citep{sparling1966, kidnie2015}. Together, variability in flammability
+\citep{wragg2018}. 
+Finally, while highly cured, uniformly low fuel moisture content in active wildfire seasons contributes to greater energy release and faster rates of spread, fire behavior outside of these seasons\textemdash when prescribed fire is typically applied\textemdash is often limited by high fuel moisture content \citep{sparling1966, kidnie2015}.
+Together, variability in flammability
 traits and curing rates among species that comprise grassland fuelbeds
@@
 interpreting thermocouple data \citep[see review by][]{mcgranahan2020}.
-Firstly, a considerable amount of variability in temperature is
-attributable to sensor placement relative to both the ground and the
-fire. There is neither a standard for placing thermocouple probes in the
-wildland fire environment nor consistency in vertical temperature
-profiles. Most observations of surface fire temperature profiles
-describe an inverse, linear relationship between height and temperature
-\citep{smith1966, patten1984, archibold2003}, although
-\citet{ramsay1996} found the highest temperatures at the top of a 1 m
-profile and the lowest temperatures at 30 cm, while \citet{frost1987}
-and \citet{bailey1980} present evidence that the highest temperatures
-occur midway up the profile. At least some of this variability might be
-due to differences in surface vs.~flame temperature among head and back
-fires \citep{trollope1978}.
+A considerable amount of variability in temperature is attributable to sensor placement relative to both the ground and the fire (e.g., see conflicting reports of peak temperatures relative to soil surface described in \citet{smith1966, patten1984, archibold2003, ramsay1996, frost1987, bailey1980}).
+Additionally, many factors that contribute to variability in temperatures recorded by thermocouples are attributable to the nature of the sensor rather than the nature of the fire \citep[e.g.,][]{walker1968}, impeding comparisons between studies \citep{bova2008}.
+And because temperature of the media around a probe is a poor proxy for the thermal experience of an organism, measures of intensity or energy flux are more biologically relevant \citep{kremens2012, smith2016}.
+Rate of spread, rather than temperature, is a factor in calculating fire intensity \citep{byram1959}, and can be derived from  thermocouple arrays by simply using the timestamps of peak heating to calculate rate of spread \citep{mcgranahan2021b, finney2021}.
 
-Secondly, many factors that contribute to variability in temperatures
-recorded by thermocouples are attributable to the nature of the sensor
-rather than the nature of the fire \citep[e.g.,][]{walker1968}. Thus,
-reporting `device temperatures' alone impedes comparisons between
-studies; \citet{bova2008} present a standard calibration of thermocouple
-probes, while \citet{mcgranahan2021b} simply uses the timestamps of peak
-heating across an array of several thermocouples to calculate rate of
-spread. Using rate of spread as the response variable makes moot the
-third issue with temperature as a fire behavior metric: temperature of
-the media around a probe is a poor proxy for the thermal experience of
-an organism. Measures of intensity or energy flux are more biologically
-relevant \citep{kremens2012, smith2016}.
-
-In the North American Great Plains, most reports of grassland fire
-behavior consist of temperatures derived from thermocouples, and there
-are few data on rate of spread. Soil surface temperatures in South
-Dakota tallgrass prairie ranged from 200-500\(^\circ\)C during spring
-burns, and were greatest under lower fuel loads \citep{ohrtman2015};
-fires in Saskatchewan mixed grass prairie exceeded 300\(^\circ\)C 5-10
-cm above the soil surface in spring, summer, and fall
-\citep{archibold2003}. Mean temperatures in experimental burns in
-eastern Montana ranged from 172-222\(^\circ\)C in the summer to 253\(^\circ\)C in the spring \citep{strong2013, russell2015}. To our
-knowledge, no studies on grassland fire in the northern Great Plains
-region has explicitly tested the effect of fire weather on fire
-behavior.
+In the North American Great Plains, most reports of grassland fire behavior consist of temperatures derived from thermocouples, and there are few data on rate of spread. 
+Soil surface temperatures in South Dakota tallgrass prairie ranged from 200-500\(^\circ\)C during spring burns, and were greatest under lower fuel loads \citep{ohrtman2015}; fires in Saskatchewan mixed grass prairie exceeded 300\(^\circ\)C 5-10 cm above the soil surface in spring, summer, and fall \citep{archibold2003}.
+Mean temperatures in experimental burns in eastern Montana ranged from 172-222\(^\circ\)C in the summer to 253\(^\circ\)C in the spring \citep{strong2013, russell2015}. 
+To our knowledge, no studies on grassland fire in the northern Great Plains has explicitly tested the effects of fuel moisture and fuel load, or fire weather, on fire behavior.
 
@@
 
-\hypertarget{data-collection}{%
-\subsection{Data collection}\label{data-collection}}
+\subsection{Data collection}
 
@@
 
-\textbf{Fire behavior data} were recorded as temperature (\(^\circ\)C) associated with the advancing flame front at each of the 27 points arranged in 9, 1-m triangular microplots at the center of each burn
-unit. Data were recorded with the open-source FeatherFlame thermocouple
-datalogger system \citep{mcgranahan2021b}, logging at 1.5 Hz. The
+\textbf{Fire behavior data} were recorded as temperature (\(^\circ\)C) associated with the advancing flame front at each of the 27 points arranged in 9, 1-m triangular microplots at the center of each burn unit.
+Data were recorded with the open-source FeatherFlame thermocouple datalogger system \citep{mcgranahan2021b}, logging at 1.5 Hz. The
 FeatherFlame system reads overbraided, ceramic fiber-insulated, 20-gauge
@@
 Three thermocouples measured flame temperature 15 cm above the soil surface at each 1 m vertex while a fourth thermocouple recorded soil surface temperature at a representative point within the 1 m array.
+We chose 15 cm for our single observation as the midpoint of 10 and 20 cm observations reported elsewhere \citep[e.g.,][]{archibold2003}.
 Beads of the thermocouple probes extended at least 3 cm from the supporting apparatus, to which the insulated lead was attached with wire. 
@@
 We calculated the maximum flame temperature (\degC) 15 cm above soil surface for each microplot as the mean of the thermocouple readings from its 3 vertices. 
-We calculated the rate of spread (m min\textsuperscript{-1}) of the flame front as it passed through
-each microplot using the maximum temperature timestamps as arrival times following equations from \citet{simard1984}, which are presented in full in Supplemental Information.
+We calculated the rate of spread (m min\textsuperscript{-1}) of the flame front as it passed through each microplot using the maximum temperature timestamps as arrival times following equations from \citet{simard1984}, which are presented in full in Supplemental Information.
 
@@
 data capture hourly trends in weather at the meso-gamma scale \citep[2-20 km;][]{orlanski1975}, and are reliably connected to our fire behavior measurements via time stamps provided by the dataloggers.
-The Hettinger mesonet array is 3--8 km from burned pastures, while the Central Grasslands mesonet array is 1--7.5 km from burned pastures. We
-found a high degree of consistency between these meso-scale data and
-fire weather records made on the fireline during operational periods,
-and the open rangeland physiognomy with flat to rolling terrain
-precludes substantial microsite differences in weather between these
-records and the conditions at each fire behavior sample point.
+The Hettinger mesonet array is 3\textendash 8 km from burned pastures, while the Central Grasslands mesonet array is 1--7.5 km from burned pastures. 
+We found a high degree of consistency between these meso-scale data and fire weather records made on the fireline during operational periods, and the open rangeland physiognomy with flat to rolling terrain precludes substantial microsite differences in weather between these records and the conditions at each fire behavior sample point.
 
-\hypertarget{data-analysis}{%
-\subsection{Data analysis}\label{data-analysis}}
+\subsection{Data analysis}
 
@@
 Fixed effects consisted of weather and fuel variables, as described above. 
-The
-random-effect term was constructed to account for random variance among
+The random-effect term was constructed to account for random variance among
 locations, spatial non-independence within locations and nested variance
-within sample plots, and the effect of repeated measurements within each
-location. Due to concerns about collinearity between relative humidity
+within plots, and the effect of repeated measurements within each
+location. 
+Due to concerns about collinearity between relative humidity
 and vapor pressure deficit because they are derived from the same
@@
 
-\hypertarget{results}{%
-\section{Results}\label{results}}
+\section{Results}
 
@@
 
-Most measures of fuel, fire weather, and fire behavior showed
-considerable variability within each location, although rates of spread
-were generally low (Fig.~\ref{DataSummary}). 
+Most measures of fuel, fire weather, and fire behavior showed considerable variability within each location, although rates of spread were generally low (Fig.~\ref{DataSummary}). 
 Principal Components Analysis indicated fire behavior patterns were consistent across locations ($P$ = 0.11), although the fall season in Hettinger---our semi-arid location in southwestern North Dakota---tended to have drier air and hotter fires (Fig.~\ref{PCA}).
-Spring fires in the Central Grasslands were conducted under warmer and more evaporative (VPD) conditions than fall fires at
-Hettinger. The first two axes of the Principal Components Analysis (Fig.
-2) explained 86\% of overall variance in the fire behavior dataset. The
-first axis (52\% variance explained) was most strongly associated with
-flame temperature and rate of spread, while the second axis was more
-strongly associated with soil surface temperature. Dew point was
-marginally related ($P~<$ 0.05) and inversely associated with
-flame temperatures and rate of spread.
+Spring fires in the Central Grasslands were conducted under warmer and more evaporative (VPD) conditions than fall fires at Hettinger. 
+The first two axes of the Principal Components Analysis (Fig.~\ref{PCA}) explained 86\% of overall variance in the fire behavior dataset. 
+The first axis (52\% variance explained) was most strongly associated with flame temperature and rate of spread, while the second axis was more strongly associated with soil surface temperature. 
+Dew point was marginally related ($P~<$ 0.05) and inversely associated with flame temperatures and rate of spread.
 
@@
 Total variance explained in these two axes = 86\%.}
-\label{PCA}
+\label{PCA} % (Fig.~\ref{PCA})
 \end{figure}
 
-Fires at both locations were characterized by considerable variability
-among sub-plots. Above the soil surface, half of the fires (13) exceeded
-325\(^\circ\)C, but only four of those fires had \textgreater50\% of
-individual sample plots within the burns reach an average of
-325\(^\circ\)C. Sparse fuels meant that fire did not spread to some
-individual plots in some burns despite strip ignitions. Among plots that
-burned, fewer than half exceeded 100\(^\circ\)C at the soil surface (49
-of 121 plots). A majority of the fires (18) had at least one plot exceed
-100\(^\circ\)C at the soil surface, seven had over half of the plots
-exceed 100\(^\circ\)C at the soil surface, and only one fire had all
-plots reach this (a spring Central Grasslands fire). Only 10 fires
-reached or exceeded 325\(^\circ\)C on the soil surface, with most of
-these fires reaching this point in only one plot and never in more than
-half of the plots.
+The effects of fuel and fire weather predictor variables varied across the three response variables (Fig.~\ref{CIs}). 
+Fires spread faster with higher wind speeds ($t$ = 2.92, $P~<$ 0.01), but no other variable had a statistically-significant association with rate of spread (Table~\ref{RegResults}).
+Aboveground flame temperatures increased as fuel load increased $(t$ = 2.82, $P$ = 0.01) and decreased as fuel moisture increased ($t$ = -2.16, $P$ = 0.04). 
+No fuel or weather variable included here had statistically-significant associations with soil surface temperature (Table~\ref{RegResults}).
 
@@
 
-The effects of fuel and fire weather predictor variables varied across the three response variables (Fig.~\ref{CIs}). 
-Fires spread faster with higher wind speeds ($t$ = 2.92, $P~<$ 0.01), but no other variable had a statistically-significant association with rate of spread (Table~\ref{RegResults}).
-Aboveground flame temperatures increased as fuel load increased $(t$ = 2.82, $P$ = 0.01) and decreased as fuel moisture increased ($t$ = -2.16, $P$ = 0.04). 
-No fuel or weather variable included here had statistically-significant associations with soil surface temperature
-(Table~\ref{RegResults}).
-
 \begin{table}
 \caption{Results of generalized linear mixed effect regression models testing three measure of fire behavior against four potential predictor variables. 
-Statistics reflect pooled results of 50 imputed datasets using the \emph{mice} package in \textsf{R},  see Methods. Vapor pressure deficit
-included for Rate of spread only due to statistically-significant
-difference between GLMM regression results that included VPD compared to
-RH alone (Wald = 5.32, $P$ = 0.02), while temperature models had no such
+Statistics reflect pooled results of 50 imputed datasets using the \emph{mice} package in \textsf{R},  see Methods. 
+Vapor pressure deficit included for Rate of spread only due to statistically-significant difference between GLMM regression results that included VPD compared to RH alone (Wald = 5.32, $P$ = 0.02), while temperature models had no such
 difference. \label{RegResults} }
@@
 
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
+\section{Discussion}
 
@@
 reports from northern rangelands. In central Alberta grassland,
-\citet{bailey1980} observed that surface temperatures varied between
+\citet{bailey1980} observed that temperatures at the surface varied between
 110\(^\circ\)C and 165\(^\circ\)C for backfires and headfires,
-respectively, and headfires averaged 200\(^\circ\)C 15 cm above the
-ground; temperatures generally tracked with fuel load. Surface fires
-through jack pine barrens in Ontario had a similar range as ours
+respectively, while at 15 cm above the ground, headfires averaged 200\(^\circ\)C; temperatures generally tracked with fuel load. 
+Surface fires through jack pine barrens in Ontario had a similar range as ours
 \citep[140-545\(^\circ\)C,][]{smith1966}. In our study, mean 15-cm
@@
 Dakota; surface temperatures at both locations generally averaged just
-above 100\(^\circ\)C (Fig. 1).
+above 100\(^\circ\)C (Fig.~\ref{DataSummary}).
 
-Discrepancies between our data and others from the region are consistent
-with what would be expected when differences in the fire environment are
-considered. For example, \citet{ohrtman2015} reported a wide range of
-maximum temperatures at the soil surface---150-500\(^\circ\)C---that was
-generally explained by variability in annual productivity and clipping
-frequency, which altered fuel load. Our fires were also cooler than
-those reported by \citet{archibold2003} in Saskatchewan: using the
-mid-point of observations made at 10 cm and 20 cm as a comparison to our
+Discrepancies between our data and others from the region are consistent with what would be expected when differences in the fire environment are considered. 
+For example, ungrazed pastures in southwestern North Dakota with higher litter loads burned much hotter at the soil surface, at an average of 275$^\circ$C \citep{mcgranahan2022}.
+\citet{ohrtman2015} reported a wide range of maximum temperatures at the soil surface---150-500\(^\circ\)C---that was generally explained by variability in annual productivity and clipping frequency, which altered fuel load.
+Our fires were also cooler than those reported by \citet{archibold2003} in Saskatchewan: using the mid-point of observations made at 10 cm and 20 cm as a comparison to our
 15-cm values, spring fires reached 314\(^\circ\)C and fall fires reached
@@
 approaching 500\(^\circ\)C when fuel loads averaged 2.8-4.5 t
-ha\(^{\text{-}1}\) \citep{archibold1998}. With greater variability in
-fuel load and fuel moisture, we might also expect to see these factors
-have greater influence on aboveground flame temperatures. For example,
+ha\(^{\text{-}1}\) \citep{archibold1998}. 
+With greater variability in fuel load and fuel moisture, we might also expect to see these factors have greater influence on aboveground flame temperatures. For example,
 in Colorado, \citet{augustine2014} observed a strong linear relationship
@@
 and little opportunity for mitigation by wind or lower atmospheric
-moisture (Fig. 1). Perhaps the most variability in fire spread was
+moisture (Fig.~\ref{DataSummary}).
+Perhaps the most variability in fire spread was 
 reported by \citet{sneeuwjagt1977} from prescribed grass fires in
@@
 dropped to 3-44 m min\(^{\text{-}1}\), approaching our location averages
-of 2.1 and 3.1 m min\(^{\text{-}1}\). Consistent with our finding that
-only wind speed had a statistically-significant effect on increasing
-rate of spread (Fig. 3), \citet{cheney1993} found that wind was by far
+of 2.1 and 3.1 m min\(^{\text{-}1}\). 
+Consistent with our finding that only wind speed had a statistically-significant effect on increasing
+rate of spread (Fig.~\ref{CIs}), \citet{cheney1993} found that wind was by far
 the most important variable to spread rate.
@@
 combustion, likely because most of our fires never got very hot at the
-soil surface---50\% of our observations were less than 60\(^\circ\)C,
-and 60\% less than 100\(^\circ\)C (Fig.~\ref{DataSummary}). 
-Nor can we differentiate the direction of fire spread with the current trigonometry applied to the
-triangular thermocouple arrays \citep{simard1984}, although it would
-theoretically be possible to compare spread direction to wind direction
-if the latter data were available at a fine enough scale.
+soil surface  (Fig.~\ref{DataSummary}). 
+Nor can we differentiate the direction of fire spread with the current trigonometry applied to the triangular thermocouple arrays \citep{simard1984}, although it would theoretically be possible to compare spread direction to wind direction if the latter data were available at a fine enough scale.
 
-The functional difference between head and back fires in the fuelbeds
-reported here is likely moot. Because our fuels were often sparse,
-sometimes only marginally cured, and prescriptions precluded taking
-advantage of higher wind or lower relative humidity to mitigate fuel
+The functional difference between head and back fires in the fuelbeds reported here is likely moot. 
+Because our fuels were often sparse, sometimes only marginally cured, and prescriptions precluded taking advantage of higher wind or lower relative humidity to mitigate fuel
 limitations, we often employed substantial interior ignitions using
